---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd and run
     devtools::build_readme(). -->

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", message = FALSE)
```

# bayesim

<!-- badges: start -->
[![R-CMD-check](https://github.com/sims1253/bayesim/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sims1253/bayesim/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/sims1253/bayesim/graph/badge.svg)](https://app.codecov.io/gh/sims1253/bayesim)
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](LICENSE.md)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

bayesim runs simulation studies of Bayesian models in R. Define how to
generate data, which models to fit, and what to measure. It runs the
replicates and summarizes estimation error, interval coverage, or calibration.
Parallel execution and checkpoints support longer studies.

## Install

Install the development version from GitHub:

```r
# install.packages("pak")
pak::pak("sims1253/bayesim")
```

The example below runs without Stan. For brms and CmdStan setup, see
[brms studies](https://sims1253.github.io/bayesim/articles/brms-studies.html)
and [fitters](https://sims1253.github.io/bayesim/articles/custom-fitters.html).

## Try a study

Does a 90% credible interval cover the true slope about 90% of the time?
Compare three sample sizes using exact Bayesian linear regression:

```{r example}
library(bayesim)

gen <- function(data_spec, task_ctx) {
  x <- rnorm(data_spec$n)
  y <- 1 + 0.5 * x + rnorm(data_spec$n)
  list(
    train = data.frame(y = y, x = x),
    response = "y",
    true_params = c(x = 0.5),
    vars_of_interest = "x"
  )
}

config <- simulation_config(
  data_grid = data.frame(n = c(20, 50, 200)),
  fit_grid = data.frame(model = "linear"),
  data_generator = gen,
  fitter = LinearRegressionFitter(),
  metrics = list(coverage_metric(prob = 0.90)),
  n_replicates = 100L,
  seed = 42L
)

result <- run_simulation(config, progress = FALSE, verbose = FALSE)
summary <- summarize_simulation(result, by = "data_n", metrics = "coverage__by_param__x")
```

```{r, echo = FALSE}
knitr::kable(
  summary[c("data_n", "coverage__by_param__x_mean", "coverage__by_param__x_mcse")],
  col.names = c("Sample size", "Coverage", "MCSE"),
  digits = 3
)
```

This runs 300 fits: 3 sample sizes × 1 model × 100 replicates. Compare the
coverage with 0.90; its Monte Carlo standard error (MCSE) describes
uncertainty from the finite number of replicates. Use more replicates for a
precise estimate. Add `workers = 2` to `run_simulation()` to run in parallel.

Here the smallest sample has coverage below 0.90. Exact posterior inference
does not guarantee nominal interval coverage at a fixed parameter value;
this is what the study measures.

## Guides

- [Getting started](https://sims1253.github.io/bayesim/articles/getting-started.html): build a study and interpret its results.
- [Study design](https://sims1253.github.io/bayesim/articles/design-of-simulation-studies.html) and [SBC](https://sims1253.github.io/bayesim/articles/sbc-and-calibration.html): choose what to measure.
- [Parallel runs and checkpoints](https://sims1253.github.io/bayesim/articles/parallel-and-hpc.html): run larger studies and resume interrupted work.
- [Reproducibility](https://sims1253.github.io/bayesim/articles/reproducibility.html): seeds, fingerprints, and their limits.
- [Function reference](https://sims1253.github.io/bayesim/reference/index.html): fitters, generators, metrics, and analysis tools.

Report bugs or request features in the [issue tracker](https://github.com/sims1253/bayesim/issues).
