---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd and run
     devtools::build_readme(). -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "70%",
  dpi = 150
)
set.seed(1)
```

# bayesim

<!-- badges: start -->
[![License: GPL v3](https://img.shields.io/badge/License-GPL%20v3-blue.svg)](LICENSE)
[![R-CMD-check](https://github.com/sims1253/bayesim/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/sims1253/bayesim/actions/workflows/R-CMD-check.yaml)
[![Codecov test coverage](https://codecov.io/gh/sims1253/bayesim/graph/badge.svg)](https://app.codecov.io/gh/sims1253/bayesim)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
<!-- badges: end -->

bayesim is a framework for **Bayesian simulation studies**: parameter recovery,
calibration checking (SBC), model comparison, and method evaluation — designed
for teaching, practicing, and publishing principled Bayesian workflows.

You declare a study as *data-generating conditions x models x replicates*;
bayesim plans the tasks, runs them (in parallel, resumably, with bounded
memory), and reports the performance measures a methods paper actually needs —
bias, empirical SE, MSE, and coverage, each with its Monte Carlo standard
error, following Morris, White & Crowther (2019).

## Why bayesim

- **Deterministic by construction.** Every task gets its own L'Ecuyer-CMRG RNG
  stream derived from one study seed. Sequential, parallel, and
  interrupted-then-resumed runs produce identical results.
- **Calibration is a first-class citizen.** Simulation-based calibration with
  prior-predictive and inverse-forward-sampling generators, ESS-aware rank
  thinning, and ECDF plots with simultaneous confidence bands
  (Säilynoja, Bürkner & Vehtari, 2022).
- **The right summaries, honestly computed.** `performance_measures()` reports
  estimator performance with MCSEs; LOO-based metrics match their
  loo/brms reference implementations (with parity tests to prove it).
- **Compile once, fit thousands of times.** For brms studies, a model bank
  compiles each distinct Stan model a single time and reuses the binary across
  all replicates — with a structural guard that fails loudly instead of
  silently mis-fitting.
- **Practical at scale.** Parallel execution via purrr's mirai integration,
  checkpoint/resume for long runs, and retention profiles that keep memory
  flat no matter how large the posteriors are.
- **Runs everywhere, including your laptop and your classroom.** The built-in
  `LinearRegressionFitter()` does exact conjugate Bayesian inference in
  milliseconds — every example in the documentation is *real* inference, no
  Stan installation required to learn the workflow.

## Installation

```r
# install.packages("pak")
pak::pak("sims1253/bayesim")
```

For brms- or Stan-backed studies you additionally need
[brms](https://paulbuerkner.com/brms/) and/or
[cmdstanr](https://mc-stan.org/cmdstanr/) with a working CmdStan.

## A complete study in under a minute

Does a 90% credible interval cover the truth 90% of the time, and how does
estimation error shrink with sample size? Real Bayesian inference, no Stan:

```{r example}
library(bayesim)

# 1. How data is generated. The engine restores a per-task RNG stream before
#    each call, so just consume the ambient RNG state.
gen <- function(data_spec, task_ctx) {
  x <- rnorm(data_spec$n)
  y <- 1 + 0.5 * x + rnorm(data_spec$n, sd = data_spec$sigma %||% 1)
  list(
    train = data.frame(y = y, x = x),
    response = "y",
    true_params = c(Intercept = 1, x = 0.5, sigma = data_spec$sigma %||% 1),
    vars_of_interest = c("Intercept", "x", "sigma")
  )
}

# 2. The study: 3 sample sizes x 1 model x 100 replicates = 300 tasks.
config <- simulation_config(
  data_grid = data.frame(n = c(20, 50, 200), sigma = 1),
  fit_grid = data.frame(model = "linear"),
  data_generator = gen,
  fitter = LinearRegressionFitter(),          # exact conjugate posterior
  metrics = list(
    posterior_summary_metric(prob = 0.90),
    coverage_metric(prob = 0.90)
  ),
  n_replicates = 100L,
  seed = 42L
)

# 3. Run it (drop `workers` to run sequentially).
result <- run_simulation(config, workers = 2, progress = FALSE)

# 4. Estimator performance with Monte Carlo standard errors.
performance_measures(result, by = "data_n")
```

Truths are recorded per task, so parameter recovery is one call:

```{r recovery, fig.alt = "Posterior mean estimates against true parameter values, faceted by sample size"}
plot_recovery(result, "x", by = "data_n")
```

## Calibration checking (SBC)

Is your model self-consistent — does it recover parameters drawn from its own
prior? bayesim ships the full simulation-based calibration loop:

```{r sbc, fig.alt = "SBC rank ECDF with simultaneous 95% confidence bands"}
sbc_config <- simulation_config(
  data_grid = data.frame(n = 50),
  fit_grid = data.frame(model = "linear"),
  data_generator = gen,
  fitter = LinearRegressionFitter(),
  metrics = list(rank_metric()),   # ESS-aware thinned SBC ranks
  n_replicates = 200L,
  seed = 1L
)
sbc_result <- run_simulation(sbc_config, workers = 2, progress = FALSE)

plot_rank_ecdf(sbc_ranks(sbc_result))
```

For brms models, `prior_predictive_generator()` and `ifs_generator()` draw the
truth from the prior (or a preconditioning posterior) and forward-simulate the
data — including dependency-ordered simulation for multivariate models. See
`vignette("sbc-and-calibration")`.

## Model backends

| fitter | inference | use it for |
|---|---|---|
| `LinearRegressionFitter()` | exact conjugate (NIG) posterior | learning the workflow, method studies with analytic baselines, fast pipelines |
| `BrmsFitter()` | Stan via brms, with the compile-once model bank | most applied studies; formula/family/prior grids via `model_grid()` |
| `CmdStanFitter()` | your own `.stan` files via cmdstanr | full control over the model; supply a `log_lik` generated quantity and LOO metrics work out of the box |
| your own | anything | subclass `Fitter`, implement six generics, check with `validate_fitter()` — see `vignette("custom-fitters")` |

Comparing brms models is declarative:

```r
config <- simulation_config(
  data_grid = data.frame(n = 200),
  fit_grid = model_grid(
    gaussian  = brms_model(y ~ x, brms::brmsfamily("gaussian")),
    student   = brms_model(y ~ x, brms::brmsfamily("student")),
    lognormal = brms_model(y ~ x, brms::brmsfamily("lognormal"))
  ),
  data_generator = skewed_gen,
  fitter = BrmsFitter(chains = 2L),
  metrics = list(elpd_loo_metric(), posterior_summary_metric()),
  n_replicates = 50L,
  seed = 42L
)
```

Each of the three models compiles exactly once; all 150 fits reuse the
binaries.

## Long runs: parallel, resumable, memory-bounded

```r
config <- simulation_config(
  ...,
  result_path = "results/my-study",  # checkpoints + artifacts live here
  checkpoint_every = 50L,            # batch size for checkpoint/memory cycling
  retain = "metrics"                 # keep memory flat: drop fits and draws
)

preflight(config)                    # task count, compile count, time estimate
result <- run_simulation(config, workers = 8, resume = "auto")

# After an interruption, pick up exactly where it stopped:
result <- resume_simulation("results/my-study")
```

Parallelism uses purrr's [mirai](https://mirai.r-lib.org/) integration; for
clusters, set daemons yourself (`mirai::daemons()`, including remote/TLS
daemons) instead of passing `workers` — see `vignette("parallel-and-hpc")`.

## How it compares

- **[SimDesign](https://cran.r-project.org/package=SimDesign)** /
  **[simhelpers](https://cran.r-project.org/package=simhelpers)**: excellent
  general-purpose simulation harnesses, frequentist-first. bayesim is
  Bayes-native: posteriors, calibration, LOO, and Stan-aware execution
  (model bank, per-daemon setup) are built in.
- **[rsimsum](https://cran.r-project.org/package=rsimsum)**: analyzes
  simulation results you already have. bayesim runs the study *and* reports
  the same Morris et al. performance measures (and its MCSE formulas follow
  rsimsum).
- **[SBC](https://hyunjimoon.github.io/SBC/)**: focused on calibration
  checking. bayesim embeds SBC as one metric inside a general study engine
  with checkpointing, grids, and model comparison.

## Learn more

The documentation is organized as a course:

1. `vignette("getting-started")` — your first study, end to end.
2. `vignette("design-of-simulation-studies")` — aims, estimands, choosing the
   number of replicates from a target MCSE, reporting.
3. `vignette("sbc-and-calibration")` — the calibration workflow, including a
   deliberately miscalibrated model so you can see what failure looks like.
4. `vignette("brms-studies")` — the model bank, `model_grid()`, family
   comparison case study.
5. `vignette("custom-fitters")` / `vignette("custom-metrics")` — extending the
   framework, including raw Stan via `CmdStanFitter()`.
6. `vignette("parallel-and-hpc")` and `vignette("reproducibility")` — running
   large studies and understanding the determinism guarantees.

## References

- Morris, White & Crowther (2019). Using simulation studies to evaluate
  statistical methods. *Statistics in Medicine*, 38(11).
- Talts, Betancourt, Simpson, Vehtari & Gelman (2018). Validating Bayesian
  inference algorithms with simulation-based calibration. arXiv:1804.06788.
- Säilynoja, Bürkner & Vehtari (2022). Graphical test for discrete uniformity
  and its applications in goodness-of-fit evaluation. *Statistics and
  Computing*, 32(2).
- Vehtari, Gelman & Gabry (2017). Practical Bayesian model evaluation using
  leave-one-out cross-validation and WAIC. *Statistics and Computing*, 27(5).

## Getting help

Bug reports and feature requests:
[github.com/sims1253/bayesim/issues](https://github.com/sims1253/bayesim/issues).
