---
title: "Integration with bayesfam"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integration with bayesfam}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

# bayesfam integration

> **Status: stub.** This vignette documents the *intended* workflow for
> integrating custom response families from
> [bayesfam](https://github.com/sims1253/bayesfam). bayesfam is not yet on
> CRAN and the integration API may change before release. All code chunks in
> this vignette have `eval = FALSE` so the vignette builds today; once the
> bayesfam release is available, the chunks will be switched on and the
> placeholders replaced with verified examples.

## Why this matters

`brms` supports a rich set of built-in response families, but real-world
studies often need custom families (zero-inflated mixtures, truncated or
censored variants, bespoke link functions, ...). `bayesfam` provides a
 principled way to *define* such families. Once a custom family exists,
the question every user must answer is: **does my custom family actually
fit the data it claims to model?**

Simulation-based calibration (SBC) is the acceptance test for that
question. In bayesim, SBC is just a regular simulation study with a
prior-predictive data generator: if the family + sampler are well
calibrated, the rank statistic of the true (data-generating) parameter
under the posterior is uniformly distributed. A family that fails SBC
is not safe to use for inference.

## Intended workflow

The plan, once bayesfam is released, is a four-step loop:

1. **Define** the custom family with `bayesfam` (e.g. a new mixture or
   truncated family).
2. **Wrap** it in a `brms` custom family (`brms::custom_family()`) so a
   `BrmsFitter(precompile = TRUE)` can compile it once into bayesim's
   model bank.
3. **Run SBC** as a `run_simulation()` study with
   `prior_predictive_generator()` (forward sampling from the prior) and
   the `rank_metric()` family of metrics.
4. **Check uniformity** of the resulting ranks with `plot_rank_hist()`,
   `plot_rank_ecdf()`, and the SBC acceptance bands.

## Placeholder example

```{r}
library(bayesim)
# library(bayesfam)   # not yet released

# 1. Define a custom family with bayesfam (API illustrative only).
# my_family <- bayesfam::custom_family(
#   name     = "my_mixture",
#   dpars    = c("mu", "sigma"),
#   lb       = c(NA, 0),
#   type     = "real",
#   log_lik  = function(...) { ... },
#   rng      = function(...) { ... }
# )

# 2. Wrap it for brms so BrmsFitter can compile it.
# brms_family <- brms::custom_family(
#   name     = "my_mixture",
#   dpars    = c("mu", "sigma"),
#   lb       = c(NA, 0),
#   type     = "real",
#   stanvars = my_family$stanvars
# )

# 3. Run SBC: prior-predictive generator + rank metric.
# config <- simulation_config(
#   data_grid = data.frame(n = c(50, 100, 200)),
#   fit_grid  = data.frame(family = "my_mixture"),
#   data_generator = prior_predictive_generator(
#     formula   = y ~ x,
#     family    = brms_family,
#     priors    = my_priors,
#     n_draws   = 1L
#   ),
#   fitter    = BrmsFitter(
#     formula      = bf(y ~ x, family = brms_family),
#     precompile   = TRUE,
#     sample_prior = "only"
#   ),
#   metrics = list(
#     rank_metric(),
#     coverage_metric(),
#     bias_metric()
#   ),
#   n_replicates = 100L,
#   seed         = 42L,
#   result_path  = "sbc_my_mixture"
# )
# result <- run_simulation(config, progress = TRUE)

# 4. Check uniformity of ranks.
# ranks <- sbc_ranks(result, param = "b_x")
# plot_rank_hist(ranks)
# plot_rank_ecdf(ranks)
# sbc_acceptance_bands(ranks)   # are the ranks inside the SBC bands?
```

If the rank histogram is flat and the ECDF stays within the SBC bands,
the custom family is calibrated for the configurations in `data_grid`.
If not, iterate on the family definition, the priors, or the sampler
settings and re-run.

## Pending the bayesfam release

This vignette is intentionally a stub. When `bayesfam` reaches a stable
release it will be added to `Suggests` (see `DESCRIPTION`) and the
placeholders above will be replaced with a runnable, end-to-end worked
example. Until then, the canonical reference for the SBC procedure is the
`F8` acceptance test in `tests/testthat/test-sbc-acceptance.R`, which
exercises the full SBC pipeline for built-in families and demonstrates
the rank-uniformity checks described here.
