---
title: "brms studies: the model bank and model_grid()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{brms studies: the model bank and model_grid()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # every chunk needs brms + a working CmdStan; shown, not run
)
```

```{r setup, eval = TRUE}
library(bayesim)
```

This vignette covers running brms-backed simulation studies: how the model
bank eliminates per-task compilation, how to declare model-comparison grids
with `model_grid()` / `brms_model()`, how to pass sampler arguments, and how
to keep memory flat on long runs. All chunks require brms and a working
CmdStan install, so they are shown but not evaluated here.

## The model bank: compile once, fit thousands of times

A naive brms simulation study recompiles the Stan model for every task —
minutes of C++ compilation per seconds of sampling. `BrmsFitter()` avoids this
with a *model bank*: at the start of `run_simulation()`, each distinct model
spec in the `fit_grid` is compiled once (via `brms::brm(chains = 0)`), and
every task reuses the compiled binary through
`stats::update(recompile = FALSE)`.

Two properties control this:

- `precompile` (default `TRUE`): build the bank at run start. Set to `FALSE`
  to fall back to a fresh `brms::brm()` per task (only sensible for tiny
  studies or debugging).
- The bank is shipped to mirai daemons once per run, so parallel workers reuse
  the same binaries.

brms does **not** warn when `recompile = FALSE` is used against structurally
incompatible data — it would silently reuse the binary against the wrong model
frame. bayesim therefore compares the Stan data *structure* (via
`brms::make_standata()`) between the compiled template and each task's data,
and aborts the run loudly on a mismatch. If your `data_grid` rows produce data
with different shapes (e.g. varying factor levels), either make them
structurally identical or set `precompile = FALSE`.

Some brms default priors are data-dependent (for example, the intercept prior
can be centered using the response in the template dataset). Those constants
are embedded in the compiled model and cannot be refreshed by
`update(recompile = FALSE)`. Therefore, always supply explicit priors in every
`brms_model()` when `precompile = TRUE`; bayesim raises a fatal configuration
error when a model-bank row omits them (opt in with
`BrmsFitter(allow_default_priors = TRUE)` if you deliberately want
template-derived priors). Use `precompile = FALSE` if task-specific default
priors are part of the study design.

## Declaring models: `brms_model()` and `model_grid()`

Hand-building list-columns (`fit_grid$formula <- list(...)`) works but is
error-prone. `model_grid()` assembles a tidy fit grid from named
`brms_model()` specs, validating each at construction so mistakes surface
before any compilation:

```{r, eval=FALSE}
explicit_priors <- c(
  brms::prior(normal(0, 2), class = "b"),
  brms::prior(normal(0, 5), class = "Intercept"),
  brms::prior(exponential(1), class = "sigma")
)
fit_grid <- model_grid(
  gaussian  = brms_model(y ~ x, brms::brmsfamily("gaussian"), explicit_priors),
  student   = brms_model(y ~ x, brms::brmsfamily("student"), explicit_priors),
  lognormal = brms_model(y ~ x, brms::brmsfamily("lognormal"), explicit_priors)
)
fit_grid
#> # A tibble: 3 x 5
#>   model     formula   family   prior  stanvars
#>   <chr>     <list>    <list>   <list> <list>
#> 1 gaussian  <formula> <family> <prior> <NULL>
#> 2 student   <formula> <family> <prior> <NULL>
#> 3 lognormal <formula> <family> <prior> <NULL>
```

The `model` name column lands in the result summary as `fit_model`, so
per-model aggregation is `summarize_simulation(result)` or
`performance_measures(result)` with no extra arguments.

## Case study: comparing likelihoods for skewed data

Which likelihood best describes right-skewed data — Gaussian, Student-t, or
lognormal? Generate skewed data, fit all three models, and compare expected
log predictive density (ELPD):

```{r, eval=FALSE}
skew_generator <- function(data_spec, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(n)
  mu <- data_spec$beta * x
  y <- exp(mu + stats::rnorm(n, sd = 0.5))
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(beta = data_spec$beta),
    vars_of_interest = "beta"
  )
}

config <- simulation_config(
  data_grid = data.frame(n = 200, beta = 1),
  fit_grid = model_grid(
    gaussian = brms_model(
      y ~ x, brms::brmsfamily("gaussian"), explicit_priors
    ),
    student = brms_model(
      y ~ x, brms::brmsfamily("student"), explicit_priors
    ),
    lognormal = brms_model(
      y ~ x, brms::brmsfamily("lognormal"), explicit_priors
    )
  ),
  data_generator = skew_generator,
  fitter = BrmsFitter(chains = 2L, iter = 1000L, warmup = 500L),
  metrics = list(
    elpd_loo_metric(),
    posterior_summary_metric(),
    sampler_diagnostics_metric()
  ),
  n_replicates = 50L,
  seed = 42L
)

result <- run_simulation(config, workers = 4)
summarize_simulation(result, metrics = "elpd_loo__elpd")
```

Three models compile exactly once each; all 150 fits reuse the binaries.

## Sampler arguments: `stan_args`

`BrmsFitter(stan_args = ...)` passes sampler controls through to brms/Stan.
`adapt_delta` and `max_treedepth` are mapped into the `control` list;
`init` and `threads` pass through directly:

```{r, eval=FALSE}
fitter <- BrmsFitter(
  chains = 4L,
  stan_args = list(adapt_delta = 0.95, max_treedepth = 12, init = 0.1)
)
```

## Warning-conditional retention

Diagnosing occasional divergent transitions across thousands of tasks needs
the fit object — but only for the tasks that misbehaved. Retention specs can
be conditional on warnings, keeping heavy artifacts only where something went
wrong:

```{r, eval=FALSE}
config <- simulation_config(
  ...,
  retain = list(
    success = c("metrics", "diagnostics"),
    warning = c("metrics", "diagnostics", "fit", "draws")
  )
)
```

Clean tasks stay light; tasks that emitted sampler warnings keep their full
fit for post-hoc inspection.

## SBC with brms generators

For calibration checking of brms models, `prior_predictive_generator()` draws
the truth from the model prior (a `sample_prior = "only"` fit) and
`ifs_generator()` draws it from a preconditioning posterior. Both
forward-simulate the response — including dependency-ordered simulation for
multivariate models. See `vignette("sbc-and-calibration")` for the full
workflow.

## Next steps

- `vignette("parallel-and-hpc")` — daemons, checkpoint/resume, memory.
- `vignette("custom-fitters")` — raw Stan via `CmdStanFitter()` when brms
  cannot express your model.
- `vignette("reproducibility")` — what the config fingerprint covers.
