---
title: "brms studies: the model bank and model_grid()"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{brms studies: the model bank and model_grid()}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # every chunk needs brms + a working CmdStan; shown, not run
)
```

```{r setup, eval = TRUE}
library(bayesim)
```

Use `BrmsFitter()` to run a study with brms models. The examples require brms
and a working CmdStan installation; they are shown without running during
the documentation build.

## The model bank: compile once, fit thousands of times

With `precompile = TRUE` (the default), `BrmsFitter()` compiles each distinct
model specification at the start of a run. Tasks reuse these compiled models,
including on parallel workers. Set `precompile = FALSE` to fit each task
with a fresh `brms::brm()` call.

brms does **not** warn when `recompile = FALSE` is used against structurally
incompatible data. It can reuse the binary against the wrong model
frame. bayesim therefore compares the Stan data *structure* (via
`brms::make_standata()`) between the compiled template and each task's data,
and stops the run on a mismatch. If your `data_grid` rows produce data
with different shapes (e.g. varying factor levels), either make them
structurally identical or set `precompile = FALSE`.

Some brms default priors are data-dependent (for example, the intercept prior
can be centered using the response in the template dataset). Those constants
are embedded in the compiled model and cannot be refreshed by
`update(recompile = FALSE)`. Therefore, always supply explicit priors in every
`brms_model()` when `precompile = TRUE`; bayesim raises a fatal configuration
error when a model-bank row omits them (opt in with
`BrmsFitter(allow_default_priors = TRUE)` if you deliberately want
template-derived priors). Use `precompile = FALSE` if task-specific default
priors are part of the study design.

## Declaring models: `brms_model()` and `model_grid()`

Hand-building list-columns (`fit_grid$formula <- list(...)`) works but is
error-prone. `model_grid()` assembles a tidy fit grid from named
`brms_model()` specs, validating each at construction so mistakes surface
before any compilation:

```{r, eval=FALSE}
explicit_priors <- c(
  brms::prior(normal(0, 2), class = "b"),
  brms::prior(normal(0, 5), class = "Intercept"),
  brms::prior(exponential(1), class = "sigma")
)
fit_grid <- model_grid(
  gaussian  = brms_model(y ~ x, brms::brmsfamily("gaussian"), explicit_priors),
  student   = brms_model(y ~ x, brms::brmsfamily("student"), explicit_priors),
  lognormal = brms_model(y ~ x, brms::brmsfamily("lognormal"), explicit_priors)
)
fit_grid
#> # A tibble: 3 x 5
#>   model     formula   family   prior  stanvars
#>   <chr>     <list>    <list>   <list> <list>
#> 1 gaussian  <formula> <family> <prior> <NULL>
#> 2 student   <formula> <family> <prior> <NULL>
#> 3 lognormal <formula> <family> <prior> <NULL>
```

The `model` name column lands in the result summary as `fit_model`, so
per-model aggregation is `summarize_simulation(result)` or
`performance_measures(result)` with no extra arguments.

## Case study: comparing likelihoods for skewed data

Which likelihood best describes right-skewed data: Gaussian, Student-t, or
lognormal? Generate skewed data, fit all three models, and compare expected
log predictive density (ELPD):

```{r, eval=FALSE}
skew_generator <- function(data_spec, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(n)
  mu <- data_spec$beta * x
  y <- exp(mu + stats::rnorm(n, sd = 0.5))
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(beta = data_spec$beta),
    vars_of_interest = "beta"
  )
}

config <- simulation_config(
  data_grid = data.frame(n = 200, beta = 1),
  fit_grid = model_grid(
    gaussian = brms_model(
      y ~ x, brms::brmsfamily("gaussian"), explicit_priors
    ),
    student = brms_model(
      y ~ x, brms::brmsfamily("student"), explicit_priors
    ),
    lognormal = brms_model(
      y ~ x, brms::brmsfamily("lognormal"), explicit_priors
    )
  ),
  data_generator = skew_generator,
  fitter = BrmsFitter(chains = 2L, iter = 1000L, warmup = 500L),
  metrics = list(
    elpd_loo_metric(),
    posterior_summary_metric(),
    sampler_diagnostics_metric()
  ),
  n_replicates = 50L,
  seed = 42L
)

result <- run_simulation(config, workers = 4)
summarize_simulation(result, metrics = "elpd_loo__elpd")
```

Three models compile exactly once each; all 150 fits reuse the binaries.

With `newdata = NULL`, brms predictions and log-likelihoods use the fitted
model's stored data. This preserves fitted latent variables for LOO metrics;
passing a data frame explicitly requests brms' new-data prediction behavior.

`BrmsFitter()` rejects fits that silently drop training rows. Handle missing
values or row filtering in the data generator, or specify a brms missing-data
model that retains the rows. The training response must align with the fitted
observations. The usual brms requirements for each model still apply, including
saving latent parameters when needed for prediction.

## Sampler arguments: `stan_args`

`BrmsFitter(stan_args = ...)` passes sampler controls through to brms/Stan.
`adapt_delta` and `max_treedepth` are mapped into the `control` list;
`init` and `threads` pass through directly:

```{r, eval=FALSE}
fitter <- BrmsFitter(
  chains = 4L,
  stan_args = list(adapt_delta = 0.95, max_treedepth = 12, init = 0.1)
)
```

## Warning-conditional retention

Keep fit objects for tasks that emit warnings by setting a conditional
retention policy:

```{r, eval=FALSE}
config <- simulation_config(
  ...,
  retain = list(
    success = c("metrics", "diagnostics"),
    warning = c("metrics", "diagnostics", "fit", "draws")
  )
)
```

Clean tasks stay light; tasks that emitted sampler warnings keep their full
fit for post-hoc inspection.

## SBC with brms generators

For calibration checking of brms models, `prior_predictive_generator()` draws
the truth from the model prior (a `sample_prior = "only"` fit) and
`ifs_generator()` draws it from a preconditioning posterior. Both
forward-simulate the response, including dependency-ordered simulation for
multivariate models. See `vignette("sbc-and-calibration")` for the full
workflow.

## Next steps

- `vignette("parallel-and-hpc")`: daemons, checkpoint/resume, memory.
- `vignette("custom-fitters")`: raw Stan via `CmdStanFitter()` when brms
  cannot express your model.
- `vignette("reproducibility")`: what the config fingerprint covers.
