---
title: "Case Studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Case studies

This vignette presents two case studies: an executable MockFitter-based power
sweep, and a sketch of a brms-based model-comparison study (code shown,
evaluated only where Stan is available).

## Case study 1: power vs sample size (executable)

We sweep sample size and effect size to see how posterior uncertainty and
recovery behave. This runs with the built-in `MockFitter()`.

```{r}
library(bayesim)

power_generator <- function(data_spec, seed, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(n)
  y <- data_spec$beta * x + stats::rnorm(n, sd = 1)
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(beta = data_spec$beta),
    vars_of_interest = "beta"
  )
}

config <- simulation_config(
  data_grid = data.frame(
    n = c(30, 60, 120),
    beta = c(0, 0.5, 1)
  ),
  fit_grid = data.frame(model = "linear"),
  data_generator = power_generator,
  fitter = MockFitter(),
  metrics = list(rmse_metric(), bias_metric(), posterior_summary_metric()),
  n_replicates = 4L,
  seed = 7L
)

result <- run_simulation(config, progress = FALSE)
```

Aggregate by condition to see how the posterior mean and spread of `beta`
vary with `n` and effect size:

```{r}
agg <- summarize_simulation(result, by = c("n", "beta"))
agg
```

## Case study 2: comparing brms families (sketch)

This study compares Gaussian, Student-t, and lognormal likelihoods for
right-skewed data. It requires brms + cmdstanr; the chunks are shown but not
evaluated in this vignette.

```{r, eval=FALSE}
# Define a generator producing right-skewed data.
skew_generator <- function(data_spec, seed, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(n)
  mu <- data_spec$beta * x
  y <- mu + stats::rgamma(n, shape = 2, scale = 1) # right-skewed noise
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(beta = data_spec$beta),
    vars_of_interest = "beta"
  )
}

# The fit grid carries brms objects as list-columns. The model bank compiles
# each distinct (formula, family) combination once.
fit_grid <- data.frame(model = c("gaussian", "student", "lognormal"))
fit_grid$formula <- list(y ~ x, y ~ x, y ~ x)
fit_grid$family <- list(
  brms::brmsfamily("gaussian"),
  brms::brmsfamily("student"),
  brms::brmsfamily("lognormal")
)

config <- simulation_config(
  data_grid = data.frame(n = 200, beta = 1),
  fit_grid = fit_grid,
  data_generator = skew_generator,
  fitter = BrmsFitter(chains = 2L, iter = 1000L, warmup = 500L),
  metrics = list(
    rmse_metric(),
    bias_metric(),
    posterior_summary_metric(),
    elpd_loo_metric()
  ),
  n_replicates = 10L,
  seed = 42L
)

result <- run_simulation(config, progress = TRUE)
agg <- summarize_simulation(result, by = "model")
```

Because `BrmsFitter` uses the model bank, the three Stan models are compiled
once each at the start of the run and reused across all replicates.

## Case study 3: Simulation-Based Calibration (sketch)

SBC checks whether a model is calibrated by ranking the data-generating
parameter against the posterior draws. Use `ifs_generator()` (inverse forward
sampling) or `prior_predictive_generator()` and the `rank_metric()`:

```{r, eval=FALSE}
# Compile a preconditioning fit on pilot data (posterior, not prior-only).
prefit <- brms::brm(y ~ x, data = pilot_data, backend = "cmdstanr",
                    chains = 2L, iter = 1000L, warmup = 500L)

config <- simulation_config(
  data_grid = data.frame(task = "sbc"),
  fit_grid = data.frame(model = "gaussian", formula = I(list(y ~ x)),
                        family = I(list(brms::brmsfamily("gaussian")))),
  data_generator = ifs_generator(
    prefit = prefit,
    vars_of_interest = "b_x"
  ),
  fitter = BrmsFitter(),
  metrics = list(rank_metric()),
  n_replicates = 200L, # enough replicates for the rank histogram
  seed = 1L
)

result <- run_simulation(config, progress = TRUE)

# Rank histograms + ECDF with simultaneous confidence bands.
plot_rank_hist(sbc_ranks(result))
plot_rank_ecdf(sbc_ranks(result))
```

Under a correctly calibrated model, the rank histogram is approximately flat
and the ECDF stays within the confidence band.
