---
title: "Case Studies: Simulation-Based Research with bayesim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Case Studies: Simulation-Based Research with bayesim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE
)
```

```{r setup}
library(bayesim)
library(dplyr)
library(ggplot2)
```

## Introduction

This vignette demonstrates two case studies from the simulation-based Bayesian model validation literature:

1. **Prediction vs Explanation** (arXiv:2210.06927): When does good predictive performance imply good parameter recovery?

2. **Primed Priors** (arXiv:2408.06504): Using calibration data to generate realistic simulation scenarios without hand-specifying priors.

::: {.callout-warning}
All code chunks that require `brms` and Stan use `eval=FALSE`. To run these examples, you need:

- `brms` package installed
- `cmdstanr` or `rstan` as the Stan backend  
- A working C++ toolchain
:::

---

## Case Study 1: Prediction as a Proxy for Explanation

### Background

The paper (arXiv:2210.06927) investigates when predictive performance (PP) can serve as a proxy for parameter recoverability (PR) in Bayesian GLMs. The question: if a model predicts well, can we trust its parameter estimates?

### The Causal Structure

We work with a fixed causal directed acyclic graph (DAG):

```
    z3
     |
     v
     x -----> y
     ^        ^
     |        |
    z1       z2
     |        |
     +--------+
     
     x -----> z4 <----- y
```

**Variable roles:**

| Variable | Role | Relationships |
|----------|------|---------------|
| `x` | Treatment | `x → y` (causal effect of interest) |
| `y` | Outcome | Target of prediction |
| `z1` | Fork/Confounder | `z1 → x` and `z1 → y` |
| `z2` | Ancestor of y only | `z2 → y` |
| `z3` | Ancestor of x only | `z3 → x` |
| `z4` | Collider | `x → z4 ← y` (NOT adjusted for) |

The target parameter is `b_x`: the causal effect of `x` on `y`. The correct adjustment set is `{z1, z2}` (not `z3` or `z4`).

### Data Generator with Varying Likelihood/Link

The generator preserves the DAG structure while varying the outcome distribution:

```{r case1-data-generator}
generate_case1_data <- function(data_spec, seed, task_ctx) {
  # Use a shared seed so all models fit to the same dataset
  # within a replicate see identical data
  shared_seed <- as.integer(
    data_spec$base_seed + 100000L * task_ctx$data_idx + task_ctx$rep_idx
  )
  set.seed(shared_seed)

  n <- data_spec$n

  # --- Generate covariates following the DAG ---
  z1 <- rnorm(n)                    # Confounder
  z2 <- rnorm(n)                    # Ancestor of y only
  z3 <- rnorm(n)                    # Ancestor of x only
  
  # x depends on z1 and z3 (not z2)
  x <- 0.7 * z1 + 0.8 * z3 + rnorm(n, sd = 0.6)

  # Linear predictor for y (correct adjustment: z1, z2)
  b0  <- data_spec$b0
  bx  <- data_spec$bx   # True causal effect
  bz1 <- data_spec$bz1
  bz2 <- data_spec$bz2
  eta <- b0 + bx * x + bz1 * z1 + bz2 * z2

  # --- Generate outcome from selected DGP ---
  if (identical(data_spec$dgp_family, "gaussian")) {
    sigma <- data_spec$sigma
    y <- rnorm(n, mean = eta, sd = sigma)
    extra_true <- c(sigma = sigma)
    
  } else if (identical(data_spec$dgp_family, "gamma")) {
    shape <- data_spec$shape
    mu <- exp(eta)  # log link
    y <- rgamma(n, shape = shape, rate = shape / mu)
    extra_true <- c(shape = shape)
    
  } else if (identical(data_spec$dgp_family, "beta")) {
    phi <- data_spec$phi
    mu <- plogis(eta)  # logit link
    mu <- pmin(pmax(mu, 1e-4), 1 - 1e-4)  # Bound away from 0,1
    y <- rbeta(n, shape1 = mu * phi, shape2 = (1 - mu) * phi)
    extra_true <- c(phi = phi)
    
  } else {
    stop("Unsupported dgp_family: ", data_spec$dgp_family)
  }

  # Collider (generated for completeness, NOT used in fitting)
  z4 <- 0.6 * x + 0.6 * y + rnorm(n, sd = 0.4)

  true_params <- c(
    b_Intercept = b0,
    b_x = bx,
    b_z1 = bz1,
    b_z2 = bz2,
    extra_true
  )

  list(
    train = data.frame(y = y, x = x, z1 = z1, z2 = z2, z3 = z3, z4 = z4),
    test = NULL,
    response = "y",
    true_params = true_params,
    vars_of_interest = names(true_params),
    references = setNames(rep(0, length(true_params)), names(true_params)),
    meta = list(
      dgp_family = data_spec$dgp_family,
      dag = "z3->x->y; z1->x; z1->y; z2->y; x->z4<-y"
    )
  )
}
```

### Fit Grid: Multiple Families on Same Data

We fit three different models to each dataset, all with the same formula but different likelihoods:

```{r case1-fit-grid, eval=FALSE}
library(brms)

fit_grid_case1 <- tibble::tibble(
  model_id = c("gaussian", "gamma", "beta"),
  formula = list(
    bf(y ~ x + z1 + z2),
    bf(y ~ x + z1 + z2),
    bf(y ~ x + z1 + z2)
  ),
  family = list(
    gaussian(),
    Gamma(link = "log"),
    Beta()
  )
)
```

### Custom Metrics for PP and PR

We define two custom metrics:

- **PP (Predictive Performance)**: LOO-ELPD from leave-one-out cross-validation
- **PR (Parameter Recovery)**: Absolute error in recovering `b_x`, the true causal effect

```{r case1-metrics}
# PP: LOO-ELPD metric (needs LOO computation in context)
LooElpdMetric <- S7::new_class(
  "LooElpdMetric",
  parent = Metric,
  properties = list(
    name = S7::new_property(S7::class_character, default = "loo_elpd"),
    needs = S7::new_property(S7::class_character, default = "loo"),
    required = S7::new_property(S7::class_logical, default = FALSE)
  )
)

S7::method(compute, LooElpdMetric) <- function(metric, fit_result, data_bundle, context, task_ctx) {
  if (is.null(context$loo)) {
    return(list(elpd = NA_real_, elpd_se = NA_real_, p_loo = NA_real_))
  }
  list(
    elpd = context$loo$elpd,
    elpd_se = context$loo$elpd_se,
    p_loo = context$loo$p_loo
  )
}

# PR: Recovery of the causal effect b_x
BxRecoveryMetric <- S7::new_class(
  "BxRecoveryMetric",
  parent = Metric,
  properties = list(
    name = S7::new_property(S7::class_character, default = "bx_recovery"),
    needs = S7::new_property(S7::class_character, default = character()),
    required = S7::new_property(S7::class_logical, default = FALSE)
  )
)

S7::method(compute, BxRecoveryMetric) <- function(metric, fit_result, data_bundle, context, task_ctx) {
  if (is.null(fit_result$draws) || !("b_x" %in% colnames(fit_result$draws))) {
    return(list(abs_error = NA_real_, sq_error = NA_real_, bias = NA_real_))
  }

  est <- mean(fit_result$draws[, "b_x"])
  truth <- unname(data_bundle$true_params[["b_x"]])

  list(
    abs_error = abs(est - truth),
    sq_error = (est - truth)^2,
    bias = est - truth
  )
}
```

### Simulation Configuration

Small-scale illustration with 3 DGP families × 3 fit models × 8 replicates:

```{r case1-config, eval=FALSE}
config_case1 <- simulation_config(
  data_grid = data.frame(
    n = 120,
    dgp_family = c("gaussian", "gamma", "beta"),
    b0 = c(0.4, 0.2, -0.2),
    bx = c(0.6, 0.5, 0.8),
    bz1 = 0.3,
    bz2 = 0.4,
    sigma = 0.7,
    shape = 8,
    phi = 25,
    base_seed = 42L,
    stringsAsFactors = FALSE
  ),
  fit_grid = fit_grid_case1,
  data_generator = generate_case1_data,
  fitter = BrmsFitter(
    backend = "cmdstanr",
    chains = 2L,
    iter = 1000L,
    warmup = 500L,
    refresh = 0L
  ),
  metrics = list(
    LooElpdMetric(),
    BxRecoveryMetric(),
    coverage_metric(),
    rmse_metric()
  ),
  n_replicates = 8L,
  seed = 101L
)

result_case1 <- run_simulation(config_case1, progress = TRUE)
```

### Analysis: When Do PP and PR Agree?

The key analysis compares whether the model with best PP also has best PR:

```{r case1-analysis, eval=FALSE}
# Extract summary and identify replicates
summary_case1 <- result_case1$summary |>
  mutate(
    dgp = paste0("DGP: ", data_dgp_family),
    rep_id = as.integer(gsub(".*_r(\\d+)", "\\1", task_id))
  )

# For each DGP and replicate, find PP winner and PR winner
agreement_tbl <- summary_case1 |>
  group_by(dgp, rep_id) |>
  summarise(
    winner_pp = fit_model_id[which.max(loo_elpd__elpd)],
    winner_pr = fit_model_id[which.min(bx_recovery__abs_error)],
    agree = winner_pp == winner_pr,
    .groups = "drop"
  )

# Agreement rate by DGP
agreement_tbl |>
  group_by(dgp) |>
  summarise(
    agreement_rate = mean(agree, na.rm = TRUE),
    n_replicates = n(),
    .groups = "drop"
  )
```

### Visualization

```{r case1-plot, fig.width=9, fig.height=5, eval=FALSE}
summary_case1 |>
  ggplot(aes(
    x = loo_elpd__elpd,
    y = bx_recovery__abs_error,
    color = fit_model_id
  )) +
  geom_point(alpha = 0.7, size = 2) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.8, alpha = 0.5) +
  facet_wrap(~ dgp, scales = "free") +
  labs(
    title = "Predictive Performance vs Parameter Recovery",
    subtitle = "Each point is one model fit; lower right is ideal",
    x = "LOO-ELPD (higher = better prediction)",
    y = "|Error in b_x| (lower = better recovery)",
    color = "Fit Model"
  ) +
  scale_color_brewer(palette = "Set1") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

**Expected pattern:** When the fitted model matches the data-generating family, PP and PR tend to agree. Misspecification (e.g., fitting gamma to gaussian data) can decouple them—good PP but poor PR, or vice versa.

---

## Case Study 2: Primed Priors for Simulation-Based Calibration

### Background

Simulation-based calibration (SBC) requires specifying priors that define the "true" parameter values. The paper (arXiv:2408.06504) proposes **primed priors**: use a calibration dataset to learn a realistic prior distribution, then sample from the posterior to generate simulation scenarios.

This avoids hand-specifying priors while ensuring simulation parameters are realistic for your application domain.

### The Pattern

The primed prior workflow:

1. **Calibration Data**: A real dataset representative of your application
2. **Prime**: Fit the model to a subset of calibration data
3. **Draw**: Sample one parameter set from the posterior
4. **Simulate**: Generate new outcome data from posterior predictive
5. **True params**: The drawn parameters become the "ground truth" for SBC

### Primed Prior Generator

This generator implements the primed prior pattern:

```{r case2-primed-generator, eval=FALSE}
library(brms)
library(posterior)

make_primed_prior_generator <- function(
    calibration_data,
    model_formula,
    model_family,
    calibration_n = 20L,
    model_priors = NULL
) {
  # Cache for the priming fit (fitted once, reused across replicates)
  cache <- new.env(parent = emptyenv())

  function(data_spec, seed, task_ctx) {
    # --- Step 1: Prime once (cached) ---
    if (is.null(cache$priming_fit)) {
      set.seed(data_spec$prime_seed)

      # Sample subset of calibration data
      cal_idx <- sample(seq_len(nrow(calibration_data)), calibration_n)
      cal_subset <- calibration_data[cal_idx, , drop = FALSE]

      # Fit model to calibration data (this "primes" the prior)
      cache$priming_fit <- brms::brm(
        formula = model_formula,
        family = model_family,
        data = cal_subset,
        prior = model_priors %||% c(
          prior(normal(0, 5), class = "b"),
          prior(normal(0, 5), class = "Intercept")
        ),
        chains = 2,
        iter = 1000,
        warmup = 500,
        refresh = 0,
        seed = data_spec$prime_seed
      )

      cache$draws_df <- posterior::as_draws_df(cache$priming_fit)
    }

    # --- Step 2: Build covariates for new simulated dataset ---
    # Note: seed is a scalar task seed.
    # The simulation engine also restores the task RNG stream before each call,
    # so the RNG is already in the correct state for this task.
    n <- data_spec$n
    new_idx <- sample(seq_len(nrow(calibration_data)), n, replace = TRUE)

    # Extract covariate columns (everything except response)
    resp_var <- all.vars(model_formula[[2]])[1]
    cov_cols <- setdiff(names(calibration_data), resp_var)
    new_covariates <- calibration_data[new_idx, cov_cols, drop = FALSE]

    # --- Step 3: Draw one parameter set from primed posterior ---
    draw_idx <- sample(seq_len(nrow(cache$draws_df)), 1)
    draw_params <- cache$draws_df[draw_idx, , drop = FALSE]

    # --- Step 4: Simulate new outcome from posterior predictive ---
    y_sim <- as.integer(
      brms::posterior_predict(
        cache$priming_fit,
        newdata = new_covariates,
        draw_ids = draw_idx
      )[1, ]
    )

    # --- Step 5: Extract true params from the draw ---
    param_cols <- intersect(
      c("b_Intercept", paste0("b_", cov_cols), "shape", "phi", "sigma"),
      colnames(cache$draws_df)
    )
    true_params <- as.numeric(draw_params[1, param_cols])
    names(true_params) <- param_cols

    # Assemble training data
    train_data <- cbind(
      setNames(data.frame(y_sim), resp_var),
      new_covariates
    )

    list(
      train = train_data,
      test = NULL,
      response = resp_var,
      true_params = true_params,
      vars_of_interest = names(true_params),
      references = setNames(rep(0, length(true_params)), names(true_params)),
      meta = list(
        primed = TRUE,
        calibration_n = calibration_n,
        draw_idx = draw_idx
      )
    )
  }
}
```

### Example: Count Data with Negative Binomial

Using the `epilepsy` dataset from `brms` as calibration data:

```{r case2-calibration-data, eval=FALSE}
# Use brms epilepsy data as calibration source
calibration_data <- brms::epilepsy |>
  select(count, Base, Age) |>
  mutate(
    count = as.integer(count),
    Base = scale(Base)[, 1],
    Age = scale(Age)[, 1]
  )

# Create the primed generator
primed_generator <- make_primed_prior_generator(
  calibration_data = calibration_data,
  model_formula = bf(count ~ Base + Age),
  model_family = negbinomial(link = "log"),
  calibration_n = 25L
)
```

### Simulation Configuration

Compare negative binomial vs Poisson (overdispersion misspecification):

```{r case2-config, eval=FALSE}
fit_grid_case2 <- tibble::tibble(
  model_id = c("negbinomial", "poisson"),
  formula = list(
    bf(count ~ Base + Age),
    bf(count ~ Base + Age)
  ),
  family = list(
    negbinomial(link = "log"),
    poisson(link = "log")
  )
)

config_case2 <- simulation_config(
  data_grid = data.frame(
    n = c(40, 80),
    prime_seed = 2024L
  ),
  fit_grid = fit_grid_case2,
  data_generator = primed_generator,
  fitter = BrmsFitter(
    backend = "cmdstanr",
    chains = 2L,
    iter = 1000L,
    warmup = 500L,
    refresh = 0L
  ),
  metrics = list(
    LooElpdMetric(),
    coverage_metric(prob = 0.95),
    posterior_mean_metric(),
    rmse_metric()
  ),
  n_replicates = 6L,
  seed = 77L
)

result_case2 <- run_simulation(config_case2, progress = TRUE)
```

### Why Primed Priors Enable SBC

Traditional SBC requires:

1. Hand-specify prior distributions
2. Sample "true" parameters from prior
3. Simulate data given parameters
4. Fit model and check coverage

**The challenge:** Hand-specified priors may be unrealistic for your application, leading to simulation scenarios that do not reflect real-world performance.

**Primed priors solution:** The calibration dataset defines what "realistic" means for your domain. The priming fit learns posterior distributions that are:
- Data-informed (from calibration data)
- Realistic for the model family
- Properly correlated across parameters

Each simulated dataset carries the parameter values used to generate it (`true_params`), enabling coverage checks.

### Post-Run Analysis

```{r case2-analysis, eval=FALSE}
# Summary by model and sample size
result_case2$summary |>
  group_by(fit_model_id, data_n) |>
  summarise(
    mean_elpd = mean(loo_elpd__elpd, na.rm = TRUE),
    mean_coverage = mean(coverage__mean, na.rm = TRUE),
    mean_rmse = mean(rmse__value, na.rm = TRUE),
    .groups = "drop"
  )

# Coverage by parameter
result_case2$summary |>
  select(fit_model_id, data_n, starts_with("coverage__")) |>
  pivot_longer(
    cols = starts_with("coverage__"),
    names_to = "param",
    values_to = "coverage"
  ) |>
  filter(!is.na(coverage)) |>
  group_by(fit_model_id, param) |>
  summarise(mean_coverage = mean(coverage), .groups = "drop") |>
  ggplot(aes(x = param, y = mean_coverage, fill = fit_model_id)) +
  geom_col(position = "dodge") +
  geom_hline(yintercept = 0.95, linetype = "dashed", color = "red") +
  labs(
    title = "Coverage by Parameter and Model",
    x = "Parameter",
    y = "Mean Coverage",
    fill = "Model"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

---

## Implementation Notes

### For CRAN Compliance

All `brms`-dependent chunks have `eval=FALSE` because:

1. `brms` has heavy dependencies (Stan, C++ toolchain)
2. Vignette builds must work on CRAN machines without these
3. Users can copy-paste code to run locally

### Design Patterns Used

1. **Shared seeds in Case Study 1**: Ensures all models fit to the same dataset within a replicate for fair comparison

2. **Closure caching in Case Study 2**: The priming fit is cached inside the generator closure, computed once and reused

3. **Metric composition**: Custom `LooElpdMetric` and `BxRecoveryMetric` alongside built-in `coverage_metric()`, `rmse_metric()`

### Scaling Up

For a full research study, increase:

- `n_replicates`: 100-1000 for stable estimates
- `chains`: 4 for better MCMC diagnostics
- `iter` / `warmup`: 2000/1000 for more precise posteriors
- `data_grid` conditions: more sample sizes, effect sizes, etc.

---

## References

1. **Prediction vs Explanation**: Lundberg, I., et al. (2022). "Prediction can be safely used as a proxy for explanation in causally consistent Bayesian generalized linear models." arXiv:2210.06927

2. **Primed Priors**: Schürmann, C., et al. (2024). "Primed priors for simulation-based validation of Bayesian models." arXiv:2408.06504
