---
title: "Custom Fitters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Fitters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Custom Fitters

bayesim ships three fitters: `LinearRegressionFitter()` (exact conjugate
inference, no Stan), `BrmsFitter()` (Stan via brms, with the compile-once
model bank), and `CmdStanFitter()` (your own `.stan` programs via cmdstanr).
To use a different backend, implement a custom fitter by extending the
`Fitter` class and providing methods for the required generics.

## The Fitter contract

Every fitter must implement these S7 generics:

| generic | signature | returns |
|---|---|---|
| `fit_model(fitter, data_bundle, fit_spec, seed, task_ctx)` | training data + spec | a `bayesim_fit_result` (use `new_fit_result()` internally) |
| `extract_draws(fitter, fit_result, variables)` | a fit result | a draws matrix (rows = draws, cols = parameters, with colnames) |
| `predict_fit(fitter, fit_result, newdata, seed)` | a fit result | list with `predicted_mean` (length N), `predicted_samples` (S x N), `predicted_sd` (length N) |
| `log_lik_matrix(fitter, fit_result, newdata)` | a fit result | log-likelihood matrix (S x N) |
| `loo_fit(fitter, fit_result)` | a fit result | list with `elpd`, `p_loo`, `elpd_se`, `pareto_k` |
| `fit_diagnostics(fitter, fit_result)` | a fit result | list with `rhat_max`, `ess_bulk_min`, `divergent`, ... |

**All matrices are draws x observations (S x N)** — `predicted_samples`,
`log_lik_matrix()`, and `predict_epred()` all put posterior draws in rows and
observations in columns, matching the brms/loo convention. This is the single
most common custom-fitter bug; `validate_fitter(smoke_test = TRUE)` rejects
transposed matrices.

The generics avoid masking common names: `loo_fit` (not `loo::loo`),
`log_lik_matrix` (not `brms::log_lik`), `fit_model` (not `generics::fit`).

## A minimal fitter: LinearFitter

This fitter fits a linear model by OLS and synthesizes posterior draws by
bootstrapping the residual variance. It is executable without Stan.

```{r}
library(bayesim)

# Define the class
LinearFitter <- S7::new_class(
  "LinearFitter",
  parent = Fitter,
  properties = list(
    name = S7::new_property(S7::class_character, default = "linear"),
    supports_predictions = S7::new_property(S7::class_logical, default = TRUE),
    supports_log_lik = S7::new_property(S7::class_logical, default = TRUE),
    supports_loo = S7::new_property(S7::class_logical, default = FALSE),
    n_draws = S7::new_property(S7::class_integer, default = 500L)
  )
)

# fit(): estimate the model and package draws
S7::method(fit_model, LinearFitter) <- function(fitter, data_bundle, fit_spec, seed, task_ctx) {
  train <- data_bundle$train
  fit <- lm(y ~ x, data = train)
  coefs <- coef(fit)
  vc <- vcov(fit)
  draws <- MASS::mvrnorm(fitter@n_draws, mu = coefs, Sigma = vc)
  colnames(draws) <- c("intercept", "slope")

  # Package into a fit result. new_fit_result() is an internal helper.
  result <- list(
    fit = fit,
    draws = draws,
    diagnostics = list(
      rhat_max = 1.0, ess_bulk_min = as.numeric(fitter@n_draws),
      ess_tail_min = as.numeric(fitter@n_draws), divergent = 0L
    ),
    timing = list(total = 0, warmup = 0, sample = 0),
    success = TRUE,
    warnings = character(0),
    error = NULL,
    data_bundle = data_bundle
  )
  class(result) <- "bayesim_fit_result"
  result
}

# extract_draws(): return the draws matrix
S7::method(extract_draws, LinearFitter) <- function(fitter, fit_result, variables = NULL) {
  fit_result$draws
}

# predict_fit(): posterior-mean predictions
# Convention: predicted_samples is S x N (draws as rows, observations as cols),
# matching log_lik() and predict_epred().
S7::method(predict_fit, LinearFitter) <- function(fitter, fit_result, newdata = NULL, seed = NULL) {
  data <- fit_result$data_bundle$train
  # draws is S x P; design matrix X is N x P, so draws %*% t(X) is S x N.
  X <- as.matrix(cbind(1, data$x))
  preds <- fit_result$draws %*% t(X)
  list(
    predicted_mean = colMeans(preds),
    predicted_samples = preds,
    predicted_sd = apply(preds, 2, sd)
  )
}

# log_lik(): per-observation log-likelihood of each draw
S7::method(log_lik_matrix, LinearFitter) <- function(fitter, fit_result, newdata = NULL) {
  data <- newdata %||% fit_result$data_bundle$train
  fit <- fit_result$fit
  sigma2 <- sum(residuals(fit)^2) / fit$df.residual
  X <- as.matrix(cbind(1, data$x))
  mu <- t(fit_result$draws %*% t(X))
  # matrix: draws (rows) x observations (cols)
  t(dnorm(t(data$y), mean = mu, sd = sqrt(sigma2), log = TRUE))
}

# loo_fit(): not supported by this fitter (supports_loo = FALSE)
S7::method(loo_fit, LinearFitter) <- function(fitter, fit_result) {
  list(elpd = NA_real_, p_loo = NA_real_, elpd_se = NA_real_, pareto_k = NA_real_)
}

# diagnostics()
S7::method(fit_diagnostics, LinearFitter) <- function(fitter, fit_result) {
  fit_result$diagnostics
}
```

## Using the custom fitter

```{r}
my_generator <- function(data_spec, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(n)
  y <- data_spec$slope * x + stats::rnorm(n)
  list(
    train = data.frame(y = y, x = x), test = NULL, response = "y",
    true_params = c(slope = data_spec$slope),
    vars_of_interest = c("slope")
  )
}

config <- simulation_config(
  data_grid = data.frame(n = 50, slope = 2),
  fit_grid = data.frame(model = "linear"),
  data_generator = my_generator,
  fitter = LinearFitter(),
  metrics = list(pred_rmse_metric(), pred_bias_metric(), posterior_summary_metric()),
  n_replicates = 4L,
  seed = 42L
)

result <- run_simulation(config, progress = FALSE)
head(result$summary)
```

## Validation

Use `validate_fitter()` to confirm a fitter implements all required
generics before running a study:

```{r}
validate_fitter(LinearFitter())
```

# Raw Stan: CmdStanFitter

When brms cannot express your model, write the `.stan` file yourself and wire
it in with `CmdStanFitter()` — no custom fitter class needed. You supply the
Stan program, a `stan_data` function mapping a data bundle to the Stan `data`
list, and (optionally) the names of `log_lik` / `epred` generated quantities:

```{r, eval=FALSE}
stan_file <- system.file("stan", "linear_regression.stan", package = "bayesim")

fitter <- CmdStanFitter(
  stan_file = stan_file,
  stan_data = function(data_bundle, fit_spec) {
    X <- stats::model.matrix(~x, data_bundle$train)
    list(
      N = nrow(X), K = ncol(X) - 1L, X = X,
      y = data_bundle$train[[data_bundle$response]]
    )
  },
  log_lik = "log_lik",   # generated-quantities vector -> LOO metrics work
  epred = "mu",          # optional: in-sample expectation matrix
  chains = 2L, iter_warmup = 500L, iter_sampling = 500L
)
```

The shipped `inst/stan/linear_regression.stan` shows the pattern: a
`generated quantities` block computing `log_lik[n]` (pointwise
log-likelihood) and `mu[n]` (the expectation). With `log_lik` declared,
`supports_log_lik`/`supports_loo` are set automatically and `elpd_loo_metric()`
works out of the box; without it, an informative error names the convention.

Limitations, by design: raw Stan has no newdata semantics, so
`supports_predictions` is `FALSE` (test-set prediction metrics need brms or a
custom fitter), and `predict_epred()` returns the in-sample GQ matrix only
when `epred` is declared. Compilation is cached by cmdstanr (keyed on the
file hash), so daemons compile-or-cache-hit on first use — no model-bank
involvement.

# Custom Metrics

Metrics are the other half of the extension surface — see
`vignette("custom-metrics")` for the Metric contract, the output schema,
`summary_type`, and externalization of large outputs.

# A brms-based fitter (sketch)

For a real Stan fitter, `BrmsFitter()` is the standard path and benefits from
the model bank (one compile per distinct spec). A custom brms fitter would:

1. Call `brms::brm()` (or reuse a prefit via the model bank) in `fit_model()`.
2. Use `brms::as_draws_matrix()` in `extract_draws()`.
3. Use `brms::posterior_predict()` / `brms::log_lik()` in the other generics.

See the
[`BrmsFitter` source](https://github.com/sims1253/bayesim/blob/main/R/brms-fitter.R)
for a complete reference implementation. Because the model bank keys
on `fit_spec` columns (`formula`, `family`, `prior`, `stanvars`), supplying
these as `fit_grid` columns lets the engine compile once and reuse the binary
across all replicates of the same condition.
