---
title: "Custom Fitters and Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Fitters and Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Custom Fitters

bayesim ships with `MockFitter()` (no Stan, for testing the framework) and
`BrmsFitter()` (real Bayesian inference via brms + Stan). To use a different
backend, implement a custom fitter by extending the `Fitter` class and
providing methods for the required generics.

## The Fitter contract

Every fitter must implement these S7 generics:

| generic | signature | returns |
|---|---|---|
| `fit(fitter, data_bundle, fit_spec, seed, task_ctx)` | training data + spec | a `bayesim_fit_result` (use `new_fit_result()` internally) |
| `extract_draws(fitter, fit_result)` | a fit result | a draws matrix (rows = draws, cols = parameters) |
| `predict_fit(fitter, fit_result, seed)` | a fit result | list with `predicted_mean` (length N), `predicted_samples` (S x N), `predicted_sd` (length N) |
| `log_lik(fitter, fit_result, newdata)` | a fit result | log-likelihood matrix (draws x observations) |
| `loo_fit(fitter, fit_result)` | a fit result | list with `elpd`, `p_loo`, `elpd_se`, `pareto_k` |
| `diagnostics(fitter, fit_result)` | a fit result | list with `rhat_max`, `ess_bulk_min`, `divergent`, ... |

The generic is named `loo_fit` (not `loo`) to avoid clashing with
`loo::loo()`.

## A minimal fitter: LinearFitter

This fitter fits a linear model by OLS and synthesizes posterior draws by
bootstrapping the residual variance. It is executable without Stan.

```{r}
library(bayesim)

# Define the class
LinearFitter <- S7::new_class(
  "LinearFitter",
  parent = Fitter,
  properties = list(
    name = S7::new_property(S7::class_character, default = "linear"),
    supports_predictions = S7::new_property(S7::class_logical, default = TRUE),
    supports_log_lik = S7::new_property(S7::class_logical, default = TRUE),
    supports_loo = S7::new_property(S7::class_logical, default = FALSE),
    n_draws = S7::new_property(S7::class_integer, default = 500L)
  )
)

# fit(): estimate the model and package draws
S7::method(fit_model, LinearFitter) <- function(fitter, data_bundle, fit_spec, seed, task_ctx) {
  train <- data_bundle$train
  fit <- lm(y ~ x, data = train)
  coefs <- coef(fit)
  vc <- vcov(fit)
  draws <- MASS::mvrnorm(fitter@n_draws, mu = coefs, Sigma = vc)
  colnames(draws) <- c("intercept", "slope")

  # Package into a fit result. new_fit_result() is an internal helper.
  result <- list(
    fit = fit,
    draws = draws,
    diagnostics = list(
      rhat_max = 1.0, ess_bulk_min = as.numeric(fitter@n_draws),
      ess_tail_min = as.numeric(fitter@n_draws), divergent = 0L
    ),
    timing = list(total = 0, warmup = 0, sample = 0),
    success = TRUE,
    warnings = character(0),
    error = NULL,
    data_bundle = data_bundle
  )
  class(result) <- "bayesim_fit_result"
  result
}

# extract_draws(): return the draws matrix
S7::method(extract_draws, LinearFitter) <- function(fitter, fit_result, variables = NULL) {
  fit_result$draws
}

# predict_fit(): posterior-mean predictions
# Convention: predicted_samples is S x N (draws as rows, observations as cols),
# matching log_lik() and predict_epred().
S7::method(predict_fit, LinearFitter) <- function(fitter, fit_result, newdata = NULL, seed = NULL) {
  data <- fit_result$data_bundle$train
  # draws is S x P; design matrix X is N x P, so draws %*% t(X) is S x N.
  X <- as.matrix(cbind(1, data$x))
  preds <- fit_result$draws %*% t(X)
  list(
    predicted_mean = colMeans(preds),
    predicted_samples = preds,
    predicted_sd = apply(preds, 2, sd)
  )
}

# log_lik(): per-observation log-likelihood of each draw
S7::method(log_lik_matrix, LinearFitter) <- function(fitter, fit_result, newdata = NULL) {
  data <- newdata %||% fit_result$data_bundle$train
  fit <- fit_result$fit
  sigma2 <- sum(residuals(fit)^2) / fit$df.residual
  X <- as.matrix(cbind(1, data$x))
  mu <- t(fit_result$draws %*% t(X))
  # matrix: draws (rows) x observations (cols)
  t(dnorm(t(data$y), mean = mu, sd = sqrt(sigma2), log = TRUE))
}

# loo_fit(): not supported by this fitter (supports_loo = FALSE)
S7::method(loo_fit, LinearFitter) <- function(fitter, fit_result) {
  list(elpd = NA_real_, p_loo = NA_real_, elpd_se = NA_real_, pareto_k = NA_real_)
}

# diagnostics()
S7::method(fit_diagnostics, LinearFitter) <- function(fitter, fit_result) {
  fit_result$diagnostics
}
```

## Using the custom fitter

```{r}
my_generator <- function(data_spec, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(n)
  y <- data_spec$slope * x + stats::rnorm(n)
  list(
    train = data.frame(y = y, x = x), test = NULL, response = "y",
    true_params = c(slope = data_spec$slope),
    vars_of_interest = c("slope")
  )
}

config <- simulation_config(
  data_grid = data.frame(n = 50, slope = 2),
  fit_grid = data.frame(model = "linear"),
  data_generator = my_generator,
  fitter = LinearFitter(),
  metrics = list(rmse_metric(), bias_metric(), posterior_summary_metric()),
  n_replicates = 4L,
  seed = 42L
)

result <- run_simulation(config, progress = FALSE)
head(result$summary)
```

## Validation

Use `validate_fitter_interface()` to confirm a fitter implements all required
generics before running a study:

```{r}
validate_fitter_interface(LinearFitter())
```

# Custom Metrics

Extend the `Metric` class and implement `compute()`. The method receives the
fit result, the data bundle, a precomputed `context` (predictions, log_lik,
loo -- only what your metric's `needs` declares), and the task context.

```{r}
MADMetric <- S7::new_class(
  "MADMetric",
  parent = Metric,
  properties = list(
    name = S7::new_property(S7::class_character, default = "mad"),
    needs = S7::new_property(S7::class_character, default = "predictions"),
    required = S7::new_property(S7::class_logical, default = FALSE)
  )
)

S7::method(compute_metric, MADMetric) <- function(metric, fit_result, data_bundle, context, task_ctx) {
  if (is.null(context$predictions)) return(list(value = NA_real_))
  test_data <- data_bundle$test %||% data_bundle$train
  actual <- test_data[[data_bundle$response]]
  predicted <- context$predictions$predicted_mean
  list(value = stats::median(abs(predicted - actual)))
}

mad_metric <- function(name = "mad") MADMetric(name = name, needs = "predictions", required = FALSE)
```

The `needs` property tells the engine which fitter capabilities to precompute
into `context`: `"predictions"`, `"log_lik"`, `"loo"` (or `character()` for
none). Built-in metrics (`rmse_metric`, `bias_metric`, `coverage_metric`,
`posterior_summary_metric`, ...) cover the common surface; see
`?Metric` and the getting-started vignette for the full list.

# A brms-based fitter (sketch)

For a real Stan fitter, `BrmsFitter()` is the standard path and benefits from
the model bank (one compile per distinct spec). A custom brms fitter would:

1. Call `brms::brm()` (or reuse a prefit via the model bank) in `fit()`.
2. Use `brms::as_draws_matrix()` in `extract_draws()`.
3. Use `brms::posterior_predict()` / `brms::log_lik()` in the other generics.

See `BrmsFitter` source (`bayesim:::brms-fitter.R` or the package source on
GitHub) for a complete reference implementation. Because the model bank keys
on `fit_spec` columns (`formula`, `family`, `prior`, `stanvars`), supplying
these as `fit_grid` columns lets the engine compile once and reuse the binary
across all replicates of the same condition.
