---
title: "Custom Metrics"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Metrics}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bayesim)
```

A metric computes one row's worth of values per task — a summary of a single
fit against a single generated dataset. Aggregation *across* tasks (bias,
coverage, MCSEs) happens later, in `summarize_simulation()` and
`performance_measures()`. This vignette covers writing your own metric:
the contract, the output schema, `summary_type`, and how large outputs are
externalized.

## The Metric contract

Extend the `Metric` S7 class and implement `compute_metric()`:

```{r}
MADMetric <- S7::new_class(
  "MADMetric",
  parent = Metric,
  properties = list(
    name = S7::new_property(S7::class_character, default = "mad"),
    needs = S7::new_property(S7::class_character, default = "predictions"),
    required = S7::new_property(S7::class_logical, default = FALSE)
  )
)

S7::method(compute_metric, MADMetric) <- function(
  metric, fit_result, data_bundle, context, task_ctx
) {
  if (is.null(context$predictions) || is.null(data_bundle$test)) {
    return(list(value = NA_real_))
  }
  actual <- data_bundle$test[[data_bundle$response]]
  predicted <- context$predictions$predicted_mean
  list(value = stats::median(abs(predicted - actual)))
}

# Construct the subclass directly; the defaults declared above are honored.
mad_metric <- function(name = "mad") MADMetric(name = name)
```

::: {.callout-note}
### `Metric` is abstract

`Metric` has no direct instances — `Metric()` itself errors. Subclasses are
instantiated directly (`MADMetric(name = "mad")`), and because the parent is
abstract, S7 honors the defaults your subclass declares for the inherited
`name`, `needs`, `required`, `summary_type`, and `schema` properties:
`MADMetric()` alone yields `name = "mad"` and `needs = "predictions"`.
Explicit arguments (as in `mad_metric(name = ...)`) always win over the
declared defaults.
:::

`compute_metric()` receives:

- `fit_result` — the `bayesim_fit_result` (with `$draws`, `$diagnostics`, and
  `$fit` if retained).
- `data_bundle` — the generator's output (`$train`, `$test`, `$response`,
  `$true_params`, `$vars_of_interest`).
- `context` — precomputed values, driven by your metric's `needs` property:
  `"predictions"` (a `predict_fit()` result), `"log_lik"` (an S x N matrix),
  `"loo"` (elpd summary plus PSIS machinery). Declare only what you use; the
  engine computes each at most once per task and shares it across metrics.
  Predictions and log-lik are evaluated on the **test set** when one exists,
  otherwise on the training data.
- `task_ctx` — task identity (`task_id`, `data_idx`, `fit_idx`, `rep_idx`,
  `seed`).

Degrade to `NA` when inputs are missing (as above) rather than erroring:
metrics with `required = FALSE` (the default) record NA on failure, while
`required = TRUE` metrics fail the whole task.

## The output schema

`compute_metric()` must return a named list where every element is either a
scalar atomic value or a **named** numeric vector. No matrices, data frames,
or nested lists — task results must stay flat and cheap to store.

The engine flattens the output into summary columns as
`<metric_name>__<field>`, and named vectors expand per element:

```{r, eval=FALSE}
list(value = 0.3, n_obs = 50L)
#> mad__value, mad__n_obs

list(by_param = c(Intercept = 1, x = 0))
#> mad__by_param__Intercept, mad__by_param__x
```

`validate_metric_output()` enforces this schema. Even better, run
`validate_metric()` with representative values in your tests — it executes
`compute_metric()` once, checks the output schema, and verifies that every
field your `schema` declares is actually produced:

```{r, eval=FALSE}
validate_metric(
  mad_metric(),
  fit_result = new_fit_result(
    draws = cbind(intercept = rnorm(20), slope = rnorm(20))
  ),
  data_bundle = list(
    train = data.frame(y = rnorm(30), x = rnorm(30)),
    test = data.frame(y = rnorm(10), x = rnorm(10)),
    response = "y"
  ),
  context = list(
    predictions = list(
      predicted_mean = rnorm(10),
      predicted_samples = matrix(rnorm(20 * 10), 20, 10),
      predicted_sd = rep(1, 10)
    )
  )
)
```

## summary_type: how your metric aggregates

`summarize_simulation()` needs to know how to aggregate each metric's columns
across replicates. Declare it with the `summary_type` property:

- `"mean"` (default) — report mean/median/sd with an `sd/sqrt(n)` MCSE.
- `"proportion"` — for 0/1 outcomes like coverage; MCSE is `sqrt(p(1-p)/n)`.
- `"none"` — never aggregate (e.g. SBC ranks, which are analyzed as a
  distribution via `sbc_ranks()`, not averaged).

```{r, eval=FALSE}
HitMetric <- S7::new_class(
  "HitMetric",
  parent = Metric,
  properties = list(
    name = S7::new_property(S7::class_character, default = "hit"),
    summary_type = S7::new_property(S7::class_character, default = "proportion")
  )
)
```

## Parallel safety

Under `run_simulation(config, workers = N)`, metric objects are crated and
shipped to daemon processes. Keep `compute_metric()` methods self-contained:
call package functions by namespace (`stats::median`, not a re-exported
alias) and do not reference variables captured from your interactive session.

## Externalization of large outputs

Named numeric vectors longer than 50 elements (or above 64 kB) are not
inlined into the summary — with a `result_path` set, they are written to
`<result_path>/artifacts/metrics/` and the summary records a pointer
(`<metric>__<field>__artifact_path`, `__artifact_hash`, `__n_values`). This
keeps summaries navigable when a metric emits, say, per-observation Pareto-k
values across ten thousand tasks. Without a `result_path`, large vectors are
inlined as usual.

## Using a custom metric

```{r}
gen <- function(data_spec, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(2 * n)
  y <- 2 * x + stats::rnorm(2 * n)
  idx <- seq_len(n)
  list(
    train = data.frame(y = y, x = x)[idx, ],
    test = data.frame(y = y, x = x)[-idx, ],
    response = "y",
    true_params = c(Intercept = 0, x = 2, sigma = 1),
    vars_of_interest = c("Intercept", "x", "sigma")
  )
}

config <- simulation_config(
  data_grid = data.frame(n = 50),
  fit_grid = data.frame(model = "linear"),
  data_generator = gen,
  fitter = LinearRegressionFitter(n_draws = 200L),
  metrics = list(mad_metric(), posterior_summary_metric()),
  n_replicates = 4L,
  seed = 42L
)

result <- run_simulation(config, progress = FALSE)
summarize_simulation(result, metrics = "mad__value")
```

## Next steps

- `vignette("custom-fitters")` — the other half of the extension surface.
- `?Metric` — the canonical contract reference.
- `vignette("design-of-simulation-studies")` — which performance measures to
  compute across tasks, and their MCSEs.
