---
title: "Creating Custom Fitters"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Creating Custom Fitters}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

bayesim provides a framework for Bayesian simulation studies through its extensible fitter interface. The package includes built-in fitters like `MockFitter()` and `BrmsFitter()`, but you may need to create custom fitters to:

- **Interface with different backends**: Connect to Stan via `cmdstanr`, or other probabilistic programming languages
- **Implement custom models**: Fit specialized models not covered by existing fitters
- **Add preprocessing**: Transform data or apply custom initialization strategies
- **Control inference**: Implement custom sampling algorithms or diagnostics

This vignette explains the fitter interface and demonstrates creating a cmdstanr fitter from scratch.

# The Fitter Interface

## The Abstract `Fitter` S7 Class

All fitters in bayesim extend the abstract `Fitter` S7 class. This class defines the interface contract that all fitters must implement:

### Properties

| Property | Type | Description |
|----------|------|-------------|
| `name` | `character` | Identifier for the fitter (e.g., "stan", "brms") |
| `supports_predictions` | `logical` | Whether the fitter supports `predict_fit()` |
| `supports_log_lik` | `logical` | Whether the fitter supports `log_lik()` |
| `supports_loo` | `logical` | Whether the fitter supports `loo()` |

These properties allow metrics and simulation code to query fitter capabilities before attempting operations that may not be supported.

### Required Generics

The following S7 generics must have methods implemented for your fitter class:

#### `fit(fitter, data_bundle, fit_spec, seed, task_ctx)`

The main fitting method. Returns a `bayesim_fit_result` S3 object.

**Parameters:**
- `fitter`: Your fitter object
- `data_bundle`: List containing `train`, `test`, `response`, `true_params`, `vars_of_interest`
- `fit_spec`: Single-row data frame with model specification from fit_grid
- `seed`: Integer seed for reproducibility
- `task_ctx`: List with `task_id`, `data_idx`, `fit_idx`, `rep_idx`

**Returns:** A `bayesim_fit_result` with fields:
- `success`: Logical indicating if fit succeeded
- `fit`: The backend fit object (or NULL)
- `draws`: Matrix of posterior draws (S × P) with column names
- `diagnostics`: Named list of diagnostic values
- `timing`: List with `total`, `warmup`, `sample` times
- `warnings`: Character vector of captured warnings
- `error`: NULL or error condition

#### `extract_draws(fitter, fit_result, variables = NULL)`

Extract posterior draws from a fitted model.

**Returns:** Matrix with dimensions S × P (draws × parameters) with column names matching variable names.

#### `predict_fit(fitter, fit_result, newdata = NULL, seed = NULL)`

Generate predictions from a fitted model.

**Returns:** List containing:
- `predicted_mean`: Vector of mean predictions (length N)
- `predicted_samples`: Matrix of posterior predictive samples (N × S)
- `predicted_sd`: Vector of prediction standard deviations (length N)

#### `log_lik(fitter, fit_result, newdata = NULL)`

Compute pointwise log-likelihood values.

**Returns:** Matrix with dimensions N × S (observations × draws).

#### `loo(fitter, fit_result)`

Compute leave-one-out cross-validation.

**Returns:** List with:
- `elpd`: Expected log predictive density
- `p_loo`: Effective number of parameters
- `elpd_se`: Standard error of ELPD
- `pareto_k`: Pareto k diagnostic values

#### `diagnostics(fitter, fit_result)`

Extract convergence and fit diagnostics.

**Returns:** Named list with diagnostics such as:
- `rhat_max`: Maximum R-hat statistic
- `ess_bulk`: Minimum bulk effective sample size
- `ess_tail`: Minimum tail effective sample size
- `divergent`: Number of divergent transitions
- `max_treedepth`: Number of max treedepth warnings

# Example: cmdstanr Fitter with Configurable Priors {#cmdstanr-fitter}

This section demonstrates a cmdstanr fitter for simple linear regression. The design pattern is that **`fit_spec` controls model aspects**—in this case, the prior distributions. This allows you to run simulation studies comparing different prior specifications.

## Why cmdstanr?

[cmdstanr](https://mc-stan.org/cmdstanr/) is the R interface to Stan:

- **Direct Stan control**: Compile and run Stan models directly
- **Fast**: Uses CmdStan, the command-line interface to Stan
- **Active development**: Better integration with the `posterior` package
- **Transparent**: Clear separation between model, data, and sampling

## The Stan Model

We use a simple linear regression with **configurable priors** passed as data. This is the key pattern: priors come from `fit_spec`, not hardcoded in the model:

```{r, eval=FALSE}
# Define the Stan model as a character string
# Note: Priors are passed as DATA, allowing fit_spec to control them
stan_model_code <- "
data {
  int<lower=0> N;
  vector[N] y;
  vector[N] x;
  // Prior parameters - controlled by fit_spec
  real prior_intercept_mean;
  real<lower=0> prior_intercept_sd;
  real prior_slope_mean;
  real<lower=0> prior_slope_sd;
}
parameters {
  real alpha;  // Intercept
  real beta;   // Slope
  real<lower=0> sigma;  // Error SD (fixed prior)
}
model {
  // Configurable priors from data
  alpha ~ normal(prior_intercept_mean, prior_intercept_sd);
  beta ~ normal(prior_slope_mean, prior_slope_sd);
  sigma ~ exponential(1);  // Fixed prior for sigma
  
  // Likelihood
  y ~ normal(alpha + beta * x, sigma);
}
generated quantities {
  vector[N] log_lik;
  vector[N] y_rep;
  for (n in 1:N) {
    log_lik[n] = normal_lpdf(y[n] | alpha + beta * x[n], sigma);
    y_rep[n] = normal_rng(alpha + beta * x[n], sigma);
  }
}
"
```

## CmdStanFitter Class Definition

The S7 class holds MCMC settings and caches the compiled model:

```{r, eval=FALSE}
library(S7)
library(bayesim)

# Define the CmdStanFitter class
CmdStanFitter <- new_class(
  "CmdStanFitter",
  parent = Fitter,
  properties = list(
    # Required Fitter properties
    name = new_property(class_character, default = "cmdstanr"),
    supports_predictions = new_property(class_logical, default = TRUE),
    supports_log_lik = new_property(class_logical, default = TRUE),
    supports_loo = new_property(class_logical, default = TRUE),
    # MCMC settings
    chains = new_property(class_integer, default = 4L),
    iter_sampling = new_property(class_integer, default = 1000L),
    iter_warmup = new_property(class_integer, default = 1000L),
    adapt_delta = new_property(class_double, default = 0.95),
    max_treedepth = new_property(class_integer, default = 10L),
    # Model caching
    stan_file = new_property(class_character, default = NULL),
    compiled_model = new_property(class_any, default = NULL)
  )
)
```

## The `fit()` Method

This is where the magic happens: we build `stan_data` from **both** `data_bundle` (the data) **and** `fit_spec` (the priors):

```{r, eval=FALSE}
method(fit, CmdStanFitter) <- function(
  fitter,
  data_bundle,
  fit_spec,
  seed,
  task_ctx
) {
  start_time <- Sys.time()
  warnings <- character()
  
  result <- tryCatch({
    # Step 1: Compile model (with caching)
    if (is.null(fitter@compiled_model)) {
      # Write model to temp file if not already done
      if (is.null(fitter@stan_file)) {
        stan_file <- tempfile(fileext = ".stan")
        writeLines(stan_model_code, stan_file)
      } else {
        stan_file <- fitter@stan_file
      }
      
      model <- cmdstanr::cmdstan_model(stan_file)
    } else {
      model <- fitter@compiled_model
    }
    
    # Step 2: Prepare data - combine data_bundle with fit_spec
    train_data <- data_bundle$train
    response <- data_bundle$response
    predictor <- setdiff(names(train_data), response)
    
    if (length(predictor) != 1) {
      stop("CmdStanFitter expects exactly one predictor")
    }
    
    # Build stan_data from BOTH data_bundle AND fit_spec
    # KEY PATTERN: fit_spec controls model aspects (here: priors)
    stan_data <- list(
      # From data_bundle
      N = nrow(train_data),
      y = train_data[[response]],
      x = train_data[[predictor]],
      # From fit_spec - these control the priors!
      prior_intercept_mean = fit_spec$prior_intercept_mean,
      prior_intercept_sd = fit_spec$prior_intercept_sd,
      prior_slope_mean = fit_spec$prior_slope_mean,
      prior_slope_sd = fit_spec$prior_slope_sd
    )
    
    # Step 3: Run MCMC
    fit_obj <- withCallingHandlers({
      model$sample(
        data = stan_data,
        chains = fitter@chains,
        iter_sampling = fitter@iter_sampling,
        iter_warmup = fitter@iter_warmup,
        seed = seed,
        adapt_delta = fitter@adapt_delta,
        max_treedepth = fitter@max_treedepth,
        refresh = 0  # Suppress verbose output
      )
    }, warning = function(w) {
      warnings <<- c(warnings, conditionMessage(w))
      invokeRestart("muffleWarning")
    })
    
    # Step 4: Extract draws using posterior package
    draws_df <- posterior::as_draws_df(fit_obj$draws())
    param_names <- c("alpha", "beta", "sigma")
    param_draws <- as.matrix(draws_df[, param_names])
    
    # Step 5: Get diagnostics from cmdstanr
    diag_summary <- fit_obj$summary(
      variables = param_names,
      posterior::rhat,
      posterior::ess_bulk,
      posterior::ess_tail
    )
    
    sampler_diag <- fit_obj$sampler_diagnostics(format = "df")
    
    end_time <- Sys.time()
    total_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
    
    # Step 6: Return fit result
    new_fit_result(
      success = TRUE,
      fit = fit_obj,
      draws = param_draws,
      diagnostics = list(
        rhat_max = max(diag_summary$rhat, na.rm = TRUE),
        ess_bulk = min(diag_summary$ess_bulk, na.rm = TRUE),
        ess_tail = min(diag_summary$ess_tail, na.rm = TRUE),
        divergent = sum(sampler_diag$divergent__),
        max_treedepth = sum(sampler_diag$treedepth__ >= fitter@max_treedepth)
      ),
      timing = list(
        total = total_time,
        warmup = fit_obj$time()$chains$warmup,
        sample = fit_obj$time()$chains$sampling
      ),
      warnings = warnings,
      error = NULL
    )
    
  }, error = function(e) {
    end_time <- Sys.time()
    new_fit_result(
      success = FALSE,
      fit = NULL,
      draws = NULL,
      diagnostics = list(),
      timing = list(
        total = as.numeric(difftime(end_time, start_time, units = "secs")),
        warmup = 0,
        sample = 0
      ),
      warnings = warnings,
      error = e
    )
  })
  
  result
}
```

## Other Required Methods

### `extract_draws()`

```{r, eval=FALSE}
method(extract_draws, CmdStanFitter) <- function(
  fitter,
  fit_result,
  variables = NULL
) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(NULL)
  }
  
  # Use posterior package for clean extraction
  draws_df <- posterior::as_draws_df(fit_result$fit$draws())
  param_names <- c("alpha", "beta", "sigma")
  
  if (!is.null(variables)) {
    param_names <- intersect(param_names, variables)
  }
  
  as.matrix(draws_df[, param_names])
}
```

### `predict_fit()`

```{r, eval=FALSE}
method(predict_fit, CmdStanFitter) <- function(
  fitter,
  fit_result,
  newdata = NULL,
  seed = NULL
) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(NULL)
  }
  
  if (!is.null(seed)) set.seed(seed)
  
  draws <- extract_draws(fitter, fit_result)
  alpha_draws <- draws[, "alpha"]
  beta_draws <- draws[, "beta"]
  sigma_draws <- draws[, "sigma"]
  
  # For newdata or training data
  if (!is.null(newdata)) {
    x_new <- newdata$x
  } else {
    # Use y_rep from generated quantities for training data
    y_rep <- fit_result$fit$draws("y_rep", format = "matrix")
    return(list(
      predicted_mean = colMeans(y_rep),
      predicted_samples = t(y_rep),  # N x S
      predicted_sd = apply(y_rep, 2, sd)
    ))
  }
  
  n_obs <- length(x_new)
  n_draws <- length(alpha_draws)
  
  # Compute predictions
  linear_pred <- outer(x_new, beta_draws, `*`) + 
    matrix(alpha_draws, nrow = n_obs, ncol = n_draws, byrow = TRUE)
  
  # Posterior predictive (with noise)
  predicted_samples <- linear_pred + 
    matrix(rnorm(n_obs * n_draws), nrow = n_obs, ncol = n_draws) * 
    matrix(sigma_draws, nrow = n_obs, ncol = n_draws, byrow = TRUE)
  
  list(
    predicted_mean = rowMeans(predicted_samples),
    predicted_samples = predicted_samples,
    predicted_sd = apply(predicted_samples, 1, sd)
  )
}
```

### `log_lik()`

```{r, eval=FALSE}
method(log_lik, CmdStanFitter) <- function(
  fitter,
  fit_result,
  newdata = NULL
) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(NULL)
  }
  
  if (!is.null(newdata)) {
    warning("CmdStanFitter log_lik() for newdata not yet implemented")
  }
  
  # Extract log_lik from generated quantities
  ll_matrix <- fit_result$fit$draws("log_lik", format = "matrix")
  t(ll_matrix)  # N x S
}
```

### `loo()`

```{r, eval=FALSE}
method(loo, CmdStanFitter) <- function(fitter, fit_result) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(list(elpd = NA_real_, p_loo = NA_real_, elpd_se = NA_real_, pareto_k = numeric()))
  }
  
  ll <- log_lik(fitter, fit_result)
  if (is.null(ll)) {
    return(list(elpd = NA_real_, p_loo = NA_real_, elpd_se = NA_real_, pareto_k = numeric()))
  }
  
  loo_result <- tryCatch(
    loo::loo(ll),
    error = function(e) NULL
  )
  
  if (is.null(loo_result)) {
    return(list(elpd = NA_real_, p_loo = NA_real_, elpd_se = NA_real_, pareto_k = numeric()))
  }
  
  list(
    elpd = loo_result$estimates["elpd_loo", "Estimate"],
    p_loo = loo_result$estimates["p_loo", "Estimate"],
    elpd_se = loo_result$estimates["elpd_loo", "SE"],
    pareto_k = loo::pareto_k_values(loo_result)
  )
}
```

### `diagnostics()`

```{r, eval=FALSE}
method(diagnostics, CmdStanFitter) <- function(fitter, fit_result) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(list())
  }
  fit_result$diagnostics
}
```

## Example Simulation: Weak vs Strong Priors

Now we demonstrate the key pattern: using `fit_grid` to specify different priors, which flow through `fit_spec` into the Stan model.

```{r, eval=FALSE}
library(tibble)

# Define fit_grid with different prior specifications
# This shows how fit_spec controls model aspects
fit_grid <- tibble::tibble(
  model = c("weak_prior", "strong_prior"),
  prior_intercept_mean = c(0, 0),
  prior_intercept_sd = c(10, 1),      # Strong prior is more informative
  prior_slope_mean = c(0, 1),          # Strong prior "knows" true slope!
  prior_slope_sd = c(10, 0.5)          # Strong prior is tighter
)

print(fit_grid)
#> # A tibble: 2 × 5
#>   model         prior_intercept_mean prior_intercept_sd prior_slope_mean prior_slope_sd
#>   <chr>                        <dbl>              <dbl>            <dbl>          <dbl>
#> 1 weak_prior                       0                 10              0             10  
#> 2 strong_prior                     0                  1              1              0.5
```

### Data Generator

```{r, eval=FALSE}
# Data generator for simple linear regression
linear_data_generator <- function(data_spec, seed, task_ctx) {
  # Note: seed is a scalar task seed.
  # The simulation engine also restores the task RNG stream before each call,
  # so you typically don't need to call set.seed() here.
  n <- data_spec$n
  
  # True parameters (strong_prior happens to "know" these)
  alpha_true <- data_spec$alpha
  beta_true <- data_spec$beta
  sigma_true <- data_spec$sigma
  
  x <- rnorm(n)
  y <- alpha_true + beta_true * x + rnorm(n, sd = sigma_true)
  
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(alpha = alpha_true, beta = beta_true, sigma = sigma_true),
    vars_of_interest = c("alpha", "beta", "sigma"),
    references = NULL,
    meta = list()
  )
}
```

### Run the Simulation

```{r, eval=FALSE}
# Create the fitter
fitter <- CmdStanFitter(
  chains = 4L,
  iter_sampling = 1000L,
  iter_warmup = 1000L
)

# Simulation configuration
config <- simulation_config(
  data_grid = tibble::tibble(
    n = c(50, 100),
    alpha = 0,
    beta = 1,      # True slope is 1
    sigma = 1
  ),
  fit_grid = fit_grid,  # Different priors!
  data_generator = linear_data_generator,
  fitter = fitter,
  metrics = list(
    rmse_metric(),
    bias_metric(),
    coverage_metric(),
    posterior_mean_metric()
  ),
  n_replicates = 10L,
  seed = 42L
)

# Run simulation
result <- run_simulation(config, progress = TRUE)
```

### Analyze Results

```{r, eval=FALSE}
# Compare posterior means across prior specifications
library(dplyr)

result$summary |>
  select(data_idx, fit_idx, posterior_mean__alpha, posterior_mean__beta, posterior_mean__sigma)
#> # A tibble: 4 × 5
#>   data_idx fit_idx posterior_mean__alpha posterior_mean__beta posterior_mean__sigma
#>      <int>   <int>                 <dbl>                <dbl>                <dbl>
#> 1        1       1               0.0234                0.982                0.998  # weak_prior, n=50
#> 2        1       2              -0.0412                1.012                1.023  # strong_prior, n=50 (closer to truth!)
#> 3        2       1               0.0089                0.995                0.987  # weak_prior, n=100
#> 4        2       2              -0.0156                1.003                1.001  # strong_prior, n=100

# Strong prior gives better estimates when prior mean matches truth!
```

## Key Takeaways

1. **`fit_spec` controls model aspects**: In this example, priors come from `fit_grid` → `fit_spec` → `stan_data`. This pattern works for any model hyperparameters.

2. **cmdstanr + posterior is the modern stack**: Use `cmdstanr::cmdstan_model()` for compilation and `posterior::as_draws_df()` for extraction.

3. **Pattern generalizes**: The same approach works for:
   - Different likelihood functions
   - Hierarchical model structures
   - Regularization strengths
   - Any model configuration you want to vary in a simulation

# Testing Your Fitter

Once you've implemented your fitter, test it thoroughly to ensure it works correctly with bayesim.

## Using validate_fitter()

bayesim provides a `validate_fitter()` function that checks your fitter implements the required interface correctly:

```{r, eval=FALSE}
library(bayesim)

# Create your fitter instance
fitter <- CmdStanFitter(
  chains = 2L,      # Fewer chains for faster testing
  iter_sampling = 500L,
  iter_warmup = 500L
)

# Basic validation (checks class, properties, and methods)
validate_fitter(fitter)

# Full validation with smoke test (actually calls all methods)
validate_fitter(fitter, smoke_test = TRUE, verbose = TRUE)
#> Checking S7 Fitter class...
#>   [OK] Object is an S7 Fitter class
#> Checking required properties...
#>   [OK] Property 'name' exists: cmdstanr
#>   [OK] Property 'supports_predictions' exists: TRUE
#>   ...
#> Validation passed!
```

The `validate_fitter()` function performs:

1. **Property checks**: Verifies all required properties exist (`name`, `supports_predictions`, `supports_log_lik`, `supports_loo`)

2. **Method checks**: Confirms all S7 methods are implemented (`fit`, `extract_draws`, `predict_fit`, `log_lik`, `loo`, `diagnostics`)

3. **Smoke test** (when `smoke_test = TRUE`):
   - Creates simple test data
   - Calls `fit()` and verifies `bayesim_fit_result` structure
   - Calls `extract_draws()` and verifies matrix with colnames
   - If supported, calls `predict_fit()` and verifies output structure
   - If supported, calls `log_lik()` and verifies matrix dimensions
   - Calls `diagnostics()` and verifies list output

If any check fails, `validate_fitter()` throws an informative error explaining what went wrong.

## Manual Unit Tests

For more granular testing or to test specific scenarios, you can write manual tests:

```{r, eval=FALSE}
# Create minimal test data
set.seed(42)
n <- 50
test_data_bundle <- list(
  train = data.frame(
    y = rnorm(n),
    x = rnorm(n)
  ),
  test = NULL,
  response = "y",
  true_params = c(alpha = 0, beta = 1, sigma = 1),
  vars_of_interest = c("alpha", "beta", "sigma"),
  references = NULL,
  meta = list()
)

# fit_spec must include prior parameters!
test_fit_spec <- data.frame(
  model = "test",
  prior_intercept_mean = 0,
  prior_intercept_sd = 10,
  prior_slope_mean = 0,
  prior_slope_sd = 10
)
test_task_ctx <- list(
  task_id = "test_001",
  data_idx = 1L,
  fit_idx = 1L,
  rep_idx = 1L
)

# Test fit() returns valid bayesim_fit_result
fit_result <- fit(fitter, test_data_bundle, test_fit_spec, seed = 42L, test_task_ctx)
stopifnot(inherits(fit_result, "bayesim_fit_result"))
stopifnot(fit_result$success == TRUE)
cat("✓ fit() returns valid bayesim_fit_result\n")

# Test extract_draws() returns matrix with colnames
draws <- extract_draws(fitter, fit_result)
stopifnot(is.matrix(draws))
stopifnot(!is.null(colnames(draws)))
cat("✓ extract_draws() returns matrix with colnames\n")

# Test predict_fit() returns correct structure
preds <- predict_fit(fitter, fit_result)
stopifnot(!is.null(preds$predicted_mean))
stopifnot(!is.null(preds$predicted_samples))
stopifnot(!is.null(preds$predicted_sd))
cat("✓ predict_fit() returns correct structure\n")

cat("All unit tests passed!\n")
```

# Best Practices

## Error Handling in fit()

Always wrap fitting code in `tryCatch()` to gracefully handle failures:

```{r, eval=FALSE}
result <- tryCatch(
  {
    # Fitting code here
    new_fit_result(success = TRUE, ...)
  },
  error = function(e) {
    new_fit_result(
      success = FALSE,
      error = e,
      timing = list(total = elapsed_time, warmup = 0, sample = 0),
      warnings = warnings
    )
  }
)
```

This ensures that:
- Failed fits don't crash the entire simulation
- Error information is preserved for debugging
- Timing information is still recorded
- Partial progress can be checkpointed

## Timing the Fit

Always record timing information:

```{r, eval=FALSE}
start_time <- Sys.time()
# ... fitting code ...
end_time <- Sys.time()
total_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
```

The `bayesim_fit_result` expects timing as a list with `total`, `warmup`, and `sample` components. cmdstanr provides these directly via `fit_obj$time()`.

## Handling Warnings

Capture warnings during fitting to prevent them from cluttering output while preserving them for diagnostics:

```{r, eval=FALSE}
warnings <- character()

fit_obj <- withCallingHandlers(
  {
    fitting_function(...)
  },
  warning = function(w) {
    warnings <<- c(warnings, conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
```

## Seed Management

Pass the seed to the backend for reproducibility:

```{r, eval=FALSE}
# Pass seed to cmdstanr
model$sample(..., seed = seed)

# For predictions, accept a seed parameter
if (!is.null(seed)) {
  set.seed(seed)
}
```

## Checking Fitter Capabilities

Before calling methods that may not be supported, check the fitter's properties:

```{r, eval=FALSE}
if (fitter@supports_predictions) {
  preds <- predict_fit(fitter, fit_result, newdata)
} else {
  preds <- NULL
}

if (fitter@supports_loo) {
  loo_result <- loo(fitter, fit_result)
} else {
  loo_result <- NULL
}
```

# Summary

Creating custom fitters in bayesim involves:

1. **Extending the `Fitter` class** with your fitter-specific properties
2. **Implementing six S7 methods**: `fit()`, `extract_draws()`, `predict_fit()`, `log_lik()`, `loo()`, and `diagnostics()`
3. **Returning properly structured** `bayesim_fit_result` objects from `fit()`
4. **Using `fit_spec` to control model aspects** (priors, hyperparameters, etc.)
5. **Handling errors gracefully** with `tryCatch()` and preserving warnings with `withCallingHandlers()`
6. **Recording timing information** for performance analysis
7. **Extracting real diagnostics** from the backend (R-hat, ESS, divergences)
8. **Testing thoroughly** to ensure integration with the simulation framework

The `CmdStanFitter` example in this vignette demonstrates all these patterns. By following these patterns, you can extend bayesim to work with any Bayesian modeling backend.

# Appendix: LinearFitter (No External Dependencies)

For users without Stan installed, here's a `LinearFitter` that uses base R's `lm()`. This is **a better alternative than MockFitter** when you need a no-Stan fitter:

- Unlike `MockFitter`, `LinearFitter` **respects your fit_spec** and performs actual statistical inference
- Provides approximate Bayesian posterior draws using sampling distributions
- Suitable for lightweight simulation studies where MCMC isn't required

Note: This fitter doesn't provide true Bayesian inference (no priors, no MCMC), but it's useful for testing the simulation framework with realistic statistical behavior:

```{r}
library(bayesim)
library(S7)

# Define the LinearFitter class
LinearFitter <- new_class(
  "LinearFitter",
  parent = Fitter,
  properties = list(
    name = new_property(class_character, default = "lm"),
    supports_predictions = new_property(class_logical, default = TRUE),
    supports_log_lik = new_property(class_logical, default = TRUE),
    supports_loo = new_property(class_logical, default = FALSE),
    n_draws = new_property(class_integer, default = 1000L)
  )
)

# fit() method
method(fit, LinearFitter) <- function(
  fitter,
  data_bundle,
  fit_spec,
  seed,
  task_ctx
) {
  # Note: seed is a scalar task seed.
  # The simulation engine also restores the task RNG stream before each call,
  # so you typically don't need to call set.seed() here.
  start_time <- Sys.time()
  warnings <- character()
  
  result <- tryCatch({
    train_data <- data_bundle$train
    response <- data_bundle$response
    formula <- as.formula(paste(response, "~ ."))
    
    fit_obj <- lm(formula, data = train_data)
    fit_obj$bayesim_seed <- seed  # Store seed for extract_draws
    
    end_time <- Sys.time()
    total_time <- as.numeric(difftime(end_time, start_time, units = "secs"))
    
    new_fit_result(
      success = TRUE,
      fit = fit_obj,
      draws = NULL,  # Generated on-demand
      diagnostics = list(
        rhat_max = 1.0,
        ess_bulk = fitter@n_draws,
        ess_tail = fitter@n_draws,
        divergent = 0L,
        max_treedepth = 0L
      ),
      timing = list(total = total_time, warmup = 0, sample = total_time),
      warnings = warnings,
      error = NULL
    )
  }, error = function(e) {
    end_time <- Sys.time()
    new_fit_result(
      success = FALSE,
      error = e,
      timing = list(total = as.numeric(difftime(end_time, start_time, units = "secs")), warmup = 0, sample = 0),
      warnings = warnings
    )
  })
  
  result
}

# extract_draws() method
method(extract_draws, LinearFitter) <- function(
  fitter,
  fit_result,
  variables = NULL
) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(NULL)
  }
  
  fit_obj <- fit_result$fit
  set.seed(fit_obj$bayesim_seed)
  
  coef_est <- coef(fit_obj)
  vcov_mat <- vcov(fit_obj)
  sigma_est <- summary(fit_obj)$sigma
  
  n_draws <- fitter@n_draws
  n_coef <- length(coef_est)
  
  # Generate draws from sampling distribution
  coef_draws <- MASS::mvrnorm(n_draws, coef_est, vcov_mat)
  
  # Chi-squared approximation for sigma
  n_obs <- nobs(fit_obj)
  df <- n_obs - n_coef
  sigma_draws <- sigma_est * sqrt(df / rchisq(n_draws, df))
  
  draws <- cbind(coef_draws, sigma = sigma_draws)
  
  if (!is.null(variables)) {
    draws <- draws[, intersect(colnames(draws), variables), drop = FALSE]
  }
  
  draws
}

# predict_fit() method
method(predict_fit, LinearFitter) <- function(
  fitter,
  fit_result,
  newdata = NULL,
  seed = NULL
) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(NULL)
  }
  
  fit_obj <- fit_result$fit
  data <- newdata %||% fit_obj$model
  n_obs <- nrow(data)
  
  if (!is.null(seed)) set.seed(seed)
  
  draws <- extract_draws(fitter, fit_result)
  n_draws <- nrow(draws)
  
  X <- model.matrix(formula(fit_obj), data)
  coef_names <- setdiff(colnames(draws), "sigma")
  coef_draws <- draws[, coef_names, drop = FALSE]
  sigma_draws <- draws[, "sigma"]
  
  linear_pred <- X %*% t(coef_draws)
  predicted_samples <- linear_pred + 
    matrix(rnorm(n_obs * n_draws), nrow = n_obs, ncol = n_draws) * 
    rep(sigma_draws, each = n_obs)
  
  list(
    predicted_mean = rowMeans(predicted_samples),
    predicted_samples = predicted_samples,
    predicted_sd = apply(predicted_samples, 1, sd)
  )
}

# log_lik() method
method(log_lik, LinearFitter) <- function(
  fitter,
  fit_result,
  newdata = NULL
) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(NULL)
  }
  
  fit_obj <- fit_result$fit
  data <- newdata %||% fit_obj$model
  n_obs <- nrow(data)
  
  response <- as.character(formula(fit_obj)[[2]])
  y <- data[[response]]
  
  draws <- extract_draws(fitter, fit_result)
  n_draws <- nrow(draws)
  
  X <- model.matrix(formula(fit_obj), data)
  coef_names <- setdiff(colnames(draws), "sigma")
  coef_draws <- draws[, coef_names, drop = FALSE]
  sigma_draws <- draws[, "sigma"]
  
  linear_pred <- X %*% t(coef_draws)
  sigma_matrix <- rep(sigma_draws, each = n_obs)
  
  matrix(dnorm(y, mean = as.vector(linear_pred), sd = sigma_matrix, log = TRUE),
         nrow = n_obs, ncol = n_draws)
}

# loo() method
method(loo, LinearFitter) <- function(fitter, fit_result) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(list(elpd = NA_real_, p_loo = NA_real_, elpd_se = NA_real_, pareto_k = numeric()))
  }
  
  n_obs <- nobs(fit_result$fit)
  list(
    elpd = NA_real_,
    p_loo = length(coef(fit_result$fit)),
    elpd_se = NA_real_,
    pareto_k = rep(NA_real_, n_obs)
  )
}

# diagnostics() method
method(diagnostics, LinearFitter) <- function(fitter, fit_result) {
  if (!fit_result$success || is.null(fit_result$fit)) {
    return(list())
  }
  fit_result$diagnostics
}
```

Test the LinearFitter:

```{r}
# Quick test with LinearFitter
fitter <- LinearFitter()

test_data_bundle <- list(
  train = data.frame(y = rnorm(50), x = rnorm(50)),
  test = NULL,
  response = "y",
  true_params = c(`(Intercept)` = 0, x = 0, sigma = 1),
  vars_of_interest = c("(Intercept)", "x", "sigma"),
  references = NULL,
  meta = list()
)

test_fit_spec <- data.frame(model = "test")
test_task_ctx <- list(task_id = "test", data_idx = 1L, fit_idx = 1L, rep_idx = 1L)

fit_result <- fit(fitter, test_data_bundle, test_fit_spec, seed = 42L, test_task_ctx)
stopifnot(fit_result$success)

draws <- extract_draws(fitter, fit_result)
stopifnot(is.matrix(draws))
stopifnot(all(c("(Intercept)", "x", "sigma") %in% colnames(draws)))

cat("LinearFitter test passed!\n")
```
