---
title: "Design of simulation studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Design of simulation studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{bayesim}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(bayesim)
```

# Introduction

A simulation study is an experiment whose subjects are *datasets* simulated
from a known data-generating process (DGP), and whose measurements are the
*performance* of one or more estimators applied to those datasets. Because the
truth is known, simulation is the gold standard for evaluating statistical
methods — but only when the study itself is well-designed and honestly
reported.

This vignette follows the framework of Morris, White & Crowther (2019,
*Statistics in Medicine*), which structures a simulation study around four
questions:

1. What are the **estimands** — the quantities we want the method to recover?
2. What are the **data-generating mechanisms** — the DGPs that produce the
   simulated datasets, and how do they expose the estimands?
3. What are the **performance measures** — bias, variance/SE, MSE, coverage,
   etc. — used to judge the method, and what is their Monte Carlo standard
   error (MCSE)?
4. **How many replicates** are needed to estimate those performance measures
   to acceptable precision, and across which **varying conditions**?

bayesim provides typed primitives for each of these. This vignette runs
end-to-end with `LinearRegressionFitter()` (exact conjugate Bayesian linear
regression, no Stan) so every chunk is executable.

# Estimands and data-generating mechanisms

The **estimand** is the target of inference — the parameter whose value the
method is supposed to recover. In a simulation, the estimand is known because
*we* set it. bayesim records it automatically: any name in a generator's
`true_params` vector becomes a `truth__<param>` column in the per-task summary
(see `vignette("getting-started")`), and that column is what the analysis
layer pairs against the per-task posterior estimate.

The **data-generating mechanism** is the function that turns a parameter
setting into a dataset. bayesim generators have the signature
`(data_spec, task_ctx) -> data_bundle` and must consume the ambient RNG state
(the engine restores a per-task L'Ecuyer stream before each call; do not
re-seed inside). The factory constructors `fixed_truth_generator()`,
`prior_predictive_generator()`, and `ifs_generator()` cover common patterns;
for everything else, write an inline generator like the one below.

```{r}
# A linear-regression DGP. The estimands (Intercept, slope x, sigma) are
# recorded as true_params; bayesim surfaces them as truth__* columns.
linear_dgp <- function(data_spec, task_ctx) {
  n <- data_spec$n
  beta <- data_spec$beta
  x <- stats::rnorm(n)
  y <- 1 + beta * x + stats::rnorm(n)
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(Intercept = 1, x = beta, sigma = 1),
    vars_of_interest = c("Intercept", "x", "sigma")
  )
}
```

The `data_grid` is the table of `data_spec` rows swept over. Each row is one
data-generating condition; each condition is replicated `n_replicates` times.

# Performance measures

The performance measures of Morris et al. (2019) characterise an estimator
across replicate datasets:

- **bias** — does the estimator hit the truth on average?
- **empirical SE (emp_se)** — how variable is the estimator across datasets?
- **MSE** — mean squared error, the combined bias-plus-variance.
- **coverage** — does the nominal interval contain the truth at the nominal
  rate?
- **model SE (model_se)** — the average uncertainty the *method itself*
  reports (compare to the empirical SE to check the method's uncertainty
  calibration).

Each of these is a Monte Carlo estimate from `n` replicates, so each carries
an **MCSE** — the simulation's own standard error. Reporting a performance
estimate without its MCSE is like reporting a sample mean without a standard
error: the reader cannot tell signal from noise. bayesim computes every MCSE
for you.

Run a small study and call `performance_measures()`:

```{r}
config <- simulation_config(
  data_grid = data.frame(n = 100L, beta = 0.5),
  fit_grid = data.frame(model = "lm"),
  data_generator = linear_dgp,
  fitter = LinearRegressionFitter(n_draws = 400L),
  metrics = list(posterior_summary_metric()),
  n_replicates = 80L,
  seed = 11L
)

result <- run_simulation(config, progress = FALSE)

pm <- performance_measures(result, estimand = "x")
pm
```

The tidy tibble has one row per `estimand x measure`, with the `value` and its
`mcse`. For the conjugate fitter on this DGP we expect: bias near zero, emp_se
and model_se close to each other (the posterior correctly reports its own
uncertainty), and coverage near the nominal 0.95.

The MCSE formulas follow Morris et al. / rsimsum: bias MCSE = `sd / sqrt(n)`;
emp_se MCSE = `sd / sqrt(2(n-1))`; MSE MCSE via the delta method on the
squared errors; coverage MCSE = `sqrt(p(1-p) / n)`; model_se MCSE =
`sd(posterior_sd) / sqrt(n)`. These are exactly the formulas inverted by
`n_replicates_for_target()` below.

# How many replicates?

The MCSE shrinks as `1/sqrt(n)`, so the number of replicates is a precision
budget. `n_replicates_for_target()` inverts the MCSE formulas to tell you how
many replicates you need to hit a target MCSE *before* you run the study — a
power calculation for simulation.

For coverage, the MCSE is `sqrt(p(1-p) / n)`, maximised at `p = 0.5` (the
conservative default). Inverting gives `n = p(1-p) / MCSE^2`.

```{r}
# Target: coverage MCSE of 0.03 at the conservative p = 0.5.
n_replicates_for_target(target_mcse = 0.03, metric_type = "coverage")
```

So **278** replicates (ceiling) are needed to estimate a coverage rate to
within `+/- 0.03` (one MCSE) at the worst-case variance. If the coverage is
expected near the nominal 0.95, supply `p = 0.95` for a tighter, smaller
estimate:

```{r}
n_replicates_for_target(target_mcse = 0.03, metric_type = "coverage", p = 0.95)
```

For a continuous measure (bias, model SE), the MCSE is `sd / sqrt(n)`, so you
must supply an assumed standard deviation for the per-replicate point estimate
(e.g. a guess at the empirical SE of the estimator, perhaps from a pilot run):

```{r}
# Target bias MCSE of 0.05, assuming the point estimate has sd ~ 0.5.
n_replicates_for_target(
  target_mcse = 0.05, metric_type = "continuous", assumed_sd = 0.5
)
```

Use these numbers to size the study up front: a coverage claim with 50
replicates and MCSE 0.07 is barely more precise than a coin flip, while 1,000+
replicates give MCSEs near 0.015 — the difference between a defensible
simulation and an anecdotal one.

# Varying conditions

A simulation study that varies nothing proves nothing about the *range* of
behaviour a method exhibits. The Morris et al. framework emphasises sweeping
over the factors that plausibly change performance: sample size, effect size,
noise level, model mis-specification, etc. In bayesim, varying conditions are
just additional rows in the `data_grid` (or columns swept within it); the
analysis layer's `by` argument then groups the performance measures by
condition.

Below we vary sample size and effect size together:

```{r}
design_config <- simulation_config(
  data_grid = data.frame(
    n = c(60, 60, 120, 120),
    beta = c(0.5, 1.0, 0.5, 1.0)
  ),
  fit_grid = data.frame(model = "lm"),
  data_generator = linear_dgp,
  fitter = LinearRegressionFitter(n_draws = 400L),
  metrics = list(posterior_summary_metric()),
  n_replicates = 60L,
  seed = 23L
)

design_result <- run_simulation(design_config, progress = FALSE)
```

Group the performance measures by both conditions with the `by` argument. The
`data_grid` columns surface in the summary with a `data_` prefix, so we group
by `data_n` and `data_beta`:

```{r}
pm_design <- performance_measures(
  design_result, estimand = "x", by = c("data_n", "data_beta")
)
pm_design
```

Read each row as one condition cell. As `n` grows, the empirical SE and MSE
should shrink (more data, tighter inference); as `beta` grows the signal
strengthens. Coverage should stay near 0.95 throughout for the conjugate
fitter. The MCSE column tells you how much trust to put in each cell at 60
replicates — a reminder that small condition x condition grids with few
replicates produce noisy performance estimates.

Plotting coverage across conditions gives the visual version of the coverage
rows. `plot_coverage()` plots each condition x estimand cell as a point with
MCSE error bars against the nominal rate:

```{r coverage-by-condition}
plot_coverage(design_result, by = c("data_n", "data_beta"))
```

For a fuller picture, `plot_recovery()` shows per-task posterior-mean
estimates against the truth, faceted by a condition column — a direct visual
check that the estimator recovers the estimand across the design:

```{r recovery}
plot_recovery(design_result, estimand = "x", by = "data_n")
```

The parameter argument is named `estimand`, the same term
`performance_measures()` uses (`var =` remains accepted as a compatibility
alias).

# Summary

The Morris et al. (2019) framework, in bayesim, maps to a small set of typed
primitives:

- **Estimands** are recorded by generators as `true_params` and surfaced as
  `truth__*` columns.
- **DGPs** are generators (`fixed_truth_generator()`, `prior_predictive_generator()`,
  `ifs_generator()`, or an inline function).
- **Performance measures** with MCSEs come from `performance_measures()`.
- **Replicate sizing** comes from `n_replicates_for_target()`.
- **Varying conditions** are rows/columns of the `data_grid`, grouped in the
  analysis layer via `by`.

For the calibration-specific counterpart to performance coverage — checking
the *whole* posterior for self-consistency rather than a single interval —
see `vignette("sbc-and-calibration")`.

# Reference

Morris, White & Crowther (2019), *Using simulation studies to evaluate
statistical methods*, Statistics in Medicine 38(11): 2074-2102.
