---
title: "Experimental grammar: declare, run, reuse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimental grammar: declare, run, reuse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
source(system.file("experimental", "study-grammar.R", package = "bayesim"))
```

Describe the experiment with ordinary R functions. The runner supplies shared
data, deterministic random streams, saved results and comparison groups.
This experimental interface is sourced explicitly; the supported package API
still starts with `simulation_config()`.

```text
condition × replicate → one dataset → each method
                                       ↓
                              extract → measure → compare
                                       ↓             ↓
                              retained artifacts + results
                                       ↓
                              new measures and assessment
```

This example runs a small normal-mean study. Observations have known unit
variance. Two methods differ only in their normal prior's scale. Their
posteriors are analytic, so this vignette needs no sampler.

```{r declaration}
normal_method <- function(prior_sd) {
  study_method(
    fit = function(data, context) {
      precision <- length(data) + 1 / context$settings$prior_sd^2
      list(mean = sum(data) / precision, sd = sqrt(1 / precision))
    },
    extract = list(draws = function(fit, data, context) {
      rnorm(400, fit$mean, fit$sd)
    }),
    settings = list(prior_sd = prior_sd)
  )
}

experiment <- study(
  "normal-mean",
  conditions = data.frame(condition_id = c("small", "large"), n = c(10L, 50L)),
  generate = function(condition, context) rnorm(condition$n, mean = 1)
) |>
  with_methods(regularized = normal_method(1), diffuse = normal_method(10)) |>
  with_measure("estimate", function(artifacts, context) {
    data.frame(target = "mean", value = mean(artifacts$draws))
  }, needs = "draws") |>
  with_retention("draws")

plan_study(experiment, replicates = 3)
```

The draws here are independent posterior samples. An MCMC extractor should
preserve chains and retain any diagnostics needed for later selection.

```{r run}
run_path <- tempfile("normal-mean-")
first <- run_study(experiment, replicates = 3, path = run_path)
first$measurements[, c("condition_id", "replicate", "method_id", "value")]
```

Add a 90% interval and two replicates. The original fits stay cached; the new
measurement uses their saved draws. New replicates get new datasets and fits.

```{r reuse}
extended <- experiment |>
  with_measure("interval90", function(artifacts, context) {
    interval <- quantile(artifacts$draws, c(0.05, 0.95))
    data.frame(target = "mean", lower = interval[[1]], upper = interval[[2]])
  }, needs = "draws")

later <- run_study(extended, replicates = 5, path = run_path)
stopifnot(identical(first$attempts, later$attempts[later$attempts$replicate <= 3, ]))
table(later$measurements$measurement)
```

This reuse requires unchanged scientific dependencies. Cache keys track function
bodies, settings and explicit versions, but cannot detect changed global helpers,
package versions or external files. Increment the relevant declaration's
`version` when those dependencies change.

The case studies test the boundaries of this interface:

| Study | What the declaration must express |
|---|---|
| [Choice of likelihood](experimental-likelihood-study.html) | Full candidate grids, shared datasets, predictive comparisons and changing eligibility |
| [Primed priors](experimental-primed-priors.html) | Preparation shared across inner replicates and data reused in refits |
| [Method comparison](experimental-method-comparison.html) | Direct CmdStan methods, sampler settings and pointwise comparison inputs |

Those vignettes build plans without running their expensive fits. This runner
is a reference implementation: individual RDS records and in-memory result tables
still need replacement or validation for a study with millions of fits. It also
requires exclusive access to a run directory. Its purpose is to test whether the
study declaration captures the scientific decisions before committing to a
production execution and storage design.

```{r cleanup, include=FALSE}
unlink(run_path, recursive = TRUE)
```
