---
title: "Experimental grammar: a likelihood comparison study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Experimental grammar: a likelihood comparison study}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
source(system.file("experimental", "study-grammar.R", package = "bayesim"))
source(system.file("experimental", "likelihood-case.R", package = "bayesim"))
```

How do likelihood, link and predictor choices affect prediction and coefficient
recovery? The [choice-of-likelihood study](https://github.com/sims1253/choice-of-likelihood-paper/blob/2a96b0d6ac668739363f98f1211c4b0cbc6b9069/README.md#L97-L131)
compares these choices on positive and unit-interval responses. This vignette
expresses its configured designs with an experimental study grammar. Knitting
builds plans and exercises a comparison on invented scores; it fits no models.

The design tables follow the study's pinned source. The executable generator
and brms adapter below cover a smaller, explicitly defined example. They are
not an exact reproduction of the historical simulations. The source scripts
rely on historical helpers and custom families that this example does not
restore. These declarations are separate from the supported
`simulation_config()` engine.

The helper file is installed with the package and contains ordinary R functions:

```{r helper-path}
system.file("experimental", "likelihood-case.R", package = "bayesim")
```

## Conditions and candidates

The full positive-response design contains eight generating families, three
response shapes, and null or nonzero effects. Its filters constrain the allowed
links. The unit-interval design contains six generating families and its own
shape-specific coefficient tables. Each candidate combines a fitting family,
a link and one of five predictor formulas. The helper retains all source
coefficient and dispersion values, including the altered structural coefficients
for the unit-interval bathtub shape. [Positive source](https://github.com/sims1253/choice-of-likelihood-paper/blob/2a96b0d6ac668739363f98f1211c4b0cbc6b9069/simulation_study/cont_positive/continuous_positive_simulation.R#L5-L112),
[unit-interval source](https://github.com/sims1253/choice-of-likelihood-paper/blob/2a96b0d6ac668739363f98f1211c4b0cbc6b9069/simulation_study/unit_interval/unit_interval_simulation.R#L5-L198).

```{r designs}
positive <- likelihood_design("positive")
unit <- likelihood_design("unit")
stopifnot(
  nrow(positive$conditions) == 84L, nrow(positive$candidates) == 85L,
  nrow(unit$conditions) == 72L, nrow(unit$candidates) == 80L
)
positive$conditions[1:3, c("condition_id", "data_N", "dataset_N",
                          "x_y_coef", "sigma_y", "y_intercept")]
unique(positive$candidates$formula)
```

`data_N = 100` specifies observations; `dataset_N = 200` specifies dataset
replicates. The historical runner [generated one dataset before looping over
candidate fits](https://github.com/sims1253/bayesim/blob/1617348fd6b60c0862ea90539a9b4b83a251a46f/R/simulation.R#L90-L119).
The experimental grammar also puts methods under a shared dataset identity.
The cited historical commit establishes that structure; it does not establish
which version produced the study's saved results.

The source leaves some link helpers implicit. Here the inverse-softplus link
is `log(expm1(mu))`, logit is `qlogis(mu)`, cauchit is `qcauchy(mu)`, and
complementary log-log is `log(-log1p(-mu))`. These are explicit interpretations
used to resolve the coefficient tables, not verified copies of those helpers.

## Literal links and coefficient meaning

A lognormal candidate has literal link `identity` in the source, but its
coefficient acts on a logarithmic scale. The postprocessing [records effective
links separately](https://github.com/sims1253/choice-of-likelihood-paper/blob/2a96b0d6ac668739363f98f1211c4b0cbc6b9069/simulation_study/cont_positive/models/postprocess_sims_positive.R#L18-L46).
Both fields belong in the declaration:

```{r effective-links}
unique(positive$candidates[c("fit_family", "fit_link", "effective_link")])
```

For the illustrative generator, recovery of the generating `x` coefficient is
reported only when the candidate's effective link matches the generating
link. This is an explicit estimand convention for this example. Matching names
or links alone does not establish coefficient comparability for another study.
Predictive scores can still enter a comparison when recovery is unavailable.

## Scientific functions and artifact needs

`likelihood_generate_subset(condition, context)` implements an illustrative
structural model. It generates independent normal `z1`, `z2` and `z3`, then
sets `x` from `z1`, `z3` and normal noise. Its response predictor is
`y_intercept + x_y_coef*x + z1_y_coef*z1 + z2_y_coef*z2`. Gamma/log conditions
use mean `exp(eta)` and shape `sigma_y`; Beta/logit conditions use mean
`plogis(eta)` and precision `sigma_y`. Finally, `z4` depends on `x`, `y` and
normal noise. Training and testing data use separate draws of this model.

This generator supports Gamma/log and Beta/logit only. It omits the source's
response-bound rejection/resampling settings. It stops on other conditions.
The full grids therefore document planning coverage; the subset below
identifies execution coverage. A replacement generator must state its own
parameterization and handling of response bounds.

`likelihood_fit(data, context)` calls `brms::brm()` using the method's formula,
family, link and seed. It uses CmdStan, four chains and 2,000 iterations. These
are example settings, not recovered historical settings. It requires brms and
CmdStan when executed; unsupported custom families report an error. Namespace
calls sit inside functions, so planning loads neither backend nor models.

The extractors produce chain-preserving slope draws, sampler diagnostics and
pointwise LOO scores. `likelihood_measure()` computes bias, posterior root mean
squared error against truth, sign probability, interval exclusion of zero and
summed LOO ELPD. The posterior error quantity is explicitly
`sqrt(mean((draws - truth)^2))`; its `rmse_s` name supports the example policy
but does not certify equivalence to an unavailable historical metric helper.
These choices are ordinary functions, rather than inferred metric names.

```{r declarations}
positive_study <- likelihood_declaration(positive, "positive-likelihoods")
unit_study <- likelihood_declaration(unit, "unit-likelihoods")

positive_plan <- plan_study(positive_study, replicates = 200L, seed = 1235813L)
unit_plan <- plan_study(unit_study, replicates = 200L, seed = 1235813L)

configured_counts <- data.frame(
  domain = c("positive", "unit interval"),
  conditions = c(nrow(positive$conditions), nrow(unit$conditions)),
  replicates = 200L,
  candidates = c(nrow(positive$candidates), nrow(unit$candidates))
)
configured_counts$fits <- with(configured_counts, conditions * replicates * candidates)
configured_counts
```

These are configured combinations, not observed completed fits or promises
that all historical family/link combinations are supported. The counts include
Gaussian with every listed link because that is what the source filters allow.
`plan_study()` validates declarations and artifact dependencies. It does not
call the generator, validate a backend family, prepare a model or sample a prior.

A smaller declaration has a fully specified generator and built-in candidate
families. It keeps all five source formulas and both null and nonzero effects:

```{r executable-subsets}
gamma_example <- list(
  conditions = subset(positive$conditions, data_family == "gamma" & data_link == "log"),
  candidates = subset(positive$candidates, fit_family == "gamma" & fit_link == "log")
)
beta_example <- list(
  conditions = subset(unit$conditions, data_family == "beta" & data_link == "logit"),
  candidates = subset(unit$candidates, fit_family == "beta" & fit_link == "logit")
)
gamma_study <- likelihood_declaration(gamma_example, "illustrative-gamma")
beta_study <- likelihood_declaration(beta_example, "illustrative-beta")
gamma_plan <- plan_study(gamma_study, replicates = 2L, seed = 42L)
beta_plan <- plan_study(beta_study, replicates = 2L, seed = 42L)
stopifnot(nrow(gamma_example$conditions) == 6L,
          nrow(beta_example$conditions) == 6L,
          nrow(gamma_example$candidates) == 5L,
          nrow(beta_example$candidates) == 5L)
```

## A comparison is a grouped calculation

The declaration uses `by = c("dataset_id", "method_formula")`. Within each
group, `likelihood_elpd_gap()` subtracts the best finite ELPD from each score.
It returns the size of the reference set alongside each gap. This is a
measurement-row comparison; it needs no retained native fit.

The source [filters diagnostics, missing values and extreme errors before
computing its maxima](https://github.com/sims1253/choice-of-likelihood-paper/blob/2a96b0d6ac668739363f98f1211c4b0cbc6b9069/simulation_study/cont_positive/models/postprocess_sims_positive.R#L49-L98).
Changing that policy can change the reference model. This cheap example uses
invented scores, where the best predictor has an unacceptable R-hat:

```{r selection-example}
example_rows <- data.frame(
  method_id = c("A", "B", "C"), measurement = "performance",
  value = c(0.1, 0.2, 0.3), rmse_s = 0.5, comparable = TRUE,
  elpd_loo = c(-102, -100, -99), rhat = c(1, 1, 1.1),
  ess_bulk = 800, ess_tail = 800, divergents = 0
)
raw_gaps <- likelihood_elpd_gap(example_rows, list(), list())
selected <- likelihood_selection(example_rows, data.frame())
selected_gaps <- likelihood_elpd_gap(selected, list(), list())
raw_gaps
selected_gaps
stopifnot(identical(raw_gaps$value, c(-3, -1, 0)),
          identical(selected_gaps$value, c(-2, 0)))
```

`likelihood_selection(measurements, attempts)` states a recovery-analysis
policy: comparable coefficients, finite scores and recovery measures, at most
10 divergences, R-hat at most 1.01, both ESS measures above 400, absolute bias
below 10 and posterior error below 10. This example does not reproduce every
historical filter: it omits predictive RMSE, new-data ELPD and later Pareto-k
filters. Keep those choices explicit when extending it.

After an authorized run, `assess_study(run, policy = likelihood_selection)`
keeps raw rows and exclusions, then reruns this comparison on selected rows.
Failed attempts remain in the run's attempt records. Downstream brms, lmer or
custom plotting code can use the resulting tables and dataset identities.
No downstream analysis is needed to construct these plans.
