---
title: "Getting Started with bayesim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bayesim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

bayesim is a modern simulation framework for Bayesian modeling studies. It provides:

- **Deterministic reproducibility** across sequential, parallel, and resumed runs
- **Checkpoint/resume** capabilities for long-running simulations
- **Memory-bounded execution** with configurable artifact retention
- **Extensible interfaces** for custom fitters and metrics

# Basic Usage

## Setting Up a Simulation

```{r setup}
library(bayesim)
```

### Define a Data Generator

The data generator creates synthetic datasets for your simulation:

```{r}
my_data_generator <- function(data_spec, seed, task_ctx) {
  set.seed(seed)
  n <- data_spec$n
  
  # Generate synthetic data
  x <- rnorm(n)
  y <- data_spec$intercept + data_spec$slope * x + rnorm(n, sd = data_spec$sigma)
  
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(intercept = data_spec$intercept, slope = data_spec$slope, sigma = data_spec$sigma),
    vars_of_interest = c("intercept", "slope", "sigma"),
    references = c(intercept = 0, slope = 0, sigma = 1),
    meta = list()
  )
}
```

### Create a Configuration

```{r}
config <- simulation_config(
  data_grid = data.frame(
    n = c(100, 500),
    intercept = 1,
    slope = 2,
    sigma = 1
  ),
  fit_grid = data.frame(
    model = "linear"
  ),
  data_generator = my_data_generator,
  fitter = MockFitter(),  # Use mock for this example
  metrics = c("rmse", "bias"),
  n_replicates = 10L,
  seed = 42L
)
```

### Run the Simulation

```{r}
result <- run_simulation(config, progress = FALSE)
```

### Examine Results

```{r}
print(result)

# Summary tibble
head(result$summary)
```

# Checkpointing and Resume

For long-running simulations, use checkpointing:

```{r, eval=FALSE}
config <- simulation_config(
  data_grid = my_data_grid,
  fit_grid = my_fit_grid,
  data_generator = my_data_generator,
  fitter = my_fitter,
  metrics = c("rmse"),
  n_replicates = 1000L,
  seed = 42L,
  result_path = "my_simulation",
  checkpoint_every = 50L
)

# Run (can be interrupted and resumed)
result <- run_simulation(config, resume = TRUE)
```

If interrupted, resume with:

```{r, eval=FALSE}
result <- run_simulation(config, resume = TRUE)
```

# Custom Metrics

Create custom metrics by extending the Metric class:

```{r}
MyMetric <- S7::new_class(
  "MyMetric",
  parent = Metric,
  properties = list(
    name = S7::new_property(S7::class_character, default = "my_metric"),
    needs = S7::new_property(S7::class_character, default = "predictions"),
    required = S7::new_property(S7::class_logical, default = FALSE)
  )
)

S7::method(compute, MyMetric) <- function(metric, fit_result, data_bundle, context, task_ctx) {
  # Your metric computation
  list(value = 0.5)
}

# Register and use
register_metric(MyMetric())
```

# Next Steps

- See `vignette("custom-fitters")` for creating custom model fitters
- See `vignette("reproducibility")` for understanding determinism guarantees
- See `vignette("memory-management")` for handling large simulations
