---
title: "Getting Started with bayesim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bayesim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

This vignette defines a data generator, runs a Bayesian linear regression
study, and summarizes its results. It uses `LinearRegressionFitter()` for
exact conjugate inference without Stan. For Stan models, use `BrmsFitter()`
or `CmdStanFitter()`.

# Run a study

```{r setup}
library(bayesim)
```

## Generate data

A data generator is a function with the signature
`(data_spec, task_ctx)` returning a named list. `data_spec` contains one row
of `data_grid`; `task_ctx` identifies the task. Use R's random functions
normally. Set the study seed in `simulation_config()`, not inside the generator.

```{r}
my_data_generator <- function(data_spec, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(n)
  y <- data_spec$intercept + data_spec$slope * x +
    stats::rnorm(n, sd = data_spec$sigma)
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(
      Intercept = data_spec$intercept,
      x = data_spec$slope,
      sigma = data_spec$sigma
    ),
    vars_of_interest = c("Intercept", "x", "sigma")
  )
}
```

For common patterns (fixed truth, prior-predictive, inverse forward sampling)
use the factory constructors `fixed_truth_generator()`,
`prior_predictive_generator()`, `ifs_generator()`.

## Choose conditions and metrics

```{r}
config <- simulation_config(
  data_grid = data.frame(
    n = c(50, 100),
    intercept = 1,
    slope = 2,
    sigma = 1
  ),
  fit_grid = data.frame(model = "linear"),
  data_generator = my_data_generator,
  fitter = LinearRegressionFitter(n_draws = 500L),
  metrics = list(
    posterior_summary_metric(),
    sampler_diagnostics_metric()
  ),
  n_replicates = 4L,
  seed = 42L
)
```

## Run the simulation

```{r}
result <- run_simulation(config, progress = FALSE)
print(result)
```

## Inspect task results

```{r}
result$summary[c("task_id", "data_n", "status", "posterior_summary__mean__x")]
```

Each row is one task. Columns include `task_id`, `status`, `timing_total`,
the grid columns, and one column per metric field
(`posterior_summary__mean__x`, `truth__x`, ...). The `truth__*` columns record
the data-generating truth for parameter-recovery analysis.

# Summarize results

Each sample size is a separate condition. Group by `data_n` to keep their
results separate. Select the slope's posterior mean to keep the table small:

```{r}
summarize_simulation(
  result,
  by = "data_n",
  metrics = "posterior_summary__mean__x"
)
```

The output reports the mean, median, standard deviation, and Monte Carlo
standard error (MCSE) across replicates, plus task and failure counts.
Use `metric_cols(result, "posterior_summary")` to find other metric columns.

# Performance measures

Use `performance_measures()` for bias, empirical SE, MSE, coverage, and
average model SE, each with its MCSE (Morris, White & Crowther, 2019):

```{r}
pm <- performance_measures(result, estimand = "x", by = "data_n")
pm
```

This pairs `truth__x` with the per-task posterior summary to give, per
condition cell, the bias and calibration of the estimator.

# Run a larger study

This example uses four replicates per condition to keep it quick. Choose
more replicates for a precise estimate; the
[study design guide](design-of-simulation-studies.html) explains how MCSE
helps you decide.

Add `workers = 4` to `run_simulation()` to use four parallel workers.
For long runs, set `result_path` and `checkpoint_every` when constructing
the configuration. Resume interrupted work with
`resume_simulation("my_simulation", config = config)`, using the original
configuration. Script-defined generators require that configuration because
the saved manifest cannot reconstruct their closures.

See [parallel runs and checkpoints](parallel-and-hpc.html) for a complete
example, and [reproducibility](reproducibility.html) for seed and resume
requirements.

# Next steps

- `vignette("design-of-simulation-studies")` for aims, estimands, and choosing
  the number of replicates
- `vignette("sbc-and-calibration")` for the calibration workflow
- `vignette("brms-studies")` for Stan-backed studies and `model_grid()`
- `vignette("custom-fitters")` / `vignette("custom-metrics")` for extending
  the framework
- `vignette("parallel-and-hpc")` and `vignette("reproducibility")` for large
  runs and the determinism guarantees
