---
title: "Getting Started with bayesim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with bayesim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

bayesim is a simulation framework for Bayesian modeling studies. It provides:

- **Reproducible execution** across sequential, parallel (mirai), and resumed runs
- **A model bank** that compiles each distinct Stan model once and reuses it
  across tasks, eliminating per-task recompilation
- **Checkpoint/resume** capabilities for long-running simulations
- **Memory-bounded execution** with configurable artifact retention
- **Typed generators, metrics, and an analysis layer** -- no hand-rolled dplyr

This vignette runs end-to-end with the built-in `MockFitter` (no Stan required).
For real Bayesian inference, swap `MockFitter()` for `BrmsFitter()`.

# Basic Usage

## Setup

```{r setup}
library(bayesim)
```

## Define a Data Generator

A data generator is a function with the signature
`(data_spec, task_ctx)` returning a named list. It must consume the
**ambient** RNG state (the engine restores a per-task L'Ecuyer stream before
each call), so do not re-seed inside.

```{r}
my_data_generator <- function(data_spec, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(n)
  y <- data_spec$intercept + data_spec$slope * x +
    stats::rnorm(n, sd = data_spec$sigma)
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(
      intercept = data_spec$intercept,
      slope = data_spec$slope,
      sigma = data_spec$sigma
    ),
    vars_of_interest = c("intercept", "slope", "sigma"),
    references = c(intercept = 0, slope = 0, sigma = 1)
  )
}
```

For common patterns (fixed truth, prior-predictive, inverse forward sampling)
use the factory constructors `fixed_truth_generator()`,
`prior_predictive_generator()`, `ifs_generator()`.

## Create a Configuration

```{r}
config <- simulation_config(
  data_grid = data.frame(
    n = c(50, 100),
    intercept = 1,
    slope = 2,
    sigma = 1
  ),
  fit_grid = data.frame(model = "linear"),
  data_generator = my_data_generator,
  fitter = MockFitter(),
  metrics = list(
    rmse_metric(),
    bias_metric(),
    posterior_summary_metric(),
    convergence_metric()
  ),
  n_replicates = 4L,
  seed = 42L
)
```

## Run the Simulation

```{r}
result <- run_simulation(config, progress = FALSE)
print(result)
```

## Examine Results

```{r}
head(result$summary)
```

Each row is one task. Columns include `task_id`, `status`, `timing_total`,
the grid columns, and one column per metric field
(`rmse__value`, `bias__value`, `posterior_summary__mean__slope`, ...).

# Summarizing Results

`summarize_simulation()` aggregates the per-task summary by condition, with
Monte Carlo standard errors (MCSE):

```{r}
agg <- summarize_simulation(result, by = c("n"))
agg
```

For each metric you get `<metric>_mean`, `<metric>_median`, `<metric>_sd`, and
`<metric>_mcse`, plus `n_reps`, `n_failed`, and `failure_rate`.

# Parallel Execution (mirai)

Parallelism is owned by the engine and uses [mirai](https://mirai.r-lib.org/).
Set daemons before calling `run_simulation()`; otherwise execution is
sequential.

```{r, eval=FALSE}
mirai::daemons(4)
result <- run_simulation(config, progress = FALSE)
mirai::daemons(0) # reset to sequential
```

`SimulationConfig` accepts a `daemon_setup` function run once per daemon (e.g.
to configure the cmdstan path when using `BrmsFitter` on a cluster).

# Checkpointing and Resume

For long-running simulations, write checkpoints to `result_path`:

```{r, eval=FALSE}
config <- simulation_config(
  data_grid = data_grid,
  fit_grid = fit_grid,
  data_generator = my_data_generator,
  fitter = MockFitter(),
  metrics = list(rmse_metric()),
  n_replicates = 1000L,
  seed = 42L,
  result_path = "my_simulation",
  checkpoint_every = 50L
)
result <- run_simulation(config, resume = "auto")
```

If interrupted, resume with:

```{r, eval=FALSE}
result <- resume_simulation("my_simulation")
```

# Custom Metrics

Extend the `Metric` class and implement `compute()`:

```{r}
MyMetric <- S7::new_class(
  "MyMetric",
  parent = Metric,
  properties = list(
    name = S7::new_property(S7::class_character, default = "my_metric"),
    needs = S7::new_property(S7::class_character, default = "predictions"),
    required = S7::new_property(S7::class_logical, default = FALSE)
  )
)

S7::method(compute_metric, MyMetric) <- function(metric, fit_result, data_bundle, context, task_ctx) {
  list(value = 0.5)
}

config <- simulation_config(
  data_grid = data.frame(n = 100),
  fit_grid = data.frame(model = "test"),
  data_generator = my_data_generator,
  fitter = MockFitter(),
  metrics = list(MyMetric()),
  n_replicates = 2L,
  seed = 42L
)
```

The built-in metric library covers the standard surface: `bias_metric`,
`rmse_metric`, `mae_metric`, `mse_metric`, `coverage_metric`,
`pos_prob_metric`, `posterior_summary_metric`, `convergence_metric`,
`sampler_diagnostics_metric`, `rank_metric` (SBC ranks), `rstar_metric`, and
LOO/test-set variants (`elpd_loo_metric`, `rmse_loo_metric`, `r2_loo_metric`,
`elpd_test_metric`, `rmse_test_metric`, `r2_test_metric`).

# Reproducibility

Same seed + same package/backend versions + same platform produces identical
results. Across platforms (or Stan version changes) results are statistically
equivalent but may differ in the least significant bits due to floating-point
and backend differences.

# Next Steps

- `vignette("custom-fitters")` for creating custom model fitters (incl. brms)
- `vignette("reproducibility")` for the determinism guarantees
- `vignette("memory-management")` for large simulations
