---
title: "HPC and remote daemons"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{HPC and remote daemons}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{bayesim}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bayesim)
```

# Introduction

bayesim parallelizes simulation tasks across [mirai](https://mirai.r-lib.org/)
daemons. The `workers=` argument to `run_simulation()` covers the common case of
local multi-core parallelism: it sets up `mirai::daemons(workers)` for the run
and tears them down on exit.

High-performance computing (HPC) clusters are different. Compute happens across
many nodes, the dispatcher and the workers do not share a process, and the
dispatcher must reach its workers over a network. For that, mirai exposes a URL
scheme that lets daemons live on remote nodes and connect back to a dispatcher
on the head node. bayesim is designed to slot into that workflow: set up your own
daemons before calling `run_simulation()`, pass `workers = NULL`, and bayesim
will dispatch its task grid over them.

This vignette walks through the moving parts: the local-recap, the mirai URL
scheme on SLURM, TLS configuration, the `daemon_setup` hook for per-daemon
initialization (e.g. the cmdstan path), and checkpoint/resume across SLURM
submissions.

# Local parallelism

For local multi-core work, pass `workers` to `run_simulation()`:

```{r, eval=FALSE}
result <- run_simulation(config, workers = 4)
# mirai::daemons(4) is called for you and torn down on exit.
```

This is the simplest path and is also the recommended path for development and
for moderate task counts. The rest of this vignette is only relevant when you
need daemons that the `workers=` argument cannot manage itself: daemons on
remote nodes, behind TLS, or with per-daemon setup.

# Remote daemons (mirai URL scheme)

mirai's URL scheme splits the work into two roles:

- A **dispatcher** (the controller process) that calls `mirai::daemons(url, n)`
  to advertise `n` worker slots at a URL.
- **Daemons** (the workers), launched separately, that each call
  `mirai::daemon(def, url, ...)` to connect back to the dispatcher and pull
  tasks.

This split maps cleanly onto a batch scheduler: the dispatcher runs on the head
node (or wherever `srun` puts the driver), and the daemons run on the compute
nodes.

```{r, eval=FALSE}
library(bayesim)
library(mirai)

# On the dispatcher: advertise 16 worker slots at a TLS-secured URL.
url <- "tls://10.0.0.1:5555"
daemons(url = url, n = 16)

# Now run bayesim with workers = NULL so it uses the daemons you set up.
result <- run_simulation(config, workers = NULL)

# Tear down when finished.
daemons(0)
```

## A SLURM submission script

A typical pattern is one SLURM job that launches `N` daemon workers across
nodes, then runs the dispatcher R script on the head node. The script below uses
`tasks` (one task per CPU) and `srun` to launch a daemon on each task slot.

```bash
#!/bin/bash
#SBATCH --job-name=bayesim-sim
#SBATCH --nodes=4
#SBATCH --tasks-per-node=4
#SBATCH --cpus-per-task=1
#SBATCH --time=12:00:00
#SBATCH --output=sim-%j.out

module load R
module load gcc

# TLS material (see "TLS configuration" below).
export BAYESIM_TLS=/scratch/$USER/bayesim-tls

# Dispatcher URL: the head node's primary address, on a fixed port.
DISPATCHER_URL="tls://$(hostname -I | awk '{print $1}'):5555"
N_DAEMONS=$((SLURM_NTASKS - 1))   # reserve one task for the dispatcher

# Launch one daemon per remaining task, across all allocated nodes.
srun --ntasks=$N_DAEMONS --overlap Rscript -e "
  mirai::daemon('dispatcher.R', url = '$DISPATCHER_URL',
                tls = mirai::tls_config(
                  ca = '$BAYESIM_TLS/ca.pem',
                  key = '$BAYESIM_TLS/server-key.pem',
                  cert = '$BAYESIM_TLS/server.pem'))
" &

# Give the daemons a moment to connect.
sleep 10

# Run the dispatcher on the head node.
Rscript dispatcher.R "$DISPATCHER_URL" "$N_DAEMONS"

wait
```

And the dispatcher R script:

```{r, eval=FALSE}
# dispatcher.R
args <- commandArgs(trailingOnly = TRUE)
url <- args[1]
n <- as.integer(args[2])

library(bayesim)
library(mirai)

mirai::daemons(url = url, n = n)

config <- simulation_config(
  data_grid = my_data_grid,
  fit_grid = my_fit_grid,
  data_generator = my_data_generator,
  fitter = BrmsFitter(...),
  metrics = list(pred_rmse_metric(), coverage_metric()),
  n_replicates = 500L,
  seed = 42L,
  result_path = "sim-checkpoints"
)

result <- run_simulation(config, workers = NULL)

mirai::daemons(0)
```

The exact `srun`/socket plumbing depends on your cluster's network topology;
the load-bearing idea is the dispatcher/daemon split.

# TLS configuration

Remote daemons should connect over TLS. mirai uses standard PEM material; see
the [mirai security
documentation](https://mirai.r-lib.org/articles/tls.html#secure-connections-1)
for the full reference. A self-signed CA plus server/client certs is enough to
get started:

```{r, eval=FALSE}
# Run once, on a trusted host. Keep the private keys private.
openssl req -x509 -newkey rsa:2048 -nodes -days 365 \
  -keyout ca-key.pem -out ca.pem \
  -subj "/CN=bayesim-CA"

openssl req -newkey rsa:2048 -nodes \
  -keyout server-key.pem -out server-csr.pem \
  -subj "/CN=$(hostname)"

openssl x509 -req -in server-csr.pem -CA ca.pem -CAkey ca-key.pem \
  -CAcreateserial -out server.pem -days 365

# Distribute ca.pem, server.pem, and server-key.pem to every node
# that will run a daemon or the dispatcher.
```

Then pass the material into both sides:

```{r, eval=FALSE}
# Dispatcher
tls <- mirai::tls_config(
  ca = "ca.pem",
  key = "server-key.pem",
  cert = "server.pem"
)
mirai::daemons(url = "tls://...", n = 16, tls = tls)

# Each daemon
mirai::daemon(url = "tls://...", tls = tls)
```

# The daemon_setup hook

Daemons start from a fresh R process. Anything the workers need at run time ---
the cmdstan install path, environment modules, a warm model cache --- must be
set up on each daemon before tasks start. bayesim exposes this through the
`daemon_setup` argument of `simulation_config()`, which is run once per daemon
via `mirai::everywhere()` before the task grid ships.

```{r, eval=FALSE}
config <- simulation_config(
  data_grid = my_data_grid,
  fit_grid = my_fit_grid,
  data_generator = my_data_generator,
  fitter = BrmsFitter(precompile = TRUE),
  metrics = list(pred_rmse_metric(), coverage_metric()),
  n_replicates = 500L,
  seed = 42L,
  result_path = "sim-checkpoints",
  daemon_setup = function() {
    # Point cmdstanr at the cluster's cmdstan install (modules are usually
    # loaded by the SLURM environment; the path still needs to be set in R).
    cmdstanr::cmdstan_path_set("/opt/cmdstan-2.35.0")
    # Optionally seed package-level caches, set options, etc.
  }
)

# With daemons already set up via mirai::daemons(url = "tls://...", n = 16):
result <- run_simulation(config, workers = NULL)
```

`daemon_setup` is ignored when no daemons are set, so the same config works for
local and HPC runs.

# Checkpointing across submissions

SLURM jobs can be preempted or hit a wall-clock limit mid-run. Point
`result_path` at a directory on shared scratch and pass `resume = "auto"` so an
interrupted run picks up where it left off:

```{r, eval=FALSE}
config <- simulation_config(
  data_grid = my_data_grid,
  fit_grid = my_fit_grid,
  data_generator = my_data_generator,
  fitter = BrmsFitter(precompile = TRUE),
  metrics = list(pred_rmse_metric(), coverage_metric()),
  n_replicates = 2000L,
  seed = 42L,
  result_path = "/scratch/$USER/bayesim-sim",
  checkpoint_every = 50L,
  daemon_setup = function() cmdstanr::cmdstan_path_set("/opt/cmdstan-2.35.0")
)

# Each submission either starts fresh or resumes from the shared checkpoint.
result <- run_simulation(config, resume = "auto", workers = NULL)
```

The config fingerprint is written into the checkpoint manifest, so a resumed run
is rejected if you change the design (data/fit grid, seed, generator signature)
between submissions. For a forced resume from a specific directory, use
`resume_simulation()`:

```{r, eval=FALSE}
result <- resume_simulation("/scratch/$USER/bayesim-sim")
```

# Next steps

- `vignette("getting-started")` for the basics of `simulation_config()` and
  `run_simulation()`.
- `vignette("reproducibility")` for why resuming produces identical results to
  an uninterrupted run.
- `vignette("memory-management")` for keeping memory flat across very large
  task grids.
- `vignette("targets")` for running bayesim as a step in a `targets` pipeline.
