---
title: "Memory Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Memory Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Memory management

Large simulation studies (thousands of tasks, each with a full posterior and
predictions) can exhaust memory. bayesim addresses this with **retention
profiles** (control what each task keeps) and **chunked execution**
(control how many tasks are in flight).

## Retention profiles

The `retain` argument of `simulation_config()` selects which artifacts each
task result keeps. The standard profiles are:

| profile | keeps | drops |
|---|---|---|
| `"metrics"` | computed metric values | fit object, draws, predictions |
| `"diagnostics"` | metrics + sampler diagnostics | fit object, draws |
| `c("metrics", "diagnostics")` (default) | metrics + diagnostics | fit object, full draws |
| `"full"` | everything (fit object, draws, predictions) | nothing |

For a metrics-only sweep over thousands of conditions, use `retain = "metrics"`
to keep memory flat regardless of posterior size:

```{r, eval=FALSE}
config <- simulation_config(
  data_grid = big_data_grid,
  fit_grid = big_fit_grid,
  data_generator = my_generator,
  fitter = BrmsFitter(),
  metrics = list(rmse_metric(), bias_metric(), convergence_metric()),
  n_replicates = 100L,
  seed = 42L,
  retain = "metrics" # keep only metric values; drop the fit and draws
)
```

The default (`c("metrics", "diagnostics")`) keeps sampler diagnostics
(r-hat, divergences) alongside metrics but drops the fit object and full
posterior draws -- the right tradeoff for most studies.

Use `"full"` when you need per-task posteriors (e.g. for offline SBC rank
computation), at the cost of higher memory.

## Chunked execution

`chunk_size` controls how many tasks are dispatched and held in memory at
once. The engine processes tasks in batches of `chunk_size`, lightening each
batch's results (per the retention profile) before starting the next.

```{r, eval=FALSE}
config <- simulation_config(
  data_grid = big_data_grid,
  fit_grid = big_fit_grid,
  data_generator = my_generator,
  fitter = BrmsFitter(),
  metrics = list(rmse_metric()),
  n_replicates = 1000L,
  seed = 42L,
  retain = "metrics",
  chunk_size = 50L,        # process 50 tasks per batch
  checkpoint_every = 50L   # write a checkpoint after each batch
)
```

## Checkpointing for long runs

For very long runs, combine `result_path` with `checkpoint_every` so an
interrupted run can resume from the last checkpoint without losing work:

```{r, eval=FALSE}
config <- simulation_config(
  data_grid = big_data_grid,
  fit_grid = big_fit_grid,
  data_generator = my_generator,
  fitter = BrmsFitter(),
  metrics = list(rmse_metric()),
  n_replicates = 1000L,
  seed = 42L,
  result_path = "results/big_study",
  checkpoint_every = 50L,
  chunk_size = 50L,
  retain = "metrics"
)

# Interruptible: can be stopped and resumed
result <- run_simulation(config, resume = "auto")

# Resume after an interruption
# result <- resume_simulation("results/big_study")
```

## Parallel memory considerations

When using mirai daemons, each daemon holds its own copies of dispatched
tasks' artifacts. `chunk_size` bounds the in-flight work per daemon, and the
retention profile bounds per-task size. For memory-constrained HPC nodes,
prefer `retain = "metrics"` with a modest `chunk_size`.
