---
title: "Reproducibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Reproducibility guarantees

bayesim is designed so that simulation results are reproducible across
sequential, parallel, interrupted-and-resumed, and re-run executions, given a
fixed seed. Full result objects also record wall-clock timings, and those
change between runs, so the guarantee covers scientific outputs and canonical
task outcomes, not timing fields.

## What is guaranteed

- **Same seed + same package/backend versions + same platform**: scientific
  outputs and canonical task outcomes match across sequential, parallel,
  re-run, and interrupted-and-resumed execution once volatile timing fields
  are excluded. Timing fields record wall-clock time, so they differ between
  runs.
- **Across platforms** (or when the Stan/brms/cmdstanr version changes),
  results are **statistically equivalent** but may differ in the least
  significant bits due to floating-point ordering and backend differences.

Bit-identical reproducibility of the full result object is not promised on
any platform: the object records wall-clock timings, and those change between
runs.

## How determinism is achieved

### Per-task RNG streams

At the start of a run, bayesim derives one L'Ecuyer-CMRG RNG stream per task
from the simulation seed. Each task restores its stream before its data
generator and fitter run, so the RNG state a task sees depends only on its
position in the grid -- not on execution order, parallelism, or which other
tasks have completed.

Stochastic metrics receive deterministic sub-seeds derived from the task seed
and metric name. Adding or reordering metrics therefore does not change the
random draws used by another metric. `run_simulation()` also restores the
caller's RNG kind and state when it returns.

```{r}
library(bayesim)

config <- simulation_config(
  data_grid = data.frame(n = c(50, 100)),
  fit_grid = data.frame(model = "m"),
  data_generator = function(data_spec, task_ctx) {
    # consume the AMBIENT RNG state (restored by the worker before this call);
    # do NOT call set.seed() or withr::with_seed() here
    n <- data_spec$n
    x <- stats::rnorm(n)
    y <- x + stats::rnorm(n)
    list(
      train = data.frame(y = y, x = x), test = NULL, response = "y",
      true_params = c(slope = 1), vars_of_interest = "slope"
    )
  },
  fitter = LinearRegressionFitter(n_draws = 200L),
  metrics = list(posterior_summary_metric()),
  n_replicates = 2L,
  seed = 42L
)
```

### Determinism across executors

Because the RNG is per-task and stream-derived, sequential runs and mirai
daemon runs produce the same scientific outputs and canonical task outcomes.
Volatile timing fields are the only difference, so the example drops them
before comparing:

```{r}
# Sequential
seq_result <- run_simulation(
  config, resume = "never", progress = FALSE, verbose = FALSE
)

# Parallel via mirai
mirai::daemons(2)
par_result <- run_simulation(
  config, resume = "never", progress = FALSE, verbose = FALSE
)
mirai::daemons(0)

# Timing columns record wall-clock time, so drop them before comparing.
drop_timing <- function(df) df[!grepl("^timing_", names(df))]
identical(
  drop_timing(seq_result$summary),
  drop_timing(par_result$summary)
) # TRUE
```

### Determinism under resume

When a run is interrupted and resumed, the task grid (including each task's
RNG stream) is recomputed deterministically from the seed, and only terminal
task statuses are copied from the checkpoint. Resuming therefore completes
the remaining tasks with the exact RNG streams they would have had in a full
run. The resumed study then matches a single uninterrupted run once volatile
timing fields are excluded.

## The config fingerprint

Each simulation configuration is hashed into a stable fingerprint that is
written to the checkpoint manifest. On resume, bayesim verifies the
fingerprint matches, so a resumed run cannot silently use a different config
than the one that wrote the checkpoint.

## Generators and determinism

The factory generators (`fixed_truth_generator()`,
`prior_predictive_generator()`, `ifs_generator()`) all consume the ambient
RNG state. IFS and prior-predictive generators additionally select their
truth-draw parameter vector by a **deterministic** index derived from
`task_ctx$rep_idx` -- never by random sampling of draw indices -- so SBC
ranks are well-defined and resume is reproducible.

## What can break reproducibility

- Changing the seed.
- Changing the data generator's RNG consumption (e.g. adding an extra `rnorm`
  call) -- this shifts all downstream draws for that task.
- Renaming a stochastic metric, because its name is part of its deterministic
  metric-specific seed.
- Reordering the data/fit grid (tasks are identified by grid position).
- Upgrading Stan/brms/cmdstanr -- results stay statistically equivalent but
  not bit-identical.
