---
title: "Reproducibility in bayesim"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducibility in bayesim}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>")
```

# Introduction

Reproducibility is a cornerstone of rigorous scientific research. In simulation studies, reproducibility means that running the same simulation with the same inputs should produce identical results---every time, on any machine, regardless of interruptions or execution mode.

bayesim is designed from the ground up with deterministic reproducibility as a core guarantee. This vignette explains:

- **Why** reproducibility matters in Bayesian simulation studies
- **How** bayesim ensures reproducibility through careful RNG management
- **What** is and isn't guaranteed to be deterministic
- **Best practices** for ensuring your simulations remain reproducible

## Why Reproducibility Matters

Simulation studies are experiments, and like any experiment, they must be reproducible to be scientifically valid:

1. **Verification**: Others (and future you) must be able to verify your results
2. **Debugging**: Deterministic behavior makes it possible to reproduce and fix bugs
3. **Collaboration**: Team members can share and build upon each other's work
4. **Publication**: Reviewers and readers can replicate your findings
5. **Longevity**: Results remain valid even as hardware and software evolve

Without reproducibility, simulation results are essentially unverifiable claims.

# How bayesim Ensures Reproducibility

bayesim achieves reproducibility through four key mechanisms:

## 1. L'Ecuyer-CMRG RNG for Parallel-Safe Streams

bayesim uses the **L'Ecuyer-CMRG** (Combined Multiple Recursive Generator) random number generator instead of R's default Mersenne-Twister. This choice is deliberate:

- **Multiple independent streams**: L'Ecuyer-CMRG can generate multiple independent RNG streams from a single seed
- **Reproducible parallel execution**: Each task gets its own stream, ensuring results don't depend on execution order
- **Long period**: The generator has an extremely long period, preventing overlap between streams

```{r rng-setup}
library(bayesim)

# bayesim automatically sets L'Ecuyer-CMRG when you run a simulation
# You can also set it manually for custom code:
setup_global_rng(42)

# Check the RNG kind
RNGkind()
```

## 2. Pre-computed RNG Streams per Task

Before any tasks execute, bayesim pre-computes an independent RNG stream for each task:

```{r task-streams}
# Create 5 independent RNG streams from seed 42
# Note: create_task_rng_streams() is an internal API used here for demonstration
streams <- bayesim:::create_task_rng_streams(42, 5)

# Each stream is a complete RNG state
str(streams[[1]])

# These streams are independent---advancing one doesn't affect others
# This is crucial for reproducibility regardless of execution order
```

Each task receives its own stream via the `rng_seed` column in the task grid. A scalar task seed is also derived from that stream and passed into the data generator and fitter context. This means:

- Task execution order doesn't matter---each task uses its pre-assigned stream
- Parallel execution produces identical results to sequential execution
- Resuming from a checkpoint produces identical results to an uninterrupted run

## 3. Deterministic Task Ordering

Tasks are always processed in a deterministic, lexicographic order based on their task IDs:

```{r task-order}
# Task IDs follow the pattern: d..._f..._r...
# - d...: zero-padded data grid index
# - f...: zero-padded fit grid index
# - r...: zero-padded replicate number
# Padding width is chosen from the study size so ordering stays stable.

# Example: With 2 data configs, 2 fit configs, and 3 replicates:
# d001_f001_r00001, d001_f001_r00002, d001_f001_r00003,
# d001_f002_r00001, d001_f002_r00002, d001_f002_r00003,
# d002_f001_r00001, d002_f001_r00002, d002_f001_r00003,
# d002_f002_r00001, d002_f002_r00002, d002_f002_r00003
```

This deterministic ordering ensures that:

- Tasks are always processed in the same sequence
- Checkpoint resume always continues from the same point
- Results are assembled in a predictable order

## 4. Configuration Fingerprinting

Every simulation configuration gets a unique cryptographic fingerprint:

```{r fingerprint}
# Create a simple configuration
config <- simulation_config(
  data_grid = data.frame(n = c(100, 200)),
  fit_grid = data.frame(model = "linear"),
  data_generator = function(data_spec, seed, task_ctx) {
    # Note: seed is a scalar task seed.
    # The simulation engine also restores the task RNG stream before each call.
    n <- data_spec$n
    x <- rnorm(n)
    y <- 1 + 2 * x + rnorm(n)
    list(
      train = data.frame(y = y, x = x),
      test = NULL,
      response = "y",
      true_params = c(intercept = 1, slope = 2),
      vars_of_interest = c("intercept", "slope"),
      references = c(intercept = 0, slope = 0),
      meta = list()
    )
  },
  fitter = MockFitter(),
  metrics = list(rmse_metric(), bias_metric()),
  n_replicates = 3L,
  seed = 42L
)

# Compute the configuration fingerprint
fingerprint <- compute_config_fingerprint(config)
print(fingerprint)
```

The fingerprint is a SHA256 hash of the normalized configuration, excluding only runtime-specific fields (`result_path`, `checkpoint_every`). This fingerprint:

- Uniquely identifies a simulation configuration
- Enables validation during checkpoint resume
- Allows detection of configuration changes

# Reproducibility Across Execution Modes

## Sequential vs Parallel Execution

Supported sequential and parallel executions use the same pre-assigned task streams, so results stay aligned when the R, package, and model-backend environment matches:

```{r parallel-guarantee, eval=FALSE}
# Sequential execution
config <- simulation_config(..., seed = 42L)
result_seq <- run_simulation(config)

# Parallel execution
future::plan(future::multisession, workers = 4)
result_par <- run_simulation(config)

# These will produce identical results:
identical(result_seq$summary, result_par$summary)
```

This is possible because each task has its own pre-computed RNG stream. Tasks can execute in any order without affecting their assigned randomness.

## Resume from Checkpoint

Checkpoint/resume preserves the same task streams and completed-task ledger:

```{r checkpoint-resume, eval=FALSE}
# First run (interrupted)
config <- simulation_config(
  ...,
  seed = 42L,
  result_path = "my_simulation",
  checkpoint_every = 10L
)
result1 <- run_simulation(config)
# Suppose this gets interrupted after 50 tasks

# Resume from checkpoint
result2 <- run_simulation(config, resume = "must")

# result2 matches the uninterrupted run under the same software/backend stack
```

The resume mechanism:

1. Loads the checkpoint ledger (task grid with statuses)
2. Validates the configuration fingerprint matches
3. Skips already-completed tasks
4. Continues with pending tasks using their pre-assigned RNG streams
5. Reassembles the same final results under the same software/backend stack

## Different Machines and Operating Systems

Simulations are expected to be reproducible across different machines and operating systems **as long as**:

- The R version matches (see below for details)
- The bayesim version matches
- All package dependencies match
- The configuration and seed are identical

This makes it possible to:

- Start a simulation on your laptop and resume on a cluster
- Share results with collaborators who use different operating systems
- Archive simulations for long-term reproducibility

# What IS Deterministic

bayesim guarantees the following are reproducible within a matched software/backend environment:

## Same Seed → Same Results

Given identical configuration and seed, you get identical results:

```{r same-seed}
# Create a simple simulation configuration
data_gen <- function(data_spec, seed, task_ctx) {
  # Note: seed is a scalar task seed.
  # The simulation engine also restores the task RNG stream before each call.
  n <- data_spec$n
  x <- rnorm(n)
  y <- 2 + 3 * x + rnorm(n, sd = 0.5)
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(intercept = 2, slope = 3, sigma = 0.5),
    vars_of_interest = c("intercept", "slope"),
    references = c(intercept = 0, slope = 0),
    meta = list()
  )
}

config <- simulation_config(
  data_grid = data.frame(n = 50, sigma = 0.5),
  fit_grid = data.frame(model = "test"),
  data_generator = data_gen,
  fitter = MockFitter(),
  metrics = list(rmse_metric(), bias_metric()),
  n_replicates = 5L,
  seed = 123L
)

# Run twice with same seed
result_a <- run_simulation(config, progress = FALSE)
result_b <- run_simulation(config, progress = FALSE)

# Results are identical
all.equal(result_a$summary$rmse__value, result_b$summary$rmse__value)
all.equal(result_a$summary$bias__value, result_b$summary$bias__value)
```

## Resume Produces Same Final Results

Resuming from a checkpoint produces the same final results as an uninterrupted run:

```{r resume-determinism, eval=FALSE}
# This is demonstrated conceptually - in practice you would need
# to interrupt and resume an actual run

# Uninterrupted run
config <- simulation_config(
  ...,
  seed = 42L,
  result_path = "uninterrupted_run",
  checkpoint_every = 1000L  # Won't trigger during run
)
result_full <- run_simulation(config)

# Interrupted and resumed run
config2 <- simulation_config(
  ...,
  seed = 42L,
  result_path = "resumed_run",
  checkpoint_every = 10L
)
result_resumed <- run_simulation(config2, resume = "must")

# Final results are identical
identical(result_full$summary, result_resumed$summary)
```

## Task Execution Order

Tasks are planned in a deterministic order:

1. Sorted lexicographically by task ID
2. Each task uses its pre-assigned RNG stream
3. Results are assembled in task ID order

# What is NOT Guaranteed

While bayesim strives for complete reproducibility, some aspects are inherently non-deterministic:

## Exact Timing Values

Task execution times will vary between runs:

```{r timing-varies}
# Run the same simulation twice
config <- simulation_config(
  data_grid = data.frame(n = 100),
  fit_grid = data.frame(model = "test"),
  data_generator = data_gen,
  fitter = MockFitter(),
  metrics = list(rmse_metric()),
  n_replicates = 3L,
  seed = 42L
)

result1 <- run_simulation(config, progress = FALSE)
result2 <- run_simulation(config, progress = FALSE)

# Timing will differ
timing1 <- result1$timing$total
timing2 <- result2$timing$total
cat("Run 1:", timing1, "seconds\n")
cat("Run 2:", timing2, "seconds\n")
```

This is expected and doesn't affect the scientific validity of results.

## Order of Warning Messages

If multiple tasks produce warnings, the order in which they appear may vary:

- In sequential execution, warnings appear in task order
- In parallel execution, warning order depends on task completion order
- The warnings themselves are deterministic, just their order may vary

## Results Across R Versions

**Critical**: Results may differ across different versions of R, even with the same seed. This is because:

- R's random number generation may change between versions
- Underlying algorithms (e.g., for linear algebra) may be updated
- Floating-point behavior can vary

**Best practice**: Always record the R version used for your simulations.

## Results Across Package Versions

Similarly, results may differ across versions of bayesim or its dependencies:

- Algorithm improvements may change results
- Bug fixes may alter behavior
- New features may change defaults

**Best practice**: Use package version pinning (e.g., renv, packrat) for critical simulations.

# Best Practices

## Always Set a Seed

Always explicitly set a seed in your simulation configuration:

```{r always-seed, eval=FALSE}
# Good: explicit seed
config <- simulation_config(
  ...,
  seed = 42L  # Use any integer
)

# Without a seed, your results won't be reproducible
```

Choose a seed arbitrarily (many researchers use their birthday, favorite number, or random.org). The specific value doesn't matter---only that it's documented.

## Version Control Your Configuration

Store your simulation configuration in version control:

```{r version-control, eval=FALSE}
# In your R script or R Markdown document
simulation_config(
  data_grid = read.csv("data_grid.csv"),
  fit_grid = read.csv("fit_grid.csv"),
  data_generator = my_data_gen,  # Defined in the same file
  fitter = my_fitter,            # Defined in the same file
  metrics = list(rmse_metric(), bias_metric()),
  n_replicates = 1000L,
  seed = 42L
)
```

This ensures that:

- The exact configuration is preserved
- Changes are tracked over time
- Others can reproduce your setup

## Use Checkpointing for Long Runs

For long-running simulations, always enable checkpointing:

```{r checkpointing, eval=FALSE}
config <- simulation_config(
  ...,
  result_path = "path/to/results",  # Enable checkpointing
  checkpoint_every = 50L             # Save every 50 tasks
)

# Run with resume capability
result <- run_simulation(config, resume = "auto")
```

This protects against:

- System crashes
- Power failures
- Time limits on clusters
- Accidental interruption

## Record R and Package Versions

Document your computational environment:

```{r record-versions}
# Record R version
R.version.string

# Record package versions
sessionInfo()
```

For publication, include a section like:

> "Simulations were conducted using R version 4.3.1 with bayesim version 0.1.0. The computational environment is available at [repository/link]."

## Environment Management

For stronger reproducibility guarantees, use environment management tools to ensure consistent package versions and R environments across machines and sessions.

### renv: Project-Local Dependencies

**renv** creates a project-specific package library with a lockfile capturing exact package versions:

```{r renv-examples, eval=FALSE}
# Initialize renv in your project
renv::init()

# Snapshot current package versions to lockfile
renv::snapshot()

# Restore packages from lockfile (on another machine or session)
renv::restore()
```

The `renv.lock` file ensures collaborators and future-you install identical package versions. This is essential for long-term reproducibility of simulation results.

### rv and Containers: Full Environment Capture

For capturing the complete R environment (R version + packages + system libraries):

- **rv** is a newer tool for managing R environments with version-controlled snapshots
- **Docker/containers** provide the strongest guarantees by capturing the entire OS, R version, and all dependencies

```{r rv-example, eval=FALSE}
# Example: Using rv to manage environments
rv::init()
rv::snapshot()
rv::restore()
```

Container-based approaches (Docker, Singularity) are recommended for published research where reproducibility must survive years of software evolution.

# Verifying Reproducibility

Here's a complete example demonstrating reproducibility:

```{r verify-reproducibility}
library(bayesim)

# Define a data generator
my_data_generator <- function(data_spec, seed, task_ctx) {
  # Note: seed is a scalar task seed.
  # The simulation engine also restores the task RNG stream before each call.
  n <- data_spec$n
  
  # Generate synthetic data with known parameters
  x <- rnorm(n)
  y <- data_spec$intercept + data_spec$slope * x + rnorm(n, sd = data_spec$sigma)
  
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(
      intercept = data_spec$intercept,
      slope = data_spec$slope,
      sigma = data_spec$sigma
    ),
    vars_of_interest = c("intercept", "slope"),
    references = c(intercept = 0, slope = 0),
    meta = list()
  )
}

# Create configuration
config <- simulation_config(
  data_grid = data.frame(
    n = c(50, 100),
    intercept = 1,
    slope = 2,
    sigma = 0.5
  ),
  fit_grid = data.frame(
    model = "linear"
  ),
  data_generator = my_data_generator,
  fitter = MockFitter(),
  metrics = list(
    rmse_metric(),
    bias_metric()
  ),
  n_replicates = 10L,
  seed = 42L
)

# Run simulation twice with same seed
cat("Running simulation for the first time...\n")
result1 <- run_simulation(config, progress = FALSE)

cat("Running simulation for the second time...\n")
result2 <- run_simulation(config, progress = FALSE)

# Verify results are identical
cat("\n=== Reproducibility Check ===\n")

# Check summary structure
structures_match <- identical(names(result1$summary), names(result2$summary))
cat("Summary structures match:", structures_match, "\n")

# Check metric values
rmse_match <- all.equal(result1$summary$rmse__value, result2$summary$rmse__value)
bias_match <- all.equal(result1$summary$bias__value, result2$summary$bias__value)
cat("RMSE values match:", isTRUE(rmse_match), "\n")
cat("Bias values match:", isTRUE(bias_match), "\n")

# Check task statuses
status_match <- identical(result1$task_grid$status, result2$task_grid$status)
cat("Task statuses match:", status_match, "\n")

# Overall reproducibility
is_reproducible <- structures_match && isTRUE(rmse_match) && 
                   isTRUE(bias_match) && status_match
cat("\nOverall reproducibility:", is_reproducible, "\n")
```

## Testing with Different Seeds

To verify that different seeds produce different (but still reproducible) results:

```{r different-seeds}
# Same configuration, different seeds
config1 <- simulation_config(
  data_grid = data.frame(n = 100, intercept = 1, slope = 2, sigma = 0.5),
  fit_grid = data.frame(model = "linear"),
  data_generator = my_data_generator,
  fitter = MockFitter(),
  metrics = list(rmse_metric()),
  n_replicates = 5L,
  seed = 42L
)

config2 <- simulation_config(
  data_grid = data.frame(n = 100, intercept = 1, slope = 2, sigma = 0.5),
  fit_grid = data.frame(model = "linear"),
  data_generator = my_data_generator,
  fitter = MockFitter(),
  metrics = list(rmse_metric()),
  n_replicates = 5L,
  seed = 123L  # Different seed
)

result1 <- run_simulation(config1, progress = FALSE)
result2 <- run_simulation(config2, progress = FALSE)

# Results should be different (different random data)
cat("Same seed produces same results: ",
    identical(result1$summary$rmse__value, result2$summary$rmse__value), "\n")
cat("Different seeds produce different results: ",
    !identical(result1$summary$rmse__value, result2$summary$rmse__value), "\n")
```

# Summary

bayesim provides strong determinism guarantees for simulation studies:

| Aspect | Guarantee |
|--------|-----------|
| Same seed → same results | ✅ Guaranteed |
| Sequential vs parallel | ✅ Identical in supported execution modes |
| Resume from checkpoint | ✅ Identical to uninterrupted run |
| Task execution order | ✅ Deterministic |
| Cross-platform (same R version) | ⚠️ Expected under controlled environment |
| Timing values | ❌ Not guaranteed |
| Warning message order | ❌ Not guaranteed |
| Across R versions | ❌ Not guaranteed |
| Across package versions | ❌ Not guaranteed |

By following the best practices outlined in this vignette---always setting a seed, version controlling your configuration, using checkpointing, recording your computational environment, and using environment management tools---you can ensure that your simulation studies remain reproducible and scientifically valid.

For more information, see:

- `vignette("getting-started")` for basic bayesim usage
- `vignette("simulation-study")` for a complete example
- `vignette("memory-management")` for handling large simulations
