---
title: "Reproducibility"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reproducibility}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Reproducibility guarantees

bayesim is designed so that simulation results are reproducible across
sequential, parallel, interrupted-and-resumed, and re-run executions, given a
fixed seed.

## What is guaranteed

- **Same seed + same package/backend versions + same platform** produces
  **identical** results (bit-for-bit), whether the run is sequential, uses
  mirai daemons, or is interrupted and resumed.
- **Across platforms** (or when the Stan/brms/cmdstanr version changes),
  results are **statistically equivalent** but may differ in the least
  significant bits due to floating-point ordering and backend differences.

This is the honest, qualified statement; bit-identical reproducibility cannot
be promised across Stan versions or operating systems.

## How determinism is achieved

### Per-task RNG streams

At the start of a run, bayesim derives one L'Ecuyer-CMRG RNG stream per task
from the simulation seed. Each task restores its stream before its data
generator and fitter run, so the RNG state a task sees depends only on its
position in the grid -- not on execution order, parallelism, or which other
tasks have completed.

```{r, eval=FALSE}
library(bayesim)

config <- simulation_config(
  data_grid = data.frame(n = c(50, 100)),
  fit_grid = data.frame(model = "m"),
  data_generator = function(data_spec, task_ctx) {
    # consume the AMBIENT RNG state (restored by the worker before this call);
    # do NOT call set.seed() or withr::with_seed() here
    n <- data_spec$n
    x <- stats::rnorm(n)
    y <- x + stats::rnorm(n)
    list(
      train = data.frame(y = y, x = x), test = NULL, response = "y",
      true_params = c(slope = 1), vars_of_interest = "slope"
    )
  },
  fitter = MockFitter(),
  metrics = list(rmse_metric()),
  n_replicates = 2L,
  seed = 42L
)
```

### Determinism across executors

Because the RNG is per-task and stream-derived, a study produces identical
summaries whether run sequentially or via mirai daemons:

```{r, eval=FALSE}
# Sequential
seq_result <- run_simulation(config, resume = "never", progress = FALSE)

# Parallel via mirai
mirai::daemons(2)
par_result <- run_simulation(config, resume = "never", progress = FALSE)
mirai::daemons(0)

identical(seq_result$summary, par_result$summary) # TRUE (ignoring timing columns)
```

### Determinism under resume

When a run is interrupted and resumed, the task grid (including each task's
RNG stream) is recomputed deterministically from the seed, and only terminal
task statuses are copied from the checkpoint. Resuming therefore completes
the remaining tasks with the exact RNG streams they would have had in a full
run, producing identical results.

## The config fingerprint

Each simulation configuration is hashed into a stable fingerprint that is
written to the checkpoint manifest. On resume, bayesim verifies the
fingerprint matches, so a resumed run cannot silently use a different config
than the one that wrote the checkpoint.

## Generators and determinism

The factory generators (`fixed_truth_generator()`,
`prior_predictive_generator()`, `ifs_generator()`) all consume the ambient
RNG state. IFS and prior-predictive generators additionally select their
truth-draw parameter vector by a **deterministic** index derived from
`task_ctx$rep_idx` -- never by random sampling of draw indices -- so SBC
ranks are well-defined and resume is reproducible.

## What can break reproducibility

- Changing the seed.
- Changing the data generator's RNG consumption (e.g. adding an extra `rnorm`
  call) -- this shifts all downstream draws for that task.
- Reordering the data/fit grid (tasks are identified by grid position).
- Upgrading Stan/brms/cmdstanr -- results stay statistically equivalent but
  not bit-identical.
