---
title: "SBC and calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SBC and calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{bayesim}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(bayesim)
```

# Introduction

Simulation-Based Calibration (SBC) is the standard self-consistency check for
Bayesian inference engines. The idea, due to Talts et al. (2018), is sharp:

> If we draw a parameter `theta` from the model prior, simulate data `y` from
> the likelihood given `theta`, and then fit the model to `y`, the resulting
> posterior should treat `theta` as a single uniform draw.

Equivalently, the **posterior rank** of the true `theta` among the posterior
draws should be uniformly distributed on `{0, 1, ..., S}`. If it is not, the
inference engine is miscalibrated: the posterior is either too confident
(ranks pile up at the ends) or biased (ranks drift to one side).

This matters because a Bayesian posterior is only useful insofar as its
uncertainty statements are honest. **Calibration** is the property that
empirical coverage matches the nominal rate: a 95% credible interval should
contain the truth about 95% of the time. SBC is the simulation analogue of
this guarantee, applied to the *whole* posterior rather than to a single
interval, and it is sensitive to biases that single-interval coverage can
miss.

This vignette runs end-to-end with `LinearRegressionFitter()` — exact
conjugate Normal-Inverse-Gamma Bayesian linear regression (real posteriors,
no Stan, milliseconds per fit). For Stan/brms models, swap the fitter for
`BrmsFitter()` or `CmdStanFitter()` and use `prior_predictive_generator()` /
`ifs_generator()`; the SBC workflow below is identical.

# Running SBC

The SBC recipe needs three ingredients:

1. A **data generator** that draws `theta` from the model prior and simulates
   `y ~ p(y | theta)`.
2. A **fitter** whose calibration we want to check.
3. The **rank metric** (`rank_metric()`), which counts how many posterior
   draws fall below the true `theta`, giving one rank per task per parameter.

Below, the generator draws `(Intercept, slope, sigma)` from the same weak
prior family the `LinearRegressionFitter` assumes (broad Normal coefficients,
inverse-gamma residual scale), then simulates Gaussian linear data. Because
the fitter is the exact conjugate NIG updater, SBC should pass by
construction — which is exactly what makes this a good teaching example: you
see a *correct* rank distribution and learn what passing looks like.

```{r}
sbc_generator <- function(data_spec, task_ctx) {
  n <- data_spec$n
  # Draw theta from the model prior (conjugate-friendly, weak/diffuse).
  intercept <- stats::rnorm(1, mean = 0, sd = 2)
  slope     <- stats::rnorm(1, mean = 0, sd = 2)
  sigma     <- sqrt(1 / stats::rgamma(1, shape = 2, rate = 1))
  # Simulate y from the likelihood given theta.
  x <- stats::rnorm(n)
  y <- intercept + slope * x + stats::rnorm(n, sd = sigma)
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(Intercept = intercept, x = slope, sigma = sigma),
    vars_of_interest = c("Intercept", "x", "sigma")
  )
}
```

The data generator consumes the **ambient** RNG state — bayesim restores a
per-task L'Ecuyer stream before each call, so do not call `set.seed()` inside
(see `vignette("reproducibility")`).

Now configure and run the study. We use 150 replicates and 1000 posterior
draws per fit, with `rank_metric(thin = "auto")` so the rank counts are
adjusted for autocorrelation in the draws (here the NIG draws are i.i.d., so
no thinning is needed, but the option is there for MCMC-based fitters).

```{r}
config <- simulation_config(
  data_grid = data.frame(n = 40L),
  fit_grid = data.frame(model = "lm"),
  data_generator = sbc_generator,
  fitter = LinearRegressionFitter(n_draws = 1000L),
  metrics = list(
    rank_metric(thin = "auto"),
    posterior_summary_metric()
  ),
  n_replicates = 150L,
  seed = 7L
)

result <- run_simulation(config, progress = FALSE)
```

Each task records one rank per parameter (here `Intercept`, `x`, `sigma`),
plus the posterior summaries needed for the coverage check below.

# Checking rank uniformity

`sbc_ranks()` collects the per-task ranks into a long tibble; `plot_rank_ecdf()`
plots the empirical CDF of the normalized ranks against the uniform CDF (the
diagonal), with a pointwise uniformity band following Talts et al. (2018).

```{r}
ranks <- sbc_ranks(result)
head(ranks)
```

```{r sbc-ecdf}
plot_rank_ecdf(ranks, alpha = 0.95)
```

**How to read the band.** The grey ribbon is the pointwise 95% coverage
interval for the ECDF of a truly uniform sample of this size, derived from
the beta distribution of order statistics: at the `i`-th of `n` ordered
points the lower bound is `qbeta(0.025, i, n + 1 - i)` and the upper bound is
`qbeta(0.975, i, n + 1 - i)`. Under correct calibration, the ECDF (black
step) should stay inside the ribbon everywhere. The dashed red line is the
theoretical uniform CDF.

Because `LinearRegressionFitter` is the *exact* conjugate updater, the ECDF
hugs the diagonal and stays inside the band — this is what a calibrated
inference engine looks like. With an approximate engine (e.g. a variational
fitter, or a mis-specified MCMC sampler), the ECDF would systematically
depart from the diagonal.

# Interpreting failures

When the ECDF *systematically* leaves the band, the shape of the departure
tells you what is wrong:

- **S-shape (ECDF above the diagonal on the left, below on the right):** the
  posterior is *over-confident* — the draws are too tightly clustered around
  their centre, so the true `theta` too often falls outside the bulk. This is
  the classic signature of underestimated posterior uncertainty
  (undercoverage).
- **Reflected S-shape (below on the left, above on the right):** the posterior
  is *under-confident* — wider than the likelihood justifies (overcoverage).
- **One-sided drift:** the posterior is *biased* — it systematically
  over- or under-estimates `theta`.

Random excursions outside the band at one or two points, with the ECDF
returning inside, are expected by chance at the 95% level and are not
evidence of miscalibration. Use a stricter band (`alpha = 0.99`) to confirm
suspected systematic departures, and increase `n_replicates` to sharpen the
test — SBC is a Monte Carlo procedure with its own sampling noise.

# Coverage as a complementary check

SBC checks the *whole* posterior at once. A complementary, more familiar
check is **interval coverage**: does the nominal 95% credible interval contain
the truth about 95% of the time? `plot_coverage()` shows this per estimand
with MCSE error bars; `performance_measures()` reports the same quantity in a
table alongside bias, empirical SE, MSE, and average model SE.

```{r coverage}
plot_coverage(result)
```

```{r}
pm <- performance_measures(result, estimand = "x")
pm
```

The `coverage` row reports the empirical coverage of the 95% posterior
interval for the `x` coefficient, with its Monte Carlo standard error. For
the conjugate fitter this should sit close to the nominal 0.95 (within a
couple of MCSEs).

**Coverage vs SBC.** Coverage and SBC agree when the model is well-behaved,
but they probe different things. Coverage is a single-number summary of one
interval and can miss miscalibration that SBC catches: two posteriors can
have identical 95% coverage while one has correct tail behaviour and the
other is systematically over-dispersed in the tails and under-dispersed in
the middle. SBC, by contrast, checks the full posterior shape through rank
uniformity. In practice, run both: SBC as the global self-consistency test,
and `performance_measures()` coverage as the interval-level summary a methods
paper reports.

# Further reading

- Talts et al. (2018), *Validating Bayesian Inference Algorithms with
  Simulation-Based Calibration*, arXiv:1804.06788.
- `vignette("design-of-simulation-studies")` for the Morris, White & Crowther
  (2019) framework for designing and reporting simulation studies.
- `vignette("reproducibility")` for the determinism guarantees behind SBC
  rank reproducibility.
