---
title: "SBC and calibration"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SBC and calibration}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{bayesim}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 4
)
```

```{r setup}
library(bayesim)
```

# Introduction

Simulation-based calibration (SBC) checks Bayesian inference against data
generated from the model (Talts et al., 2018):

> If we draw a parameter `theta` from the model prior, simulate data `y` from
> the likelihood given `theta`, and then fit the model to `y`, the resulting
> rank of `theta` among independent posterior draws should be uniform
> across repeated simulations.

Equivalently, the **posterior rank** of the true `theta` among the posterior
draws should be uniformly distributed on `{0, 1, ..., S}`. Departures can
indicate errors in inference or in the simulation setup.

The coverage guarantee averages over parameters drawn from the prior and
data drawn from the likelihood. It does not guarantee nominal coverage at
each fixed parameter value. Rank checks can detect errors that a check of
one credible interval misses, but passing SBC does not establish that an
inference algorithm is correct.

This vignette uses `LinearRegressionFitter()` for exact
conjugate Normal-Inverse-Gamma Bayesian linear regression without Stan.
For Stan/brms models, swap the fitter for
`BrmsFitter()` or `CmdStanFitter()` and use `prior_predictive_generator()` /
`ifs_generator()`; the SBC workflow below is identical. One caveat:
`prior_predictive_generator()` draws theta from the model prior (valid SBC by
construction when the fitting prior matches), but `ifs_generator()` draws
theta from a preconditioning posterior, so its ranks are only uniform if the
fitting prior is set to match that preconditioning distribution. An
unmatched fitting prior can produce nonuniform ranks without a sampler
error (see `?ifs_generator`).

# Running SBC

The SBC recipe needs three ingredients:

1. A **data generator** that draws `theta` from the model prior and simulates
   `y ~ p(y | theta)`.
2. A **fitter** whose calibration we want to check.
3. The **rank metric** (`rank_metric()`), which counts how many posterior
   draws fall below the true `theta`, giving one rank per task per parameter.

Below, the generator draws `(Intercept, slope, sigma)` from exactly the same
Normal-Inverse-Gamma prior passed to `LinearRegressionFitter()`, then simulates
Gaussian linear data. In an NIG prior the coefficient distribution is
conditional on the residual variance:
`beta | sigma^2 ~ Normal(0, sigma^2 Lambda^-1)`. Matching that conditional
prior is essential; a prior mismatch is itself a valid SBC failure mode.
Because the fitter is the exact conjugate updater for this same prior, SBC
should pass by construction.

```{r}
sbc_generator <- function(data_spec, task_ctx) {
  n <- data_spec$n
  # Exact NIG prior used by the fitter below:
  # sigma^2 ~ Inv-Gamma(2, 1)
  # beta | sigma^2 ~ Normal(0, sigma^2 / 0.25)
  sigma     <- sqrt(1 / stats::rgamma(1, shape = 2, rate = 1))
  intercept <- stats::rnorm(1, mean = 0, sd = sigma / sqrt(0.25))
  slope     <- stats::rnorm(1, mean = 0, sd = sigma / sqrt(0.25))
  # Simulate y from the likelihood given theta.
  x <- stats::rnorm(n)
  y <- intercept + slope * x + stats::rnorm(n, sd = sigma)
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(Intercept = intercept, x = slope, sigma = sigma),
    vars_of_interest = c("Intercept", "x", "sigma")
  )
}
```

The data generator uses the current RNG state. bayesim restores a
per-task L'Ecuyer stream before each call, so do not call `set.seed()` inside
(see `vignette("reproducibility")`).

Now configure and run the study. We use 150 replicates and 1000 posterior
draws per fit, with `rank_metric(thin = "auto")` so the rank counts are
adjusted for autocorrelation in the draws (here the NIG draws are i.i.d., so
no thinning is needed, but the option is there for MCMC-based fitters).

```{r}
config <- simulation_config(
  data_grid = data.frame(n = 40L),
  fit_grid = data.frame(model = "lm"),
  data_generator = sbc_generator,
  fitter = LinearRegressionFitter(
    n_draws = 1000L,
    prior_mean = 0,
    prior_precision = 0.25,
    a0 = 2,
    b0 = 1
  ),
  metrics = list(
    rank_metric(thin = "auto"),
    posterior_summary_metric()
  ),
  n_replicates = 150L,
  seed = 7L
)

result <- run_simulation(config, progress = FALSE)
```

Each task records one rank per parameter (here `Intercept`, `x`, `sigma`),
plus the posterior summaries needed for the coverage check below.

# Checking rank uniformity

`sbc_ranks()` collects the per-task ranks into a long tibble; `plot_rank_ecdf()`
plots the empirical CDF of the normalized ranks against the uniform CDF (the
diagonal), with the simultaneous uniformity band of Säilynoja, Bürkner, and
Vehtari (2022).

```{r}
ranks <- sbc_ranks(result)
head(ranks)
```

```{r sbc-ecdf}
plot_rank_ecdf(ranks, alpha = 0.95)
```

The red diagonal is the uniform CDF. The grey ribbon is a 95% simultaneous
band: for independent uniform ranks on a common support, the whole ECDF
stays inside it in about 95% of repetitions. This level applies to each
panel separately. A crossing can occur by chance even with exact inference.

If condition cells share truth draws, their ranks are dependent. Pooling
them can make the band too narrow. Use `by` to separate the conditions, for
example `plot_rank_ecdf(result, by = c("data_n", "fit_model"))` when those
columns identify the cells. The band does not correct for dependence from
shared draws ([#59](https://github.com/sims1253/bayesim/issues/59)).
It is also approximate when ranks within a panel have different supports.

# Interpreting departures

The shape can suggest a cause, but does not identify it uniquely:

| ECDF relative to the diagonal | Possible explanation |
|---|---|
| Above on the left, below on the right | Posterior uncertainty is too small. |
| Below on the left, above on the right | Posterior uncertainty is too large. |
| Mostly above | Posterior estimates tend to exceed the truth. |
| Mostly below | Posterior estimates tend to fall below the truth. |

Check the generator, model, and sampler diagnostics before attributing a
departure to the inference method. More replicates improve the ability to
detect small departures.

# Coverage as a complementary check

SBC checks the *whole* posterior at once. A complementary, more familiar
check is **interval coverage**: does the nominal 95% credible interval contain
the truth about 95% of the time? `plot_coverage()` shows this per estimand
with MCSE error bars; `performance_measures()` reports the same quantity in a
table. Because SBC draws a new truth from the prior for every replicate, this
is a **varying-truth** study. Accordingly, the table labels error summaries as
`mean_error`, `error_sd`, and `error_mse`; it does not label them as the
fixed-truth Morris measures `bias`, `emp_se`, or `mse`.

```{r coverage}
plot_coverage(result)
```

```{r}
pm <- performance_measures(result, estimand = "x")
pm
```

The `coverage` row reports the empirical coverage of the 95% posterior
interval for the `x` coefficient, with its Monte Carlo standard error. For
the conjugate fitter this should sit close to the nominal 0.95 (within a
couple of MCSEs). The `truth_mode` column is `"varying"`. Here `mean_error`
is the average replicate-level difference `estimate - truth`, and its MCSE is
computed from those replicate-level errors; `error_sd` describes the spread
of that error distribution. These are useful calibration summaries, but they
are not the empirical sampling SE or bias of an estimator at one fixed truth.

Coverage checks one interval level. Rank uniformity checks calibration over
a range of posterior quantiles for the quantities supplied to `rank_metric()`.
Neither check proves that every aspect of the joint posterior is correct.
Use a fixed-truth generator when you want the estimator-performance measures
`bias`, `emp_se`, and `mse`.

# Further reading

- Talts et al. (2018), *Validating Bayesian Inference Algorithms with
  Simulation-Based Calibration*, arXiv:1804.06788.
- Säilynoja, Bürkner, and Vehtari (2022), *Graphical test for discrete
  uniformity and its applications in goodness-of-fit evaluation*,
  *Statistics and Computing* 32(2).
- `vignette("design-of-simulation-studies")` for the Morris, White & Crowther
  (2019) framework for designing and reporting simulation studies.
- `vignette("reproducibility")` for the determinism guarantees behind SBC
  rank reproducibility.
