---
title: "Simulation Study Example: Assessing Model Performance"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Simulation Study Example: Assessing Model Performance}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bayesim)
library(dplyr)
library(ggplot2)
```

## Introduction

Simulation studies are essential for understanding how statistical models perform under controlled conditions. By systematically varying data characteristics and measuring estimation accuracy, we can identify scenarios where models excel or struggle. This vignette demonstrates how to conduct a realistic simulation study using bayesim 1.0.

The goal of this example is to assess how sample size and noise level affect the accuracy of parameter estimation in a simple linear regression model. We will generate synthetic data with known parameters, fit models to each dataset, and compute metrics that quantify estimation quality. This pattern generalizes to more complex models and study designs.

bayesim provides a structured framework for such studies. The key components are: (1) a data generator function that creates synthetic datasets, (2) a configuration object that defines the simulation grid and settings, and (3) a runner that executes the study with deterministic reproducibility. This vignette walks through each component in detail.

## Study Design

Our hypothetical study examines two factors that commonly affect model performance:

1. **Sample size (n)**: We test n = 50, 100, 200, and 500 observations. Larger samples should yield more precise estimates.
2. **Noise level (sigma)**: We test sigma = 0.5, 1, and 2. Higher noise makes estimation more challenging.

The full factorial design yields 4 × 3 = 12 data generation conditions. For each condition, we run 20 replicates with different random seeds, giving 240 total simulation tasks.

We assess performance using four metrics:

- **RMSE**: Root mean square error of predictions
- **Bias**: Mean bias of predictions  
- **Coverage**: Whether true parameters fall within 95% credible intervals
- **Posterior Mean**: Estimated parameter values for tracking estimation accuracy

## Data Generator

The data generator creates synthetic linear regression data. It must have the signature `(data_spec, seed, task_ctx)`:

```{r}
generate_linear_data <- function(data_spec, seed, task_ctx) {
  # Note: seed is a scalar task seed.
  # The simulation engine also restores the task RNG stream before each call.
  # so you typically don't need to call set.seed() here.
  # If you need additional randomization, the state is already set correctly.
  n <- data_spec$n
  intercept <- data_spec$intercept
  slope <- data_spec$slope
  sigma <- data_spec$sigma
  
  # Generate predictor and response
  x <- rnorm(n, mean = 0, sd = 1)
  y <- intercept + slope * x + rnorm(n, mean = 0, sd = sigma)
  
  # Return data bundle with required components
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(
      intercept = intercept,
      slope = slope,
      sigma = sigma
    ),
    vars_of_interest = c("intercept", "slope", "sigma"),
    references = c(intercept = 0, slope = 0, sigma = 1),
    meta = list()
  )
}
```

The data bundle must include:

- `train`: Training data as a data frame
- `test`: Optional test data (NULL here since we evaluate on training data)
- `response`: Name of the response variable
- `true_params`: Named vector of true parameter values for coverage calculations
- `vars_of_interest`: Parameters to track in metrics
- `references`: Reference values for certain metrics
- `meta`: Optional metadata

## Configuration

We create a simulation configuration that combines the data grid, fit grid, and execution settings:

```{r}
config <- simulation_config(
  data_grid = data.frame(
    n = c(50, 100, 200, 500),
    intercept = 1.0,
    slope = 2.0,
    sigma = rep(c(0.5, 1, 2), each = 4)
  ),
  fit_grid = data.frame(
    model = "linear_regression"
  ),
  data_generator = generate_linear_data,
  fitter = MockFitter(),
  metrics = list(
    rmse_metric(),
    bias_metric(),
    coverage_metric(),
    posterior_mean_metric()
  ),
  n_replicates = 20L,
  seed = 42L
)
```

Note that we use `MockFitter()` here for illustration purposes. In a real study with actual Bayesian models, you would use `BrmsFitter()` or a custom fitter that interfaces with your modeling backend. The MockFitter simulates realistic output structure without requiring brms or Stan to be installed, making this vignette self-contained and fast to run.

The `data_grid` specifies all combinations of sample size and noise level. The `fit_grid` contains a single model specification since we are comparing data conditions rather than model variants. The metrics list includes objects (not character strings) that define what to compute for each simulation replicate. Built-in metric constructors like `rmse_metric()` have default names; custom metrics require an explicit `name` argument.

## Running the Simulation

Execute the simulation with progress display:

```{r}
result <- run_simulation(config, progress = TRUE)
```

The progress bar shows completion status and elapsed time. With 240 tasks and the MockFitter, this completes quickly. Real studies with brms models may take minutes to hours depending on data size and model complexity.

## Results Analysis

### Examining the Result Object

The result object contains the simulation output:

```{r}
print(result)
```

### Summary Tibble

The summary tibble contains one row per task with flattened metric columns. Data grid columns are prefixed with `data_` and fit grid columns with `fit_`:

```{r}
head(result$summary)
```

Column names follow these patterns:

- `task_id`, `status`, `timing_total`: Basic task information
- `data_<colname>`: Columns from the data_grid (e.g., `data_n`, `data_sigma`)
- `fit_<colname>`: Columns from the fit_grid (e.g., `fit_model`)
- `<metric>__<field>`: Metric outputs (e.g., `rmse__value`, `coverage__mean`)
- `<metric>__<field>__<subname>`: Flattened named vectors (e.g., `coverage__by_param__intercept`)

### Aggregation with dplyr

Aggregate results to examine patterns across conditions:

```{r}
summary_stats <- result$summary |>
  group_by(data_n, data_sigma) |>
  summarise(
    n_tasks = n(),
    mean_rmse = mean(rmse__value, na.rm = TRUE),
    sd_rmse = sd(rmse__value, na.rm = TRUE),
    mean_bias = mean(bias__value, na.rm = TRUE),
    mean_coverage = mean(coverage__mean, na.rm = TRUE),
    .groups = "drop"
  )

print(summary_stats)
```

### Visualizations

Visualize how RMSE varies with sample size and noise level:

```{r, fig.width=7, fig.height=5}
result$summary |>
  mutate(
    data_n = factor(data_n, levels = c(50, 100, 200, 500)),
    data_sigma = factor(data_sigma, levels = c(0.5, 1, 2))
  ) |>
  ggplot(aes(x = data_n, y = rmse__value, fill = data_sigma)) +
  geom_boxplot() +
  labs(
    title = "RMSE by Sample Size and Noise Level",
    x = "Sample Size (n)",
    y = "Root Mean Square Error",
    fill = "Noise (σ)"
  ) +
  theme_minimal()
```

Examine coverage rates:

```{r, fig.width=7, fig.height=4}
coverage_summary <- result$summary |>
  group_by(data_n, data_sigma) |>
  summarise(
    coverage_rate = mean(coverage__mean, na.rm = TRUE),
    .groups = "drop"
  )

ggplot(coverage_summary, aes(x = data_n, y = coverage_rate, color = factor(data_sigma))) +
  geom_line(linewidth = 1) +
  geom_point(size = 3) +
  geom_hline(yintercept = 0.95, linetype = "dashed", alpha = 0.5) +
  scale_x_continuous(breaks = c(50, 100, 200, 500)) +
  labs(
    title = "Coverage Rate by Sample Size and Noise Level",
    subtitle = "Dashed line indicates nominal 95% coverage",
    x = "Sample Size (n)",
    y = "Coverage Rate",
    color = "Noise (σ)"
  ) +
  theme_minimal()
```

View posterior mean estimates for the slope parameter:

```{r, fig.width=7, fig.height=4}
result$summary |>
  ggplot(aes(x = factor(data_n), y = posterior_mean__slope)) +
  geom_boxplot(aes(fill = factor(data_sigma))) +
  geom_hline(yintercept = 2.0, linetype = "dashed", color = "red") +
  labs(
    title = "Posterior Mean Estimates of Slope",
    subtitle = "Red dashed line indicates true value (2.0)",
    x = "Sample Size (n)",
    y = "Estimated Slope",
    fill = "Noise (σ)"
  ) +
  theme_minimal()
```

### Accessing the Task Grid

The result also contains the task grid, which shows the full experimental design:

```{r}
head(result$task_grid)
```

This can be useful for custom analyses or for debugging.

## Checkpointing

For long-running simulations, enable checkpointing to save progress periodically:

```{r, eval=FALSE}
config_with_checkpoint <- simulation_config(
  data_grid = data.frame(n = c(100, 500, 1000)),
  fit_grid = data.frame(model = "complex_model"),
  data_generator = generate_linear_data,
  fitter = MockFitter(),  # Replace with BrmsFitter() for real models
  metrics = list(
    rmse_metric(),
    coverage_metric()
  ),
  n_replicates = 100L,
  seed = 42L,
  result_path = "my_simulation_results",
  checkpoint_every = 50L
)

# Run with resume capability
result <- run_simulation(config_with_checkpoint, resume = "auto")
```

If the simulation is interrupted, resume from the last checkpoint:

```{r, eval=FALSE}
result <- resume_simulation("my_simulation_results")
```

The checkpoint system validates that the configuration matches the saved state, preventing accidental mismatches between resumed runs and modified code. The supported checkpoint backend is `checkpoint_format = "rds"`.

## Summary

This vignette demonstrated a complete simulation study workflow:

1. **Defined a data generator** that creates synthetic linear regression data with configurable sample size, slope, intercept, and noise level.

2. **Created a simulation configuration** specifying the experimental grid, metrics, and execution parameters.

3. **Ran the simulation** with progress tracking and obtained reproducible results.

4. **Analyzed results** using dplyr aggregation and ggplot2 visualizations to understand how estimation accuracy varies with sample size and noise.

5. **Showed checkpointing** for long-running studies that need interruption tolerance.

Key takeaways for conducting your own simulation studies:

- Use `MockFitter()` for rapid prototyping and testing; switch to `BrmsFitter()` or custom fitters for real inference
- Structure data generators to return complete data bundles with `true_params` for coverage calculations
- Pass Metric objects with explicit names to the `metrics` parameter
- Data grid columns appear in the summary with `data_` prefix; fit grid columns with `fit_` prefix
- Leverage dplyr and ggplot2 for flexible result analysis
- Enable checkpointing for studies with many replicates or slow model fitting

For more advanced topics, see the other vignettes: `vignette("custom-fitters")` for creating custom model fitters, `vignette("reproducibility")` for understanding determinism guarantees, and `vignette("memory-management")` for handling large simulations.
