---
title: "Designing Simulation Studies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Designing Simulation Studies}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Designing simulation studies

This vignette walks through designing, running, and analyzing a simulation
study with bayesim. It runs end-to-end with `LinearRegressionFitter(n_draws = 200L)` (no Stan required).

## Data and fit grids

A study is the Cartesian product of a **data grid** (conditions that change
the generated data) and a **fit grid** (conditions that change the model),
crossed with `n_replicates`. Each cell is one task.

```{r}
library(bayesim)

# A data generator that consumes the ambient RNG state (restored per task).
my_generator <- function(data_spec, task_ctx) {
  n <- data_spec$n
  x <- stats::rnorm(n)
  y <- data_spec$slope * x + stats::rnorm(n, sd = data_spec$sigma)
  list(
    train = data.frame(y = y, x = x),
    test = NULL,
    response = "y",
    true_params = c(slope = data_spec$slope, sigma = data_spec$sigma),
    vars_of_interest = c("slope", "sigma")
  )
}

data_grid <- data.frame(
  n = c(50, 100, 200),
  slope = c(0, 1, 2),
  sigma = 1
)
fit_grid <- data.frame(model = "baseline")
```

## Configuring the study

```{r}
config <- simulation_config(
  data_grid = data_grid,
  fit_grid = fit_grid,
  data_generator = my_generator,
  fitter = LinearRegressionFitter(n_draws = 200L),
  metrics = list(
    pred_rmse_metric(),
    pred_bias_metric(),
    posterior_summary_metric(),
    convergence_metric()
  ),
  n_replicates = 4L,
  seed = 2024L
)
```

## Running

```{r}
result <- run_simulation(config, progress = FALSE)
nrow(result$summary)
```

## Aggregating with MCSE

`summarize_simulation()` collapses the per-task summary by condition, adding
Monte Carlo standard errors (following rsimsum's formulas):

```{r}
agg <- summarize_simulation(result, by = c("n", "slope"))
agg
```

Each metric column gains `_mean`, `_median`, `_sd`, and `_mcse` variants,
plus `n_reps`, `n_failed`, and `failure_rate`.

## Plotting a metric across conditions

```{r, eval=FALSE}
# Requires ggplot2
plot_metric(result, "rmse__value", x = "n")
plot_metric(result, "bias__value", x = "slope")
```

## Larger studies

For studies too large to hold in memory, use `retain = "metrics"` and a modest
`checkpoint_every`, and checkpoint to `result_path` so the run is interruptible
and resumable (see `vignette("memory-management")`). For brms-based studies the
model bank compiles each distinct model once, so a study sweeping many
replicates of the same model pays the compile cost only once
(see `vignette("custom-fitters")`).
