---
title: "Running bayesim with targets"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Running bayesim with targets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{bayesim}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bayesim)
```

# Introduction

[targets](https://docs.ropensci.org/targets/) is a Make-like pipeline engine for
R: it caches the return value of each step and only re-runs a step when its
inputs change. That maps onto bayesim naturally because every simulation has a
stable **config fingerprint** --- a SHA256 hash of the data/fit grids, the
generator/fitter/metrics specs, `n_replicates`, and the seed (see
`vignette("reproducibility")`). The fingerprint is a ready-made cache key: feed
it to a `targets` cue and the pipeline re-fits only when the simulation design
actually changes.

`targets` is not a hard dependency of bayesim. The pipeline code below is
`eval = FALSE`; install `targets` separately if you want to run it.

# A minimal targets pipeline

A typical workflow defines one target per simulation condition (so conditions
re-run independently when their inputs change) or one target for the whole
config. Both are shown below in a single `_targets.R`.

The config fingerprint drives a `tar_cue()` so that a target is rebuilt only
when its design changes --- not on every `tar_make()` call:

```{r, eval=FALSE}
# _targets.R
library(targets)
library(bayesim)
library(tarchetypes) # for tar_cue()

# Build the data and fit grids once, up front.
data_grid <- data.frame(
  n = c(100, 500, 1000),
  effect = c(0.2, 0.5, 1.0)
)
fit_grid <- data.frame(model = c("baseline", "full"))

# One row per condition; we will map over it to define one target per condition.
conditions <- expand.grid(
  data_idx = seq_len(nrow(data_grid)),
  fit_idx = seq_len(nrow(fit_grid)),
  stringsAsFactors = FALSE
)

# A cue that fires only when the fingerprint changes. The fingerprint excludes
# runtime policy (result_path, retain, checkpoint cadence, max_errors, stop_on),
# so bumping those does not invalidate the cache.
fingerprint_cue <- function(config) {
  fp <- bayesim:::compute_config_fingerprint(config)
  tar_cue(fingerprint = fp)
}

list(
  # Pre-define the (static) grids so they show up as their own targets.
  tarchetypes::tar_file_read(
    conditions,
    command = readRDS("conditions.rds"),
    read = readRDS(!!"conditions.rds")
  ),

  # One target per (data_idx, fit_idx) condition, iterating in "group" mode so
  # each condition is built independently.
  tar_target(
    sim_condition,
    {
      config <- simulation_config(
        data_grid = data_grid[conditions$data_idx[sim_condition_group], , drop = FALSE],
        fit_grid = fit_grid[conditions$fit_idx[sim_condition_group], , drop = FALSE],
        data_generator = my_data_generator,
        fitter = LinearRegressionFitter(n_draws = 500L),
        metrics = list(pred_rmse_metric(), coverage_metric()),
        n_replicates = 200L,
        seed = 42L
      )
      run_simulation(config, progress = FALSE)
    },
    pattern = group(conditions),
    iteration = "group",
    cue = tar_cue(fingerprint = bayesim:::compute_config_fingerprint(
      simulation_config(
        data_grid = data_grid,
        fit_grid = fit_grid,
        data_generator = my_data_generator,
        fitter = LinearRegressionFitter(n_draws = 500L),
        metrics = list(pred_rmse_metric(), coverage_metric()),
        n_replicates = 200L,
        seed = 42L
      )
    ))
  ),

  # A single-target alternative: one target for the whole config.
  tar_target(
    sim_all,
    {
      config <- simulation_config(
        data_grid = data_grid,
        fit_grid = fit_grid,
        data_generator = my_data_generator,
        fitter = LinearRegressionFitter(n_draws = 500L),
        metrics = list(pred_rmse_metric(), coverage_metric()),
        n_replicates = 200L,
        seed = 42L
      )
      run_simulation(config, progress = FALSE)
    },
    cue = tar_cue(fingerprint = bayesim:::compute_config_fingerprint(
      simulation_config(
        data_grid = data_grid,
        fit_grid = fit_grid,
        data_generator = my_data_generator,
        fitter = LinearRegressionFitter(n_draws = 500L),
        metrics = list(pred_rmse_metric(), coverage_metric()),
        n_replicates = 200L,
        seed = 42L
      )
    ))
  )
)
```

Notes on the cue:

- The fingerprint is computed from a representative `simulation_config()`; the
  cue re-evaluates that config on every `tar_make()` and compares the resulting
  hash to the cached one.
- Runtime policy (`result_path`, `retain`, `checkpoint_every`, `max_errors`,
  `stop_on`) is excluded from the fingerprint by design, so tweaking those does
  not force a re-fit.

# Combining results

Once the per-condition targets have built, assemble them into a single
performance table with `tar_read()` / `tar_load()`. bayesim's
`performance_measures()` expects a summary data frame, so we row-bind the
per-condition summaries first:

```{r, eval=FALSE}
# In an interactive session, after tar_make():
library(targets)
library(dplyr)

# Load all per-condition results.
conditions_results <- tar_read(sim_condition)          # list, one element per branch
# Or, if iteration = "group" returns grouped chunks:
tar_load(sim_condition)

combined <- bind_rows(lapply(conditions_results, \(r) r$summary))

pm <- performance_measures(combined, estimand = "effect")
pm
```

For the whole-config target, `tar_read(sim_all)$summary` gives the full
per-task summary directly, and `summarize_simulation(tar_read(sim_all))`
produces the aggregated condition-level table with MCSEs.

# Next steps

- `vignette("getting-started")` for `simulation_config()` and
  `performance_measures()`.
- `vignette("reproducibility")` for what exactly feeds the config fingerprint
  (and what does not).
- `vignette("hpc")` for the remote-daemon / SLURM side of large runs, which
  composes naturally with a `targets` driver.
